% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Joseph Stiglitz}
\title{ Assault on Social Spending  }
\date{2011, április 7 \thanks{Demo} }
\maketitle

\begin{abstract}
  Nobel Economist Joseph Stiglitz: Assault on Social Spending, Pro-Rich Tax Cuts Turning U.S. into Nation "Of the 1 Percent, by the 1 Percent, for the 1 Percent"
\end{abstract}


This week Republicans unveiled a budget proposal for 2012 that cuts more than \$5.8 trillion in government spending over the next decade. The plan calls for sweeping changes to Medicaid and Medicare, while reducing the top corporate and individual tax rates to 25 percent. We speak to Nobel Prize-winning economist Joseph Stiglitz, who addresses the growing class divide taking place in the United States and inequality in a new \emph{Vanity Fair} article titled "Of the 1, by the 1, for the 1\%." Stiglitz is a professor at Columbia University and author of numerous books, most recently \emph{Freefall: America, Free Markets, and the Sinking of the World Economy}. "It's not just that the people at the top are getting richer," Stiglitz says. "Actually, they're gaining, and everybody else is decreasing... And right now, we are worse than old Europe."

\textbf{JUAN GONZALEZ:} This week Republicans unveiled a budget proposal for 2012 that cuts more than \$5.8 trillion in government spending over the next decade. The plan also calls for reducing the top corporate and individual tax rates to 25 percent.

Meanwhile, Republicans remain locked in an impasse with Democrats and the White House over a budget to cover the remainder of 2011. Last night President Obama and House Speaker John Boehner met to narrow their differences in order to avoid the first government shutdown in 15 years, but they failed to come to an agreement.

On Tuesday, Speaker Boehner said Republicans might now seek up to \$40 billion in cuts---around \$7 billion more than both sides had discussed over the past week. Failure to agree on a budget would lead to a federal government shutdown beginning Friday at midnight. President Obama blamed Republicans for the political stalemate.

\textbf{PRESIDENT BARACK OBAMA:} After weeks of negotiations, we've now agreed to cut as much spending as the Republicans in Congress originally asked for. I've got some Democrats mad at me, but I said, "You know what? Let's get past last year's budget, and let's focus on the future." So we've agreed to a compromise. But somehow we still don't have a deal, because some folks are trying to inject politics in what should be a simple debate about how to pay our bills. I mean, there's stuff on all kinds of issues in there: abortion and the environment and healthcare. You know, there are time to have those discussions, but that time is not now. Right now we need to just make sure that we pay our bills and that the government stays open.

\textbf{JUAN GONZALEZ:} Well, according to a recent article by our next guest, this budget is unveiled at a time when one percent of the people in the United States take in nearly a quarter of the nation's income.

\textbf{AMY GOODMAN:} "Of the 1, by the 1, for the 1\%." That's the title of an \href{http://www.vanityfair.com/society/features/2011/05/top-one-percent-201105}{article} by Joe Stiglitz appearing in this month's issue of \emph{Vanity Fair}. The piece discusses growing inequalities in the United States.

Joseph Stiglitz is a Nobel Prize-winning economist, professor at Columbia University, author of numerous books, most recently \emph{Freefall: America, Free Markets, and the Sinking of the World Economy}.

Welcome to \emph{Democracy Now!}

\textbf{JOSEPH STIGLITZ:} Nice to be here.

\textbf{AMY GOODMAN:} So, "Of the 1, by the 1, for the 1\%." Explain.

\textbf{JOSEPH STIGLITZ:} The point is that there has been this growing inequality, not only in income, but actually the inequality of wealth is even much greater. There's a shrinking of opportunity. It's not just that the people at the top are getting richer. If they were getting richer because they were contributing more to our society and everybody else was doing well, that would be one thing. But actually, they're gaining, and everybody else is decreasing. In fact, right now, it's not just the bottom, but even the middle, the middle, the median income---half above, half below---are poorer today than they were more than a decade ago. So, all the growth that has occurred in our country over the last decade or more has gone to the upper one, two percent.

At the same time, there's really shrinking opportunity. You know, we used to think of the United States as the land of opportunity---Horatio Alger, anybody could make it. And we used to think of ourselves as different from old Europe, as we used to call it. But the statistics show otherwise. Now, yes, we have some dramatic examples of people making it from the bottom to the top, middle to the top. But the statistics look at what happens on average: what is the chance of somebody at the bottom making it to the middle or the middle making it to the top? And right now, we are worse than old Europe. This is something dramatic that has happened over the last 30, 40 years.

And what's really happened is old Europe has become more of a land of opportunity. They reformed their education systems after World War II. They said that they had---they recognized they had a problem, and they changed. Meanwhile, the United States, we've wound up with a more fractionated society, one where more---moved more to a private education system, where those who have money can get a really first-class education. But the average American is not. And, you know, those who come out of---those education statistics that came out not long ago where---show that, on average, Americans are doing more poorly than countries around the world.

And we---this is related to some of the issues that got raised in the context of the financial crisis. Remember the discussion about the bonuses? And that's related to the same idea. The question was, if people were getting rewards for contributing to our society, a theory that was in the 19th century called "marginal-productivity theory," then you could say, "OK, those who contribute more should get more." But what we saw in that crisis was that these titans of the financial industry got mega-bonuses while their companies were making mega-losses. And while they were---as a result of their actions, our economy and the global economy went into a real tailspin, from which we have still not recovered. Their salaries have recovered, but not the rest of us.

\textbf{JUAN GONZALEZ:} Well, Joe Stiglitz, I'd like to ask you---given this trend, as you say, that's been going on now for decades, we've been startled in the last few months by Democrats, at both the national level and some state and local levels, further increasing the bonanza of this one percent. We had obviously in Congress the extension of the Bush-era tax cuts for the wealthy that were supposed to be eliminated, and here in New York state we were startled that the new Democratic governor, Andrew Cuomo, abolished a millionaire's tax that had been instituted a few years earlier, so that, in both ends, if you are a multi-millionaire in New York State, you benefited from both the federal policies as well as the recent state policies. So, the Democrats are supposed to be, to some degree or other, the party of labor.

\textbf{JOSEPH STIGLITZ:} Well, they're doing---let me just be clear: they are better than the Republican positions. And they, some of them, have fought against these tax cuts for the millionaires. But to me, this makes absolutely no sense. You know, we're---big discussion, you were talking about earlier about the budget deficit. There are only two ways to address the budget deficit: raise revenues or cut spending. And with this one percent getting so much, there's only one place really to get that extra revenue. The good news is it's relatively easy. You have 25 percent---almost 25 percent of the income in the upper one percent, you raise their taxes by a few percentage points, and you get an awful lot of money. And in many of these cases, we don't even have to raise the taxes; all we have to do is to say they ought to pay a fair tax rate.

One of the curious things---and again, this is bipartisan---we lowered the tax rates on speculators, so that speculators pay a much lower tax rate than people who work for a living. Now, is there any principle in economics that says that you should earn more money by speculating than by working hard? To me, it makes no sense. So, eliminating some of these special tax provisions would raise a lot of revenue and improve our deficit position.

This raises a very important point that I raise in my article, which is that much of the wealth of this one percent comes not from hard work, not from innovation, but from good investments in Washington, investing in political capital. And you saw that in the financial sector, where they first got deregulation, and they got a massive bailout. But it's true in lots of other areas. A lot of the natural resource companies---mining, oil---get access to natural resources that are in public lands at a discount price. So, if you can buy---if you can get access to these resources at a very low price, sell them at a high price, you make a lot of money. It's actually like money being stolen from the public.

\textbf{JUAN GONZALEZ:} I wanted to ask you about another---something that you've written about: the connection, in terms of risk analysis, between the nuclear crisis in Japan and the meltdown of the reactors there and the credit default swaps. And I would even throw in perhaps the BP blowout, which was another example of a risk analysis that said it could never happen.

\textbf{JOSEPH STIGLITZ:} Yeah. Well, I just wrote an interesting article making a comparison between our ability to judge what are called small probability events, you know, rare---events that are supposed to be rare---those in the financial market said that the kind of collapse that we had should happen once in a thousand years, once in the history of the universe. But we had a collapse in the 1980s, we had a problem in the 1990s, we have them every 10 years. And that shows the models are very bad, our ability to judge rare events is very bad. Now, a lot of research in behavioral economics and psychology have explained why it is that these events that don't happen very much, we don't have a lot of experience.

But one of the points that I raised was that these people have an incentive not to see things accurately. You know, the nuclear power industry has an incentive to tell everybody, "Oh, don't worry. Nothing---no risk there." The financial sector had an incentive to say, "Don't worry about these derivatives, even if they're already a quadrillion dollars. Don't worry, because we can manage that risk. We have systems of diversifying the risk across the economy." Clearly wrong. So, you know, when there's so much money at stake, people have a way of seeing---of discounting these risks, especially because those risks are borne by everybody else in our society.

And, you know, nuclear power is a really interesting case, because that industry has never been commercially viable. It has always existed on the back of a government-provided insurance, that we provide as taxpayers, that they don't pay for. And we see now in Japan that, you know, they did the same thing, and we see the cost of that. The rest of society is paying an enormous price. There is no way that the slight savings in energy cost can make up for the loss to the Japanese economy that has resulted from the nuclear explosion. And the same thing could happen here in the United States.

\textbf{AMY GOODMAN:} I loved seeing on \emph{Meet the Press} right after the tsunami and the earthquake and the terrible tragedy in Japan, they had on the head of the Nuclear Energy Institute, so, you know, they represent the nuclear industry, and the host of the show saying, "Thank you so much for running in at the last minute to be here with us." And I could only think about---I mean, here he is speaking to save the butts of the nuclear industry in this country and saying there's nothing to worry about here, as we're saying this---well, what is looking like a partial meltdown or more.

\textbf{JOSEPH STIGLITZ:} If the industry really believed it, let them make an unlimited liability and provide us with a guarantee that they would pick up for the financial cost of the kind of disaster that Japan is facing. And I can tell you that if you made them bear those costs, if we didn't give them that free ride of limited liability, that industry would not exist in the United States today.

\textbf{AMY GOODMAN:} In your piece, Joe Stiglitz, in \emph{Vanity Fair}, "Of the 1, by the 1, for the 1\%," you say, "Americans have been watching protests against oppressive regimes that concentrate massive wealth in the hands of an elite few. Yet in our own democracy, 1 percent of the people take nearly a quarter of the nation's income---an inequality even the wealthy will come to regret." Talk about all this. We're seeing these rolling rebellions. We are seeing rebellions not only in the Middle East, though, in the Midwest. I mean, look at Madison, Wisconsin. And what about this issue of even the wealthy will regret this?

\textbf{JOSEPH STIGLITZ:} Well, there are two points that I try to make. One is that a successful economy requires collective action. There are lots of things we have to do together. We have to have infrastructure. We have to have an educated population. If you have a divided society, you start worrying more---if you're in the wealthy and you have an electorate system that can use your wealth to affect the politics, you say, "I'd rather have a small government that isn't able to redistribute money, take money away from me. I don't need public schools; I have private money. I don't need public parks; I have private---you know, my large land." So, what you have then is an erosion of the kind of collective action, and that makes a society less efficient, less productive. And you see that already happening. We are competing in education with countries in Asia that were much poorer than we were not that long ago. So that's one problem.

And the second one is that obviously a house divided can't stand, that you start getting tensions, you start not paying attention to the things that make us cohesive as a nation. And that's what you're seeing in Wisconsin. And you also see that in the budget messages that are coming across, saying, "OK, we're going to cut back on healthcare for aged and for the poor, but we're not going to do anything about overall healthcare costs." What does that mean? It means that if you're going to cut back on health expenditures for the aged and the poor, and you're going to let health costs continue to rise, that says rationing. They're not going to be able to get healthcare. Already, we spend more money with poor health outcomes than those in other countries in the advanced industrial world. And it's going to get worse as the poor and the elderly can't get access to healthcare.

\textbf{JUAN GONZALEZ:} In that vein, I wanted to ask you about a recent dissent of yours, that a bunch of eminent economists, a former chairman of the President's Council on Economic Advisers and other members, were asked to sign on to a letter about the need for deficit reduction in the future in the United States. Why did you dissent to that letter?

\textbf{JOSEPH STIGLITZ:} Well, the letter said all economists agree that you need to get deficit reduction, we ought to show that there's something called economic science. And then they came out in support of working off the Bowles-Simpson proposal, which is the bipartisan deficit reduction. I had looked at that very carefully, and I come to the view that that is not going to make America stronger, it's not going to not make our economy stronger. I said before that there are two ways---

\textbf{JUAN GONZALEZ:} You called it a potential suicide pact?

\textbf{JOSEPH STIGLITZ:} That's right. Why is it? I mentioned before that there are two ways of cutting the deficit. One is raising revenue; the other is cutting expenditures. Well, what's the best way of getting tax revenue up? It's to put America back to work. If America were growing, if we were back to our full potential, our output would be higher, and tax revenue would be substantially higher. So the best way to reduce the deficit, as I said, is to put America back to work.

Now, I thought that Bowles-Simpson is a recipe for making our economy weaker. If you start cutting budgets now, spending now, before the economy is back to health---you know, we already suffer from the deregulation, now this is going---we're going to suffer again from the thoughtless deficit reduction. That means people are going to be out of jobs. That means revenues are going to be lower, and our economy is going to be weaker.

But the second point I raise is that it goes back to: look at where the revenues are going to be. When you have so much money going to the upper one percent, the only way of fixing the revenue problem is to raise taxes and have a more fair tax system. What they proposed was eliminating some of the deductions, say, for housing, mortgages. Well, if you do that now, house prices, which have continued to decline, would decline even more. But the more fundamental point is, even if from the long-term point of view you want to get rid of these mortgage deductions, you shouldn't do that on the backs of the middle class. That should be viewed as a revenue-neutral. You lower the mortgage deduction, but you also lower the tax rates to offset it. That's not a revenue---shouldn't be a revenue enhancer. These are the people who have been doing very badly for the last 10, 15, 20 years.

\textbf{AMY GOODMAN:} Joe Stiglitz, we're going to go to Medicare and Medicaid in the crosshairs in just one minute with two healthcare activists. But speaking of the crosshairs, let's end on the issue of war. You wrote with Linda Bilmes the book \emph{The Three Trillion Dollar War: The True Cost of the Iraq Conflict}. That's not talking about Afghanistan, what, \$2 billion a week, the longest ongoing conflict in U.S. history. What about the cost of this?

\textbf{JOSEPH STIGLITZ:} It's enormous. And since we wrote that book, we did---new numbers came in, and things are worse than we said. The disability rates are higher. The cost of caring for the disabled are higher. Almost one out of two people coming back from Iraq and Afghanistan are disabled. This is an unfunded liability of---we calculate now to be almost a trillion dollars, over \$900 billion. So, one of the big ways of reducing our deficit is a---is cut back some expenditures.

I believe we could have more security with much less spending. We are spending literally hundreds of billions of dollars for weapons that don't work against enemies that don't exist. The Cold War ended more than 20 years ago. And yet, if you look at our military, nobody seems to have told it that. Another way of thinking about it, we spend more money now than all the rest of the countries of the world, or almost as much as all the rest of the countries put together. And yet, when you have a case where you might arguably want some use of it---you know, to protect people who are being killed---we say we can't do it, even in a small country of a few million people. We say, "Oh, no. Our military can't do anything." So we've been spending all this money and getting actually very little security for it. So my own feeling is that we could reduce our money, our expenditures markedly---particularly, get out of Afghanistan---and improve our security.

\textbf{AMY GOODMAN:} You know, if there's a government shutdown, it's possible that Congress members who shut it down would continue to be paid, and soldiers would stop being paid. I want to thank you very much for being with us.

\textbf{JOSEPH STIGLITZ:} Thank you.

\textbf{AMY GOODMAN:} Joseph Stiglitz, Nobel Prize-winning economist, professor at Columbia University. His latest book, \emph{Freefall: Amercia, Free Markets, and the Sinking of the World Economy}. When we come back, we'll look at Medicare and Medicaid, what's being proposed in Congress right now. Yes, healthcare in the crosshairs. Stay with us.

\end{document}

