% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Kester Eddy}
\title{ Hungary's central bank `revolution' attacked  }
\date{2013. március 20 \thanks{Fin Financial Times} }
\maketitle


New management at Hungary's central bank is threatening its independence and undermining staff morale, bank officials have told the Financial Times. Ultimately, the changes will fail to boost the flagging economy, the say.

MNB, the central bank, was the last big institution not under the direct influence of Viktor Orbán, the prime minister, and his Fidesz party, but they have moved with surprising speed to consolidate control since appointing a new bank governor earlier this month.

High quality global journalism requires investment. Please share this article with others using the link below, do not cut \& paste the article. See our Ts\&Cs and Copyright Policy for more detail. Email ftsales.support@ft.com to buy additional rights. http://www.ft.com/cms/s/0/1d0b8824-90a2-11e2-862b-00144feabdc0.html\#ixzz2O3wbAJCn

Gyorgy Matolcsy, the former economy minister responsible for \textqq{unorthodox} economic policies that have angered foreign investors since Mr Orbán came to power in 2010, became the MNB's governor barely two weeks ago, promising a new era of co-operation between the bank and government.

Since then, he has used a new layer of handpicked managers to strip his two deputies of most of their main responsibilities, dismiss three senior managers and demote a fourth, while many long-term employees feel threatened.

\textqq{It is the end of {[}Orbán's{]} `revolution', the central bank was the last pillar that was not taken over,} said one bank official on condition of anonymity.

Another said: \textqq{People don't know if they will have a chair, a desk, or an office tomorrow.}

András Simor, the independent-minded former governor, successfully fought off strong government pressure to resign.

While bank staff expected changes under Mr Matolcsy, the \textqq{speed and pettiness} of the moves -- exemplified by the reduced responsibilities of Ferenc Karvalits, deputy governor, whose term expires at the end of this month -- had caused \textqq{fear} among staff, a bank employee said.

\textqq{The whole thing is about pettiness, this childish will for control, power and revenge,} the employee said.

Another said: \textqq{The new management does not respect the culture and values of this institution. {[}The{]} culture of merit was killed off when they declared they would fire key personalities {[}including{]} the guy who headed the research department: he has a PhD from Massachusets Institute of Technology, he has no political affiliations.}

The three bank officials who spoke to the FT confirmed that many staff were seeking new jobs in the general mood of uncertainty. \textqq{Many people will be looking for jobs at the EU, the European Central Bank, the International Monetary Fund. They are killing the soul of this institution,} said one.

The bank's monetary policy council, in which government appointees outnumber its internal members, has cut interest rates for seven consecutive months in an attempt to revive Hungary's economy in the run-up to general elections next year. While manufactured exports remain strong, the economy contracted by 1.7 per cent in 2012, and most independent analysts forecast stagnation at best this year.

The new bank leadership is expected to further reduce interest rates and implement policies to stimulate lending, particularly to small and medium sized domestic businesses. But the forint had weakened by 5 per cent this year -- partly due to these rate cuts, analysts said.

Measures to stimulate lending by commercial banks were likely to prove futile, bank officials claimed. \textqq{The commercial banks are freezing lending because the business environment is uncertain and taxation {[}on the banks{]} is high,} said one.

The official said that the government's looser monetary policy had not, so far, been punished by the markets, but any downturn in international sentiment would be immediately felt in Hungary.

The central bank did not respond to requests for comment.

\end{document}

