% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Jan Puhl}
\title{ Csanád Szegedi, Metamorphosis:~A Hungarian Extremist Explores His Jewish Roots  }
\date{ \thanks{} }
\maketitle

\begin{abstract}
  \textbf{Metamorphosis:~A Hungarian Extremist Explores His Jewish Roots}

\emph{By Jan Puhl}

\textbf{Csanád Szegedi was a prominent right-wing extremist in Hungary until he discovered his own Jewish roots in 2012. Since then, he has undergone a radical reinvention and is even learning Hebrew. His grandmother, though, continues to hide her Auschwitz tattoo.}

\end{abstract}


Csanád Szegedi's second life began in the apartment of Rabbi Baruch Oberlander, located above the Synagogue in the Erzsébetváros quarter of Budapest. A mohel -\/- a circumcision specialist -\/- had arrived from Israel. And with a single cut, the anti-Semite Csanád was transformed into Dovid, a Jew.

Csanád Szegedi, 31, had been the deputy head of right-wing extremist party Jobbik, which he also represented in the European Parliament. He had made a career of claiming that the Jews sought to plunder Hungary and that they had entered into an alliance with~\href{http://www.spiegel.de/international/europe/europe-failing-to-protect-roma-from-discrimination-and-poverty-a-942057.html}{\emph{the Roma}}~to turn "pure" Hungarians into a minority in their own country. In public, he would often wear the black military pants and vest of the Hungarian Guard, the banned right-wing extremist group.

But then he learned that his family was Jewish, a revelation that turned his life on its head.

Now, he calls himself Dovid Szegedi, eats kosher, is learning Hebrew and goes to the Synagogue every Friday. "This is my true identity," says Szegedi, who is almost two meters (6" 6') tall. He wears an Italian designer suit, scruffy stubble and a black kippah.

The story of Csanád's transformation into Dovid is one of radical reinvention, and also one of a desperate search for a reliable identity, one which continues to elude Eastern Europe even 25 years after the end of communism.

It has been particularly elusive for Hungary, a country that joined Hitler's alliance in World War II and which delivered more than a half-million of its Jews to the Nazis' murderous machine. In 1989, it became the first Eastern Bloc country to open its borders to the West and was long considered to be a model student of capitalism. Until it almost went broke five years ago.

\section{Megalomania and Guilt}

Today, the nationalist-conservative Fidesz party holds the reins of power in Budapest, under the leadership of~\href{http://www.spiegel.de/international/europe/hungarian-government-puts-european-conservatives-in-tough-spot-a-902620.html}{\emph{Prime Minister Viktor Orbán}}, a man for whom democracy is something of a bother. In recent years, he has pushed through legislation limiting press freedoms and reduced the power of his country's highest court. In Hungary's April 6 general election, Fidesz will likely win an absolute majority once again, with surveys showing that the right-wing extremists from Jobbik stand to rake in an additional 15 percent of the vote.

Nowhere in Europe do competing political camps view each other with as much hatred as they do in Hungary. There is hardly another country that swings as radically between the emotional extremes of nationalist megalomania and deep-seated feelings of guilt.

Shortly before the sun sinks behind the houses of Buda and the Sabbath begins on a recent Friday, Dovid Szegedi drives up to the Obuda Synagogue together with his mother, wife and two sons. When it was completed in 1820, it was the biggest synagogue in Budapest; its columned facade gave it the look of a Corinthian temple. Following the Holocaust, it was used as a warehouse before doing service as a studio for a state-run television network.

It re-opened its doors to the Jewish faithful four years ago with a Torah ark funded by American Jews and freshly restored frescoes on the ceiling. Blessings in Hebrew, which Szegedi can read with some effort, stand above the water faucets.

There are 613 commandments that devout Jews must adhere to, says Szegedi, adding that he has managed to conform to 80 of them thus far. "I am trying, but it doesn't happen overnight."

On this Friday, some 100 worshippers are on hand, with the women taking their places in a separate area partitioned by a curtain. Szegedi's sons romp in the sanctuary, their kippahs repeatedly falling from their heads. Dovid Szegedi still seems a bit awkward as his upper body sways to and fro in time with the prayers.

\section{Doing Everything to Become Jewish}

Rabbi Slomo Köves speaks of desires, covetousness and restraint -\/- and of the Internet, which makes it so difficult to be satisfied with oneself and one's life. At the end of the service, worshippers wish each other "Shabbat Shalom" and head upstairs to the community hall.

Men and women sit together to eat at round tables. Jews are not allowed to work on the Sabbath and Christians have been hired to prepare the kosher meal: humus, cucumber salad and chicken. A small glass of red wine stands next to each plate. Every now and then, a member of the congregation stands up to wish the gathering well or relate an anecdote.

A young man from the neighboring table bends over to Szegedi and asks "And? Have you made progress with your Hebrew?" Szegedi has become a part of the congregation, but that wasn't always the case. The first time he came to a service, many got up and left the synagogue out of protest.

"They didn't believe that I would really change," Szegedi says. "Many could only forgive me my past with great difficulty. Which I can totally understand." After all, he has a hard time believing it himself.

In the summer of 2012, Szegedi left the Jobbik party, found a rabbi and began studying the Torah. He traveled to Israel, visited the Yad Vashem Holocaust memorial and went to the Western Wall. He did everything he could to become Jewish.

But he continued to wake up in a cold sweat at night, burning with shame. He would turn on his computer and watch speeches from his past on YouTube. He was never one to scream and rant. Rather he recited his~\href{http://www.spiegel.de/international/europe/trial-in-hungary-little-outrage-over-far-right-murders-of-6-roma-a-912709.html}{\emph{hatred of Roma}}, Slovakians, the EU and "international financial capital" in a calm and precise, almost cold, tone of voice. That is what made him particularly dangerous.

"Today, I want to make amends for my past with good deeds," he says.

Csanád Szegedi was born in eastern Hungary in 1982. When he started school, the old communist school books were being phased out and his history teacher was a fervent nationalist who spoke frequently of the Treaty of Trianon, the 1920 peace agreement which stripped the erstwhile Kingdom of Hungary of two-thirds of its territory, giving it to its neighbors. For Hungarian nationalists, it is an historical trauma which forms the core of their worldview.

\section{'An Addiction'}

Szegedi studied history at university in Budapest and joined together with other right-wing students, a group that was full of hatred toward the communists who had managed to regain power in Hungary just a decade after 1989. It was a different communist party, to be sure, but Szegedi didn't believe that it had truly changed.

Szegedi also remained unconvinced by the Fidesz party, finding its high-brow brand of nationalism too placid. He was more radical. He didn't want a debate; he wanted to burn European Union flags. Hate slowly became his political currency. It all began with the Roma, he says. "First, it was just the criminal Roma, then all of them. Later, you start hating the Slovaks and the Romanians. You always need more. It is like an addiction."

In 2003, he became one of the founders of the Jobbik party; his membership number is 63. They saw themselves as the avant-garde, courageous enough to say what many Hungarians secretly thought: The Roma have to go, the left-wing should be locked up and Brussels wants to enslave Hungary.

Anti-Semitism was also~\href{http://www.spiegel.de/international/europe/prime-minister-viktor-orban-downplays-rising-anti-semitism-in-hungary-a-898319.html}{\emph{widespread}}~in the party. Many Jobbik members believed that the Holocaust and extermination camps like Auschwitz were nothing but Jewish inventions in order to make others feel guilty and extort money from them.

That was Szegedi's world back then -\/- and he became something of a radical thought leader for the far right. He even wrote a book entitled "I Believe in Hungary's Resurrection." In the forward, one of his Jobbik comrades writes: "He is our party's fist." Szegedi was also making a good living selling T-shirts with runic writing and other right-wing trinkets from his online shop.

Some were envious of Szegedi's success -\/- people like Zoltan Ambrus, a right-wing acquaintance of Szegedi's who served time for possession of a pistol and plastic explosives.

\section{A Sinister Secret}

Four years ago, Ambrus was given documents -\/- likely from the archives of Hungary's communist secret police -\/- showing that Szegedi is of Jewish origin. Ambrus challenged Szegedi, who by then was a representative in European Parliament, and filmed the conversation. Szegedi reacted by trying to bribe Ambrus, offering him jobs and money in exchange for silence. But Ambrus refused and informed the party. In the summer of 2012, Jobbik leaders confronted Szegedi, with party head Gábor Vona initially enthusiastic about the news. "Super," he said. "You are now our shield against accusations that we are anti-Semitic."

But the majority of the party was of a different opinion. One member said: The best would be to shoot you in the head right now.

Suddenly, he was Jewish. And he was reminded of a subconscious feeling from his childhood that something wasn't quite right. On the one hand, Szegedi says, he learned early on at home "that being Jewish is somehow bad." On the other hand, though, his mother once slapped him when he told a Jewish joke that he had heard in school. It was only now, though, as an adult, that he learned more about his ancestors -\/- a history that had gone unmentioned until then.

Before the war, some 14,000 Jews had lived in Miskolc, where Szegedi was born. Only 105 of them returned from the concentration camps after the violence had ended. One of them was his grandmother. She kept her past from her children like a sinister secret. Even on extremely hot days, she would wear long-sleeved blouses or slapped a bandage over the prisoner number tattooed on her arm. Many Eastern European Jews continue to do so today.

It was Hungarian gendarmes who herded Szegedi's grandmother Magdolna Klein, 25-years-old at the time, into an Auschwitz-bound cattle car in 1944. The first German she saw was the notorious Holocaust doctor Josef Mengele. He sent her to the right -\/- to join those who were able to work. Most of the others were pointed to the left -\/- to the gas chambers.

One morning, Magdolna's blanket was missing, an infraction that usually resulted in death. But the guard merely slapped her and sent her to a different labor commando. It was intended as a punishment, but it likely saved her life: The work she was now asked to do was relatively easy, helping her to survive four months in Auschwitz before the Red Army liberated the camp on Jan. 27, 1945.

\section{Fear of Anti-Semitism}

Magdolna Klein returned to Hungary and tried to lead a normal life. She married a Jewish man who had also survived Auschwitz, but whose first wife and two children had been murdered. The couple went to synagogue every Friday.

But then, in 1956, Hungarians rebelled against the communists and Szegedi's grandmother saw how a mob hanged a policeman from a tree and tore down his pants. He was circumcized. For the gathered crowd, it was proof that Hungary's communists and their henchmen were Jewish. It was back,~\href{http://www.spiegel.de/international/europe/jewish-world-congress-in-budapest-protests-hungarian-anti-semitism-a-897718.html}{\emph{the old anti-Semitism}}~that Magdolna hoped had vanished along with Nazi Germany. On that day, she and her husband decided to cease being Jewish. They never again visited a synagogue and stopped speaking about Auschwitz.

With such a family background, says Szegedi, it was no longer possible to be a right-wing radical, much less a member of Jobbik. Which is why, one day, he wrote a text message to Rabbi Köves. "Please call me," he wrote.

The rabbi initially believed it was a joke, but then he agreed to a meeting in the austere conference room of the Lubavitch community. "I met with a man in freefall," Köves says. "He had lost all of his friends and all of his certainties."

Rabbi Köves, 34, consulted with his colleagues and came to the conclusion that "despite his past, Szegedi needed help." Soon, Köves arranged another meeting with Szegedi, during which they talked about anti-Semitism and Jewish faith. The discussion was akin to a wrestling match. "The Jews are buying up Hungary," Szegedi claimed, for example. "There are some Israeli investors who are creating jobs for Hungarians," the rabbi responded. It went on like that for a whole day.

\section{A Jewish Family}

Rabbi Köves prescribed Szegedi a rigorous regime of study and contemplation -\/- and told him not to speak to the media. And step by step, Csanád Szegedi overcame his past. He says he would have left Jobbik if the party hadn't kicked him out. He closed down his online shop, sent his unsold stock back to the suppliers and threw away his books. The only thing he held onto was his seat in European Parliament, so that it wouldn't be filled by another right-winger from Jobbik.

Some, such as the Budapest philosopher Gáspár Miklós Tamás, have accused Szegedi of remaining a racist. As long as he believed he was a "pure Hungarian," Miklós has said, he saw things through a nationalist lens. But as soon as he learned of his past, he began seeing the world from an exclusively Jewish point of view.

Rabbi Köves shakes his head; he believes Szegedi has experienced a true catharsis. "Szegedi learned from his party and from his former friends what it means to be discriminated against," Köves says. "Suddenly, it no longer mattered what he did for the party. It only mattered that he was Jewish." Had the Jobbik leadership been more tolerant, Köves believes, perhaps Csanád would never have become Dovid.

His entire family has now returned to being Jewish -\/- except for his grandmother. "She still thinks that the Holocaust could be repeated at any moment and that hatred of the Jews will never disappear," Szegedi says. When she learned that he was planning to become circumcized, she tried to talk him out of it. And she still doesn't go to the synagogue.

\emph{Translated from the German by Charles Hawley}

\end{document}

