% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Noam Chomsky}
\title{ America hates its poor  }
\date{2013, december 1 \thanks{Salon } }
\maketitle

\begin{abstract}
\textbf{Noam Chomsky: America hates its poor}

Noam Chomsky interviewed by Chris Steele

Excerpt from~\emph{Occupy}~(Second edition), Zucotti Park Press

  
\end{abstract}


Q: An article that recently came out in Rolling Stone, titled "Gangster Bankers: Too Big to Jail," by Matt Taibbi, asserts that the government is afraid to prosecute powerful bankers, such as those running HSBC. Taibbi says that there's "an arrestable class and an unarrestable class." What is your view on the current state of class war in the U.S.?

A: Well, there's always a class war going on. The United States, to an unusual extent, is a business-run society, more so than others. The business classes are very class-conscious -\/- they're constantly fighting a bitter class war to improve their power and diminish opposition. Occasionally this is recognized.

We don't use the term "working class" here because it's a taboo term. You're supposed to say "middle class," because it helps diminish the understanding that there's a class war going on.

It's true that there was a one-sided class war, and that's because the other side hadn't chosen to participate, so the union leadership had for years pursued a policy of making a compact with the corporations, in which their workers, say the autoworkers -\/- would get certain benefits like fairly decent wages, health benefits and so on. But it wouldn't engage the general class structure. In fact, that's one of the reasons why Canada has a national health program and the United States doesn't. The same unions on the other side of the border were calling for health care for everybody. Here they were calling for health care for themselves and they got it. Of course, it's a compact with corporations that the corporations can break anytime they want, and by the 1970s they were planning to break it and we've seen what has happened since.

This is just one part of a long and continuing class war against working people and the poor. It's a war that is conducted by a highly class-conscious business leadership, and it's one of the reasons for the unusual history of the U.S. labor movement. In the U.S., organized labor has been repeatedly and extensively crushed, and has endured a very violent history as compared with other countries.

In the late 19th century there was a major union organization, Knights of Labor, and also a radical populist movement based on farmers. It's hard to believe, but it was based in Texas, and it was quite radical. They wanted their own banks, their own cooperatives, their own control over sales and commerce. It became a huge movement that spread over major farming areas.

The Farmers' Alliance did try to link up with the Knights of Labor, which would have been a major class-based organization if it had succeeded. But the Knights of Labor were crushed by violence, and the Farmers' Alliance was dismantled in other ways. As a result, one of the major popular democratic forces in American history was essentially dismantled. There are a lot of reasons for it, one of which was that the Civil War has never really ended. One effect of the Civil War was that the political parties that came out of it were sectarian parties, so the slogan was, "You vote where you shoot," and that remains the case.

Take a look at the red states and the blue states in the last election: It's the Civil War. They've changed party labels, but other than that, it's the same: sectarian parties that are not class-based because divisions are along different lines. There are a lot of reasons for it.

The enormous benefits given to the very wealthy, the privileges for the very wealthy here, are way beyond those of other comparable societies and are part of the ongoing class war. Take a look at CEO salaries. CEOs are no more productive or brilliant here than they are in Europe, but the pay, bonuses, and enormous power they get here are out of sight. They're probably a drain on the economy, and they become even more powerful when they are able to gain control of policy decisions.

That's why we have a sequester over the deficit and not over jobs, which is what really matters to the population. But it doesn't matter to the banks, so the heck with it. It also illustrates the consider- able shredding of the whole system of democracy. So, by now, they rank people by income level or wages roughly the same: The bottom 70 percent or so are virtually disenfranchised; they have almost no influence on policy, and as you move up the scale you get more influence. At the very top, you basically run the show.

A good topic to research, if possible, would be "why people don't vote." Nonvoting is very high, roughly 50 percent, even in presidential elections -\/- much higher in others. The attitudes of people who don't vote are studied. First of all, they mostly identify themselves as Democrats. And if you look at their attitudes, they are mostly Social Democratic. They want jobs, they want benefits, they want the government to be involved in social services and so on, but they don't vote, partly, I suppose, because of the impediments to voting. It's not a big secret. Republicans try really hard to prevent people from voting, because the more that people vote, the more trouble they are in. There are other reasons why people don't vote. I suspect, but don't know how to prove, that part of the reason people don't vote is they just know their votes don't make any difference, so why make the effort? So you end up with a kind of plutocracy in which the public opinion doesn't matter much. It is not unlike other countries in this respect, but more extreme. All along, it's more extreme. So yes, there is a constant class war going on.

The case of labor is crucial, because it is the base of organization of any popular opposition to the rule of capital, and so it has to be dismantled. There's a tax on labor all the time. During the 1920s, the labor movement was virtually smashed by Wilson's Red Scare and other things. In the 1930s, it reconstituted and was the driving force of the New Deal, with the CIO organizing and so on. By the late 1930s, the business classes were organizing to try to react to this. They began, but couldn't do much during the war, because things were on hold, but immediately after the war it picked up with the Taft-Hartley Act and huge propaganda campaigns, which had massive effect. Over the years, the effort to undermine the unions and labor generally succeeded. By now, private-sector unionization is very low, partly because, since Reagan, government has pretty much told employers, "You know you can violate the laws, and we're not going to do anything about it." Under Clinton, NAFTA offered a method for employers to illegally undermine labor organizing by threatening to move enterprises to Mexico. A number of illegal operations by employers shot up at that time. What's left are private-sector unions, and they're under bipartisan attack.

They've been protected somewhat because the federal laws did function for the public-sector unions, but now they're under bipartisan attack. When Obama declares a pay freeze for federal workers, that's actually a tax on federal workers. It comes to the same thing, and, of course, this is right at the time we say that we can't raise taxes on the very rich. Take the last tax agreement where the Republicans claimed, "We already gave up tax increases." Take a look at what happened. Raising the payroll tax, which is a tax on working people, is much more of a tax increase than raising taxes on the super-rich, but that passed quietly because we don't look at those things.

The same is happening across the board. There are major efforts being made to dismantle Social Security, the public schools, the post office -\/- anything that benefits the population has to be dismantled. Efforts against the U.S. Postal Service are particularly surreal. I'm old enough to remember the Great Depression, a time when the country was quite poor but there were still postal deliveries. Today, post offices, Social Security, and public schools all have to be dismantled because they are seen as being based on a principle that is regarded as extremely dangerous.

If you care about other people, that's now a very dangerous idea. If you care about other people, you might try to organize to undermine power and authority. That's not going to happen if you care only about yourself. Maybe you can become rich, but you don't care whether other people's kids can go to school, or can afford food to eat, or things like that. In the United States, that's called "libertarian" for some wild reason. I mean, it's actually highly authoritarian, but that doctrine is extremely important for power systems as a way of atomizing and undermining the public.

That's why unions had the slogan, "solidarity," even though they may not have lived up to it. And that's what really counts: solidarity, mutual aid, care for one another and so on. And it's really important for power systems to undermine that ideologically, so huge efforts go into it. Even trying to stimulate consumerism is an effort to undermine it. Having a market society automatically carries with it an undermining of solidarity. For example, in the market system you have a choice: You can buy a Toyota or you can buy a Ford, but you can't buy a subway because that's not offered. Market systems don't offer common goods; they offer private consumption. If you want a subway, you're going to have to get together with other people and make a collective decision. Otherwise, it's simply not an option within the market system, and as democracy is increasingly undermined, it's less and less of an option within the public system. All of these things converge, and they're all part of general class war.

Q: Can you give some insight on how the labor movement could rebuild in the United States?

A: Well, it's been done before. Each time labor has been attacked -\/- and as I said, in the 1920s the labor movement was practically destroyed -\/- popular efforts were able to reconstitute it. That can happen again. It's not going to be easy. There are institutional barriers, ideological barriers, cultural barriers. One big problem is that the white working class has been pretty much abandoned by the political system. The Democrats don't even try to organize them anymore. The Republicans claim to do it; they get most of the vote, but they do it on non-economic issues, on non-labor issues. They often try to mobilize them on the grounds of issues steeped in racism and sexism and so on, and here the liberal policies of the 1960s had a harmful effect because of some of the ways in which they were carried out. There are some pretty good studies of this. Take busing to integrate schools. In principle, it made some sense, if you wanted to try to overcome segregated schools. Obviously, it didn't work. Schools are probably more segregated now for all kinds of reasons, but the way it was originally done undermined class solidarity.

For example, in Boston there was a program for integrating the schools through busing, but the way it worked was restricted to urban Boston, downtown Boston. So black kids were sent to the Irish neighborhoods and conversely, but the suburbs were left out. The suburbs are more affluent, professional and so on, so they were kind of out of it. Well, what happens when you send black kids into an Irish neighborhood? What happens when some Irish telephone linemen who have worked all their lives finally got enough money to buy small houses in a neighborhood where they want to send their kids to the local school and cheer for the local football team and have a community, and so on? All of a sudden, some of their kids are being sent out, and black kids are coming in. How do you think at least some of these guys will feel? At least some end up being racists. The suburbs are out of it, so they can cluck their tongues about how racist everyone is elsewhere, and that kind of pattern was carried out all over the country.

The same has been true of women's rights. But when you have a working class that's under real pressure, you know, people are going to say that rights are being undermined, that jobs are being under- mined. Maybe the one thing that the white working man can hang onto is that he runs his home? Now that that's being taken away and nothing is being offered, he's not part of the program of advancing women's rights. That's fine for college professors, but it has a different effect in working-class areas. It doesn't have to be that way. It depends on how it's done, and it was done in a way that simply undermined natural solidarity. There are a lot of factors that play into it, but by this point it's going to be pretty hard to organize the working class on the grounds that should really concern them: common solidarity, common welfare.

In some ways, it shouldn't be too hard, because these attitudes are really prized by most of the population. If you look at Tea Party members, the kind that say, "Get the government off my back, I want a small government" and so on, when their attitudes are studied, it turns out that they're mostly social democratic. You know, people are human after all. So yes, you want more money for health, for help, for people who need it and so on and so forth, but "I don't want the government, get that off my back" and related attitudes are tricky to overcome.

Some polls are pretty amazing. There was one conducted in the South right before the presidential elections. Just Southern whites, I think, were asked about the economic plans of the two candidates, Barack Obama and Mitt Romney. Southern whites said they preferred Romney's plan, but when asked about its particular components, they opposed every one. Well, that's the effect of good propaganda: getting people not to think in terms of their own interests, let alone the interest of communities and the class they're part of. Overcoming that takes a lot of work. I don't think it's impossible, but it's not going to happen easily.

Q: In a recent article about the Magna Carta and the Charter of the Forest, you discuss Henry Vane, who was beheaded for drafting a petition that called the people's power "the original from whence all just power arises." Would you agree the coordinated repression of Occupy was like the beheading of Vane?

A: Occupy hasn't been treated nicely, but we shouldn't exaggerate. Compared with the kind of repression that usually goes on, it wasn't that severe. Just ask people who were part of the civil rights movement in the early 1960s, in the South, let's say. It was incomparably worse, as was just showing up at anti-war demonstrations where people were getting maced and beaten and so on. Activist groups get repressed. Power systems don't pat them on the head. Occupy was treated badly, but not off the spectrum -\/- in fact, in some ways not as bad as others. I wouldn't draw exaggerated comparisons. It's not like beheading somebody who says, "Let's have popular power."

Q: How does the Charter of the Forest relate to environmental and indigenous resistance to the Keystone XL pipeline?

A: A lot. The Charter of the Forest, which was half the Magna Carta, has more or less been forgotten. The forest didn't just mean the woods. It meant common property, the source of food, fuel. It was a common possession, so it was cared for. The forests were cultivated in common and kept functioning, because they were part of people's common possessions, their source of livelihood, and even a source of dignity. That slowly collapsed in England under the enclosure movements, the state efforts to shift to private ownership and control. In the United States it happened differently, but the privatization is similar. What you end up with is the widely held belief, now standard doctrine, that's called "the tragedy of the commons" in Garrett Hardin's phrase. According to this view, if things are held in common and aren't privately owned, they're going to be destroyed. History shows the exact opposite: When things were held in common, they were preserved and maintained. But, according to the capitalist ethic, if things aren't privately owned, they're going to be ruined, and that's "the tragedy of the commons." So, therefore, you have to put everything under private control and take it away from the public, because the public is just going to destroy it.

Now, how does that relate to the environmental problem? Very significantly: the commons are the environment. When they're a common possession -\/- not owned, but everybody holds them together in a community -\/- they're preserved, sustained and cultivated for the next generation. If they're privately owned, they're going to be destroyed for profit; that's what private owner- ship is, and that's exactly what's happening today.

What you say about the indigenous population is very striking. There's a major problem that the whole species is facing. A likelihood of serious disaster may be not far off. We are approaching a kind of tipping point, where climate change becomes irreversible. It could be a couple of decades, maybe less, but the predictions are constantly being shown to be too conservative. It is a very serious danger; no sane person can doubt it. The whole species is facing a real threat for the first time in its history of serious disaster, and there are some people trying to do some- thing about it and there are others trying to make it worse. Who are they? Well, the ones who are trying to make it better are the pre-industrial societies, the pre-technological societies, the indigenous societies, the First Nations. All around the world, these are the communities that are trying to preserve the rights of nature.

The rich societies, like the United States and Canada, are acting in ways to bring about disaster as quickly as possible. That's what it means, for example, when both political parties and the press talk enthusiastically about "a century of energy independence." "Energy independence" doesn't mean a damn thing, but put that aside. A century of "energy independence" means that we make sure that every bit of Earth's fossil fuels comes out of the ground and we burn it. In societies that have large indigenous populations, like, for example, Ecuador, an oil producer, people are trying to get support for keeping the oil in the ground. They want funding so as to keep the oil where it ought to be. We, however, have to get everything out of the ground, including tar sands, then burn it, which makes things as bad as possible as quickly as possible. So you have this odd situation where the educated, "advanced" civilized people are trying to cut everyone's throats as quickly as possible and the indigenous, less educated, poorer populations are trying to prevent the disaster. If somebody was watching this from Mars, they'd think this species was insane.

Q: As far as a free, democracy-centered society, self-organization seems possible on small scales. Do you think it is possible on a larger scale and with human rights and quality of life as a standard, and if so, what community have you visited that seems closest to an example to what is possible?

A: Well, there are a lot of things that are possible. I have visited some examples that are pretty large scale, in fact, very large scale. Take Spain, which is in a huge economic crisis. But one part of Spain is doing okay -\/- that's the Mondragón collective. It's a big conglomerate involving banks, industry, housing, all sorts of things. It's worker owned, not worker managed, so partial industrial democracy, but it exists in a capitalist economy, so it's doing all kinds of ugly things like exploiting foreign labor and so on. But economically and socially, it's flourishing as compared with the rest of the society and other societies. It is very large, and that can be done anywhere. It certainly can be done here. In fact, there are tentative explorations of contacts between the Mondragón and the United Steelworkers, one of the more progressive unions, to think about developing comparable structures here, and it's being done to an extent.

The one person who has written very well about this is Gar Alperovitz, who is involved in organizing work around enterprises in parts of the old Rust Belt, which are pretty successful and could be spread just as a cooperative could be spread. There are really no limits to it other than willingness to participate, and that is, as always, the problem. If you're willing to adhere to the task and gauge yourself, there's no limit.

Actually, there's a famous sort of paradox posed by David Hume centuries ago. Hume is one of the founders of classical liberalism. He's an important philosopher and a political philosopher. He said that if you take a look at societies around the world -\/- any of them -\/- power is in the hands of the governed, those who are being ruled. Hume asked, why don't they use that power and overthrow the masters and take control? He says, the answer has to be that, in all societies, the most brutal, the most free, the governed can be controlled by control of opinion. If you can control their attitudes and beliefs and separate them from one another and so on, then they won't rise up and overthrow you.

That does require a qualification. In the more brutal and repressive societies, controlling opinion is less important, because you can beat people with a stick. But as societies become more free, it becomes more of a problem, and we see that historically. The societies that develop the most expansive propaganda systems are also the most free societies.

The most extensive propaganda system in the world is the public relations industry, which developed in Britain and the United States. A century ago, dominant sectors recognized that enough freedom had been won by the population. They reasoned that it's hard to control people by force, so they had to do it by turning the attitudes and opinions of the population with propaganda and other devices of separation and marginalization, and so on. Western powers have become highly skilled in this.

In the United States, the advertising and public relations industry is huge. Back in the more honest days, they called it propaganda. Now the term doesn't sound nice, so it's not used anymore, but it's basically a huge propaganda system which is designed very extensively for quite specific purposes.

First of all, it has to undermine markets by trying to create irrational, uninformed consumers who will make irrational choices. That's what advertising is about, the opposite of what a market is supposed to be, and anybody who turns on a television set can see that for themselves. It has to do with monopolization and product differentiation, all sorts of things, but the point is that you have to drive the population to irrational consumption, which does separate them from one another.

As I said, consumption is individual, so it's not done as an act of solidarity -\/- so you don't have ads on television saying, "Let's get together and build a mass transportation system." Who's going to fund that? The other thing they need to do is undermine democracy the same way, so they run campaigns, political campaigns mostly run by PR agents. It's very clear what they have to do. They have to create uninformed voters who will make irrational decisions, and that's what the campaigns are about. Billions of dollars go into it, and the idea is to shred democracy, restrict markets to service the rich, and make sure the power gets concentrated, that capital gets concentrated and the people are driven to irrational and self-destructive behavior. And it is self-destructive, often dramatically so. For example, one of the first achievements of the U.S. public relations system back in the 1920s was led, incidentally, by a figure honored by Wilson, Roosevelt and Kennedy -\/- liberal progressive Edward Bernays.

His first great success was to induce women to smoke. In the 1920s, women didn't smoke. So here's this big population which was not buying cigarettes, so he paid young models to march down New York City's Fifth Avenue holding cigarettes. His message to women was, "You want to be cool like a model? You should smoke a cigarette." How many millions of corpses did that create? I'd hate to calculate it. But it was considered an enormous success. The same is true of the murderous character of corporate propaganda with tobacco, asbestos, lead, chemicals, vinyl chloride, across the board. It is just shocking, but PR is a very honored profession, and it does control people and undermine their options of working together. And so that's Hume's paradox, but people don't have to submit to it. You can see through it and struggle against it.

\end{document}

