% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Noam Chomsky}
\title{ This Is The Most Remarkable Regional Uprising That I Can Remember  }
\date{2011, február 3 \thanks{Demo } }
\maketitle

In recent weeks, popular uprisings in the Arab world have led to the ouster of Tunisian dictator Zine El Abidine Ben Ali, the imminent end of Egyptian President Hosni Mubarak's regime, a new Jordanian government, and a pledge by Yemen's longtime dictator to leave office at the end of his term. We spoke to MIT Professor Noam Chomsky on Wednesday's live program about the situation in Egypt, and then continued the interview for another 50 minutes after the show to further discuss what these popular uprisings mean for the future of the Middle East and U.S. foreign policy in the region, how U.S. fear of the Muslim Brotherhood is really fear of democracy in the Arab world, and what the Egyptian protests mean for people in the United States.

\textbf{AMY GOODMAN:} Noam Chomsky, you were just talking about the significance of what's happening in the Middle East, and you were bringing it back to President Dwight Eisenhower.

\textbf{NOAM CHOMSKY:} Well, in 1958, Eisenhower---this is in internal discussions, since declassified---Eisenhower expressed his concern for what he called the "campaign of hatred against us" in the Arab world, not by the governments, but by the people. Remember, 1958, this was a rather striking moment. Just two years before, Eisenhower had intervened forcefully to compel Israel, Britain and France to withdraw from their invasion of Egyptian territory. And you would have expected enormous enthusiasm and support for the United States at that moment, and there was, briefly, but it didn't last, because policies returned to the norm. So when he was speaking two years later, there was, as he said, a "campaign of hatred against us." And he was naturally concerned why. Well, the National Security Council, the highest planning body, had in fact just come out with a report on exactly this issue. They concluded that, yes, indeed, there's a campaign of hatred. They said there's a perception in the Arab world that the United States supports harsh and brutal dictators and blocks democracy and development, and does so because we're interested in---we're concerned to control their energy resources.

\textbf{AMY GOODMAN:} Noam, I wanted to go for a minute to that famous address of the general, of the Republican president, of the president of the United States, Dwight D. Eisenhower.

\begin{quote}
\textbf{PRESIDENT DWIGHT D. EISENHOWER:} My fellow Americans, this evening I come to you with a message of leave-taking and farewell and to share a few final thoughts with you, my countrymen. We have been compelled to create a permanent armaments industry of vast proportions. Three-and-a-half million men and women are directly engaged in the defense establishment. The total---economic, political, even spiritual---is felt in every city, every state house, every office of the federal government. We recognize the imperative need for this development, yet we must not fail to comprehend its grave implications. In the councils of government, we must guard against the acquisition of unwarranted influence, whether sought or unsought, by the military-industrial complex. The potential for the disastrous rise of misplaced power exists and will persist.
\end{quote}

\textbf{AMY GOODMAN:} That was President Dwight D. Eisenhower in his farewell address in 1961. Special thanks to Eugene Jarecki and his film \emph{Why We Fight}, that brought it to us in the 21st century. Noam Chomsky, with us on the phone from his home near Boston, Noam, continue with the significance of what Eisenhower was saying and what the times were there and what they have to teach us today about this Middle East uprising.

\textbf{NOAM CHOMSKY:} Yeah, the military-industrial complex speech, the famous one, was after what I've just been talking about. That was as he was leaving office and a important speech, of course. Needless to say, the situation he described not only persists but indeed has amplified.

It should be mentioned that there's another element to the military-industrial complex issue, which he didn't bring up. At that time, in the 1950s, as he certainly knew, the Pentagon was funding what became---a lot of Pentagon funding was going into creating what became the next phase of the high-tech economy at that time: computers, micro-electronics, shortly after, the internet. Much of this developed through a Pentagon subsidy funding procurement, other mechanisms. So it was a kind of a cover for shifting---for a basic theme of contemporary economic development. That is, the public pays the costs and takes the risks, and eventual profit is privatized, in the case of computers and the internet, after decades. So that's another aspect of the military-industrial complex which is worth keeping in mind.

But Eisenhower was speaking particularly about the military aspect, what's called "defense," though in fact it's mostly aggression, intervention, subversion. It doesn't defend the country; it harms it, most of the time. But that's separate from the---not, of course, unrelated, but distinct from the Middle East problem. There, what Eisenhower and the National Security Council were describing is a persistent pattern. He was describing---they were describing it in 1950. And I'll repeat the basic conclusion: the United States does support brutal and harsh dictatorships, blocks democracy and development; the goal is to maintain control over the incomparable energy resources of the region---incidentally, not to use them. The U.S.---one of the things that Eisenhower was doing at exactly the same time was pursuing a program to exhaust U.S. energy reserves, rather than using much cheaper Middle East energy, for the benefit of Texas oil producers. That's a program that went on from the late '50s for about 15 years. So, at the time, it was not a matter of importing oil from Saudi Arabia, but just ensuring the maintenance of control over the world's major energy resources. And that, as the National Security Council concluded correctly, was leading to the campaign of hatred against us, the support for dictators, for repression, for violence and the blocking of democracy and development.

Now, that was the 1950s. And those words could be written today. You take a look at what's happening in the Middle East today. There's a campaign of hatred against the United States, in Tunisia against France, against Britain, for supporting brutal, harsh dictators, repressive, vicious, imposing poverty and suffering in the midst of great wealth, blocking democracy and development, and doing so because of the primary goal, which remains to maintain control over the energy resources of the region. What the National Security Council wrote in 1958 could be restated today in almost the same words.

Right after 9/11, the \emph{Wall Street Journal}, to its credit, did a---ran a poll in the Muslim world, not of the general population, of the kind of people they are interested in, I think what they called the moneyed Muslims or some phrase like that---professionals, directors of multinational corporations, bankers, people deeply embedded in the whole U.S.-dominated neoliberal project there---so not what's called anti-American. And it was an interesting poll. In fact, the results were very much like those that were described in 1958. There was tremendous---there wasn't a campaign of hatred against the U.S. among these people, but there was tremendous antagonism to U.S. policies. And the reasons were pretty much the same: the U.S. is blocking democracy and development; it's supporting dictators. By that time, there were salient issues that---some of which didn't exist in 1958. For example, there was a tremendous opposition in these groups to the murderous sanctions in Iraq, which didn't arouse much attention here, but they certainly did in the region. Hundreds of thousands of people were being killed. The civilian society was being destroyed. The dictator was being strengthened. And that did cause tremendous anger. And, of course, there was great anger about U.S. support for Israeli crimes, atrocities, illegal takeover of occupied territories and so on, settlement programs. Those were other issues, which also, to a limited extent, existed in '58, but not like 2001.

So that---and in fact, right now, we have direct evidence about attitudes of the Arab population. I think I mentioned this on an earlier broadcast, strikingly not reported, but extremely significant. Now, last August, the Brookings Institute released a major poll of Arab opinion, done by prestigious and respected polling agencies, one of them. They do it regularly. And the results were extremely significant. They reveal that there is again, still, a campaign of hatred against the United States. When asked about threats to the region, the ones that were picked, near unanimously, were Israel and the United States---88 percent Israel, about 77 percent the United States, regarded as the threats to the region. Of course, they asked about Iran. Ten percent of the population thought Iran was a threat. In the list of respected personalities, Erdogan was first. I think there were about 10. Neither Obama or any other Western figure was even mentioned. Saddam Hussein had higher respect.

Now, this is quite striking, especially in the light of the WikiLeaks revelations. The most---the one that won the headlines and that was---led to great enthusiasm and euphoria was the revelation, whether accurate or not---we don't know---but the claim, at least, by diplomats that the Arab dictators were supporting the U.S. in its confrontation with Iran. And, you know, enthusiastic headlines about how Arab states support---the Arabs support the United States. That's very revealing. What the commentators and the diplomats were saying is the Arab dictators support us, even though the population is overwhelming opposed, everything's fine, everything's under control, it's quiet, they're passive, and the dictators support us, so what could be a problem? In fact, Arab opinion was so antagonistic to the United States in this---as revealed in this poll, that a majority of the Arab population, 57 percent, actually thought the region would be better off if Iran had nuclear weapons. Nevertheless, the conclusion here, and in England and the continent, was it's all wonderful. The dictators support us. We can disregard the population, because they're quiet. As long as they're quiet, who cares? People don't matter. Actually, there's an analog of that internal to the United States. And it's of course the same policy elsewhere in the world. All of that reveals a contempt for democracy and for public opinion which is really profound. And one has to listen with jaws dropping when Obama, in the clip you ran, talks about how, of course, governments depend on the people. Our policy is the exact opposite.

\textbf{AMY GOODMAN:} Noam Chomsky, I wanted to read to you what Robert Fisk has written from the streets of Cairo today. Robert Fisk, the well-known reporter from \emph{The Independent} of London. He said, "One of the blights of history will now involve a U.S. president who held out his hand to the Islamic world and then clenched his fist when it fought a dictatorship and demanded democracy." Noam Chomsky, your response?

\textbf{NOAM CHOMSKY:} Well, Fisk's reporting, as usual, has been inspiring and phenomenal. And yeah, he's exactly right. And it is the old pattern. As I say, it goes back 50 years right there in Egypt and the region, and it's the same elsewhere. As long as the population is passive and obedient, it doesn't matter if there's a campaign of hatred against us. It doesn't matter if they believe that our official enemy can perhaps save them from our attacks. In fact, nothing matters, as long as the dictators support us. That's the view here.

We should remember there's an analog here. I mean, it's not the same, of course, but the population in the United States is angry, frustrated, full of fear and irrational hatreds. And the folks not far from you on Wall Street are just doing fine. They're the ones who created the current crisis. They're the ones who were called upon to deal with it. They're coming out stronger and richer than ever. But everything's fine, as long as the population is passive. If one-tenth of one percent of the population is gaining a preponderant amount of the wealth that's produced, while for the rest there 30 years of stagnation, just fine, as long as everyone's quiet. That's the scenario that has been unfolding in the Middle East, as well, just as it did in Central America and other domains.

\textbf{AMY GOODMAN:} Noam, I wanted to ask you if you think the revelations from WikiLeaks,---right?---the U.S. diplomatic cables, before that, Iraq and Afghan war logs, this massive trove of documents that have been released, Julian Assange talking about the critical issue of transparency---have played a key role here. I mean, in terms of Tunisia, a young university graduate who ended up, because there were no jobs, just selling vegetables in a market, being harassed by police, immolates himself---that was the spark. But also, the documents that came out on Tunisia confirming the U.S. knowledge, while it supported the Tunisian regime, that it was wholly corrupt, and what this means from one country to another, Yemen, as well. Do you think there is a direct relationship?

\textbf{NOAM CHOMSKY:} Well, actually, the fact of the matter is that WikiLeaks are not really telling us anything dramatically new. They're providing confirmation, often, of reasonable surmises. Tunisia was a very interesting case. So the ambassador did have a---one of the leaks comes from the ambassador, July 2009, and he describes Tunisia. He says it's a police state with little freedom of expression or association, serious human rights problems, ruled by a dictator whose family is despised for their corruption, robbery of the population and so on. That's the assessment of the ambassador. Not long after that, the U.S. singled out Tunisia for an extra shipment of military aid. Not just Tunisia, also two other Arab dictatorships---Egypt and Jordan---and of course Israel---it's routine---and one other country, namely Colombia, the country with the worst human rights record in the western hemisphere for years and the leading recipient of U.S. military aid for years, two elements that correlate quite closely, it's been shown.

Well, this tells you what the understanding was about Tunisia---namely, police state, a bitterly hated dictator and so on. But we send them more arms afterwards, because the population is quiet, so everything's fine. Actually, there was a description by---a very succinct account of all of this by a former high Jordanian official who's now director of Middle East research for the Carnegie Endowment, Marwan Muasher. He said, "This is the principle." He said, "There is nothing wrong. Everything is under control." Meaning, as long as the population is quiet, acquiescent---maybe fuming with rage, but doing nothing about it---everything's fine, there's nothing wrong, it's all under control. That's the operative principle.

\textbf{AMY GOODMAN:} He's a former Jordanian diplomat.

\textbf{NOAM CHOMSKY:} Former Jordanian official, high official.

\textbf{AMY GOODMAN:} What about what's happening now in Jordan, what you think is going to happen, and also in Saudi Arabia, how much it drives this and what you feel Obama needs to do and what you think he actually is doing?

\textbf{NOAM CHOMSKY:} Well, Jordan, the prime minister was just replaced. He was replaced with an ex-general who seems to be---is claimed to be moderately popular, at least not hated by the population. But essentially nothing changed. There are changes of the Jordanian cabinet frequently, and the basic system remains. Whether the population will accept that, whether the Muasher principle will work---nothing's wrong, everything's under control---that, we don't know.

Saudi Arabia is an interesting case. Saudi Arabia---the king of Saudi Arabia has been, along with Israel, the strongest supporter, most outspoken supporter of Mubarak. And the Saudi Arabian case should remind us of something about the regular commentary on this issue. The standard line and commentary is that, of course, we love democracy, but for pragmatic reasons we must sometimes reluctantly oppose it, in this case because of the threat of radical Islamists, the Muslim Brotherhood. Well, you know, there's maybe some---whatever one thinks of that. Take a look at Saudi Arabia. That's the leading center of radical Islamist ideology. That's been the source of it for years. The United States has---it's also the support of Islamic terror, the source for Islamic terror or the ideology that supports it. That's the leading U.S. ally, and has been for a long, long time. The U.S. supported---U.S. relations, close relations, with Israel, incidentally, after the 1967 war, escalated because Israel had struck a serious blow against secular Arab nationalism, the real enemy, Nasser's Egypt, and in defense of radical Islam, Saudi Arabia. Saudi Arabia and Egypt had been in a proxy war just before that, and there was a major conflict. And that's quite typical.

Probably the most---going back to WikiLeaks, maybe the most significant revelation has to do with Pakistan. In Pakistan, the WikiLeaks cables show that the ambassador, Ambassador Patterson, is pretty much on top of what's going on. There's enormous---the phrase "campaign of hatred against the United States" is an understatement. The population is passionately anti-American, increasingly so, largely, as she points out, as a result of U.S. actions in both Afghanistan and Pakistan, the pressure on the Pakistani military to invade the tribal zones, the drone attacks and so on. And she goes on to say that this may even lead to the---what is in fact the ultimate nightmare, that Pakistan's enormous nuclear facilities, which incidentally are being increased faster than anywhere else in the world, that these---there might be leakage of fissile materials into the hands of the radical Islamists, who are growing in strength and gaining popular support as a result of---in part, as a result of actions that we're taking.

Well, this goes back to---this didn't happen overnight. The major factor behind this is the rule of the dictator Zia-ul-Haq back in the 1980s. He was the one who carried out radical Islamization of Pakistan, with Saudi funding. He set up these extremist madrassas. The young lawyers who were in the streets recently shouting their support for the assassin of the political figure who opposed the blasphemy laws, they're a product of those madrassas. Who supported him? Ronald Reagan. He was Reagan's favorite dictator in the region. Well, you know, events have consequences. You support radical Islamization, and there are consequences. But the talk about concern about the Muslim Brotherhood in Egypt, whatever its reality, is a little bit ironic, when you observe that the U.S. and, I should say, Britain, as well, have traditionally supported radical Islam, in part, sometimes as a barrier to secular nationalism.

What's the real concern is not Islam or radicalism; it's independence. If the radical Islamists are independent, well, they're an enemy. If secular nationalists are independent, they are an enemy. In Latin America, for decades, when the Catholic Church, elements of the Catholic Church, were becoming independent, the liberation theology movement, they were an enemy. We carried out a major war against the church. Independence is what's intolerable, and pretty much for the reasons that the National Security Council described in the case of the Arab world 50 years ago.

\textbf{AMY GOODMAN:} Noam Chomsky, I wanted to read to you what two people are writing. One is Ethan Bronner in the \emph{New York Times}, saying, "Despite {[}Mr.{]} Mubarak's supportive relations with Israel, many Israelis on both the left and right are sympathetic {[}to{]} the Egyptians' desire to rid themselves of his autocracy and build a democracy. But they fear what will follow if things move too quickly." He quotes a top Israeli official saying, "We know this has to do with the desire for freedom, prosperity and opportunity, and we support people who don't want to live under tyranny, but who will take advantage of what is happening in its wake?" The official goes on to say, "The prevailing sense here is that you need a certain stability followed by reform. Snap elections are likely to bring a very different outcome," the official said.

And then there's Richard Cohen, who's writing in the \emph{Washington Post}. And Richard Cohen writes---and let me see if I can find this clip. Richard Cohen writes that---let's see if I can find it---"Things are about to go from bad to worse in the Middle East. An Israeli-Palestinian peace agreement is nowhere in sight."

Noam Chomsky, your response?

\textbf{NOAM CHOMSKY:} The comment of the Israeli official is standard boilerplate. Stalin could have said it. Yes, of course, the people want peace and freedom, democracy; we're all in favor of that. But not now, please. Because we don't like what the outcome will be. In fact, it's worth bearing---in the case---it's the same with Obama. It's more or less the same comment. On the other hand, the Israeli officials have been vociferous and outspoken in support of Mubarak and denunciation of the popular movement and the demonstrations. Perhaps only Saudi Arabia has been so outspoken in this regard. And the reason is the same. They very much fear what democracy would bring in Egypt.

After all, they've just seen it in Palestine. There has been one free election in the Arab world, exactly one really free election---namely, in Palestine, January 2006, carefully monitored, recognized to be free, fair, open and so on. And right after the election, within days, the United States and Israel announced publicly and implemented policies of harsh attack against the Palestinian people to punish them for running a free election. Why? The wrong people won. Elections are just fine, if they come out the way we want them to.

So, if in, say, Poland under Russian rule, popular movements were calling for freedom, we cheer. On the other hand, if popular movements in Central America are trying to get rid of brutal dictatorships, we send---we arm the military and carry out massive terrorist wars to crush it. We will cheer Václav Havel in Czechoslovakia standing up against the enemy, and at the very same moment, elite forces, fresh from renewed training at Fort Bragg, North Carolina, under command of the military, blow the brains out of six leading Latin American intellectuals, Jesuit priests, in El Salvador. That passes in silence. But those are the---that's exactly the pattern that we see replicated over and over again.

And it's even recognized by conservative scholarship. The leading studies of---scholarly studies of what's called "democracy promotion" happen to be by a good, careful scholar, Thomas Carruthers, who's a neo-Reaganite. He was in Reagan's State Department working on programs of democracy promotion, and he thinks it's a wonderful thing. But he concludes from his studies, ruefully, that the U.S. supports democracy, if and only if it accords with strategic and economic objectives. Now, he regards this as a paradox. And it is a paradox if you believe the rhetoric of leaders. He even says that all American leaders are somehow schizophrenic. But there's a much simpler analysis: people with power want to retain and maximize their power. So, democracy is fine if it accords with that, and it's unacceptable if it doesn't.

\textbf{AMY GOODMAN:} Noam, there's a sign, a big banner that people are holding in the square, in Tahrir, that says, "Yes, we can, too."

\textbf{NOAM CHOMSKY:} Let's what? I'm sorry, I didn't hear.

\textbf{AMY GOODMAN:} The banner says, "Yes, we can, too."

\textbf{NOAM CHOMSKY:} Oh, "Yes, we can, too." Yeah. You know where they got that from. Well, except that they mean it. Whether they can or not, no one knows. I mean, the situation has---we should recognize, has ominous aspects. The dispatch of pro-Mubarak thugs to the square is dangerous and frightening. Mubarak, presumably with U.S. backing, feels that---clearly feels that he can reestablish control. They've opened the internet again. The army is sitting by. We don't know what they'll do. But they might very well use the conflicts in the streets, caused by the pro-Mubarak gangs that have been sent in, to say, "Well, we have to establish military control," and they'll be another form of the military dictatorships that have been, you know, the effective power in Egypt for a long time.

Another crucial is how long the demonstrators can sustain themselves, not only against terror and violence, but also just against economic crisis. Within a short time, maybe beginning already, there isn't going to be bread, water. The economy is collapsing. They have shown absolutely incredible courage and determination, but, you know, there's a limit to what human flesh can bear. So, amazing as all this is, there's no guarantee of success.

If the United States, the population of the United States, Europe---if there is substantial vocal, outspoken support, that could make a difference. Now, remember the Muasher principle: as long as everyone's quiet, everything's under control, it's all fine. But when they break those bonds, it's not fine. You have to do something.

\textbf{AMY GOODMAN:} If you were president today, what would you do right now, president of the United States?

\textbf{NOAM CHOMSKY:} Well, if I were---if I had made it to the presidency, meaning with the kind of constituency and support that's required to be a president in the United States, I'd probably do what Obama's doing. But what ought to be done is what Erdogan is doing. Turkey is becoming the most significant country in the region, and it's recognized. Erdogan is far and away the most popular figure. And they've taken a pretty constructive role on many issues. And in this case, he is the one leading public figure, leader, who has been frank, outspoken, clear, and says Mubarak must go now. Now is when we must have change. That's the right stand. Nothing like that in Europe, and nothing like that here.

\textbf{AMY GOODMAN:} And what do you think of the role of the U.S. corporations? We spoke to Bill Hartung, who wrote this book, \emph{Prophets of Power}, P-R-O-P-H-E-T-S, about Lockheed Martin. The overwhelming amount of money, the billions, that have gone to Egypt, haven't really gone to Egypt; they've gone to U.S. weapons manufacturers, like General Dynamics, like Lockheed Martin, like Boeing, etc. In fact, Boeing owns Narus, which is the digital technology that's involved with surveillance of the cell phone, of the internet system there, where they can find dissident voices for the Egyptian regime. And who knows what they will do with those voices, just among others? But these corporations that have made such a killing off the repression, where are they standing right now in terms of U.S. policy?

\textbf{NOAM CHOMSKY:} Well, they don't issue press releases, so we have to speculate. But it's pretty obvious that they have a major stake in the dictatorships, not just Egypt. So, for example, a couple of months ago, Obama announced the biggest military sale in history to Saudi Arabia, \$60 billion worth of jet planes, helicopters, armored vehicles and so on and so forth. The pretext is that we have to defend Saudi Arabia against Iran. Remember that among the population, if anyone cares about them, 10 percent regard Iran as a threat, and a majority think the region would be better off if Iran had nuclear weapons. But we have to defend them against Iran by sending them military equipment, which would do them absolutely no good in any confrontation with Iran. But it does a lot of good for the American military-industrial complex that Eisenhower was referring to in that clip you ran a while back. So, yes, William Hartung was quite right about this.

In fact, a part of the reason why there is such strong support for Israel in the military lobby, in the military-industrial lobby in the United States, is that the massive arms transfers to Israel, which, whatever they're called, end up essentially being gifts, they go from the U.S.---the pocket of the U.S. taxpayer into the pocket of military industry. But there's also a secondary effect, which is well understood. They're a kind of a teaser. When the U.S. sends, you know, the most advanced jet aircraft, F-35s, to Israel, then Saudi Arabia says, "Well, we want a hundred times as much second-rate equipment," which is a huge bonanza for military industry, and it also recycles petrodollars, which is an important---a necessity for the U.S. economy. So these things are quite closely tied together.

And it's not just military industry. Construction projects, development, telecommunications---in the case of Israel, high-tech industry. So, Intel Corporation, the major---the world's major chip producer, has announced a new generation of chips, which they hope will be the next generation of chips, and they're building their main factory in Israel. Just announced an expansion of it. The relations are very close and intimate all the way through---again, in the Arab world, certainly not among the people, but we have the Muasher principle. As long as they're quiet, who cares? We can disregard them.

\textbf{AMY GOODMAN:} And the significance of Mubarak in the Israel-Palestine-Egypt axis? I mean, going back to 1979, if you could briefly remind people why he's so important, as the media keeps saying he has meant peace and stability with Israel, he gives the U.S. access to their air space, he guarantees access to the Suez Canal. Talk about that and what the change would mean.

\textbf{NOAM CHOMSKY:} We should actually go back a little further. In 1971, President Sadat of Egypt offered Israel a full peace treaty in return for withdrawal from the Occupied Territories. He cared about the Sinai, not---but Israel considered it, rejected it. Henry Kissinger, national security adviser, supported the rejection. State Department then supported Sadat. And Israel---it was a fateful decision. That's the point at which Israel quite explicitly chose expansion over security. They were then expanding into the Sinai, planning to build a city of a million people, Egyptian Sinai, settlements driving farmers out into the desert and so on. Well, that was the background for the 1973 war, which made it clear that Egypt can't simply be dismissed. Then we move on to the negotiations which led, in 1979, to the U.S. and Israel pretty much accepting Sadat's offer of 1971: withdrawal from the Sinai in return for a peace treaty. That's called a great diplomatic triumph. In fact, it was a diplomatic catastrophe. The failure to accept it in 1971 led to a very dangerous war, suffering, brutality and so on. And finally, the U.S. and Israel essentially, more or less, accepted it.

Now, as soon as that settlement was made, 1979, Israeli strategic analysts---the main one was Avner Yaniv, but others, too---recognized right away that now that Egypt is excluded from the confrontation, Israel is free to use force in other areas. And indeed, it very soon after that attacked Lebanon, didn't have to worry about an Egyptian deterrent. Now, that was gone, so we can attack Lebanon. And that was a brutal, vicious attack, killed 15,000, 20,000 people, led finally to the Sabra-Shatila massacre, destroyed lots of---most of southern Lebanon. And no defensive rationale. In fact, it wasn't even pretended. It was an effort to---as it was said, it was a war for the West Bank. It was an effort to block embarrassing Palestinian negotiation, diplomatic offers, and move forward on integrating the Occupied Territories. Well, they were free to do that once the Egyptian deterrent was gone. And that continues. Egypt is the major Arab state, the biggest military force by far, and neutralizing Egypt does free Israel---and when I say Israel, I mean the United States and Israel, because they work in tandem---it frees them to carry out the crimes of the occupation, attacks on Lebanon---there have been five invasions already, there might be another one---and Egypt does not interfere.

Furthermore, Egypt cooperates in the crushing of Gaza. That terrible free election in January 2006 not only frightened the U.S. and Israel---they didn't like the outcome, so turned instantly to punishing the Palestinians---but the same in Egypt. The victor in the election was Hamas, which is an offshoot of the Muslim Brotherhood. That was very much feared by the Egyptian dictatorship, because if they ever allowed anything like a free election, the Muslim Brotherhood would no doubt make out quite well, maybe not a majority, but it would be a substantial political force. And they don't want that, so therefore they cooperate. Egypt, under Mubarak, cooperates with Israel in crushing {[}Gaza{]}, built a huge fence on the Egyptian border, with U.S. engineering help, and it sort of monitors the flow of goods in and out of Gaza on the Egyptian side. It essentially completes the siege that the U.S. and Israel have imposed. Well, all of that could erode if there was a democratic movement that gained influence in Egypt, just as it did in Palestine.

I should mention that there's one other semi-democratic election in the Arab world, regularly. Now, that's in Lebanon. Lebanon is a complex story. It's a confessional democracy, so the Shiite population, which is the largest of the sects, is significantly underrepresented under the confessional system. But nevertheless the elections are not just state elections under dictatorships. And they have outcomes, too, which are suppressed here. So, for example, in the last election, the majority, a popular majority, was the Hezbollah-led coalition. They were the popular majority in the last election. I think about 53 percent. Well, that's not the way it was described here. If you read, say, Thomas Friedman, he wrote an ode about the election about---he was practically shedding tears of joy at free elections, in which Obama won over Ahmadinejad. Well, you know, what he meant is that in the representation under the confessional system, which seriously underrepresents the Shiite population, the pro-U.S. coalition won the most seats. That again reflects the standard contempt for democracy. All we care---we don't care that the majority of the population went the other way, as long as they're quiet and passive. And interestingly, Hezbollah quietly accepted the outcome, didn't protest about it at the time. But since then, their power has increased, and now there's a serious threat in Lebanon, which we should not overlook.

\textbf{AMY GOODMAN:} Noam, finally, as we wrap up, I've asked you a lot about what this means for the Middle East, this rolling revolution, from Tunisia to Egypt, what we're seeing in Jordan, in Yemen and beyond. But what about what these mass protests mean for people in the United States?

\textbf{NOAM CHOMSKY:} I think they mean a lot, and I've been trying to hint about that. The doctrine that everything is fine as long as the population is quiet, that applies in the Middle East, applies in Central America, it applies in the United States. For the last 30 years, we have had state-corporate policies specifically designed---specifically designed, not accidentally---to enrich and empower a tiny sector of the population, one percent---in fact, one-tenth of one percent. That's the basic source of the extreme inequality. Tax policies, rules of corporate governance, a whole mass of policies, have been very explicitly designed to achieve this end---deregulation and so on. Well, for most of the population, that's meant pretty much stagnation over a long period. Now, people have been getting by, by sharply increasing the number of work hours, far beyond Europe, by debt, by asset inflation like the recent housing bubble. But those things can't last.

And as soon as Obama came into office, he came in in the midst of the worst crisis since the Depression. In fact, Ben Bernanke, we know from recent testimony that was released, head of the Fed, said it was even worse than the banking crisis in 1929. So there was a real crisis. Who did he pick to patch up the crisis? The people who had created it, the Robert Rubin gang, Larry Summers, Timothy Geithner, basically the people who were responsible for the policies that led to the crisis. And it's not surprising. I mean, Obama's primary constituency was financial institutions. They were the core of the funding for his campaign. They expect to be paid back. And they were. They were paid back by coming out richer and more powerful than they were before the crisis that they created.

Meanwhile, the population, much of the population, is literally in depression. If you look at the unemployment figures, among the top few percent, maybe 10, 20 percent, unemployment is not particularly high. In fact, it's rather low. When you go down to the bottom of the income ladder, you know, the lower quintiles, unemployment is at Depression levels. In manufacturing industry, it is at Depression levels.

And it's different from the Depression. In the Depression, which I'm old enough to remember, it was very severe. My own family was mostly unemployed working class. But there was a sense of hopefulness. Something is---we can do something. There's CIO organizing. There's sitdown strikes, that compelled New Deal measures, which were helpful and hopeful. And there was a sense that somehow we'll get out of this, that we're in it together, we can work together, we can get out of it. That's not true now. Now there's a general atmosphere of hopelessness, despair, anger and deep irrationality. That's a very dangerous mix. Hatred of foreigners, you know, a mix of attitudes which is volatile and dangerous, quite different from the mood in the Depression.

But the same governing principle applies: as long as the population is---accepts what's going on, is directing their anger against teachers, you know, firemen, policemen, pensions and so on, as long as they're directing their anger there, and not against us, the rulers, everything's under control, everything's fine. Until it erupts. Well, it hasn't erupted here yet, and if it does erupt, it might not be at a constructive direction, given the nature of what's happening in the country now. But yes, those Egyptian lessons should be taken to heart. We can see clearly what people can do under conditions of serious duress and repression far beyond anything that we face, but they're doing it. If we don't do it, the outcome could be quite ugly.

\textbf{AMY GOODMAN:} Noam Chomsky, I want to thank you very much for being with us. Noam, author, Institute Professor Emeritus at MIT, and most recent book, \emph{Hopes and Prospects}, has written more than a hundred books.

\textbf{AMY GOODMAN:} For analysis of the Egyptian uprising and its implications for the Middle East and beyond, we're joined now by the world-renowned political dissident and linguist Noam Chomsky, Professor Emeritus at Massachusetts Institute of Technology, author of over a hundred books, including his latest, \emph{Hopes and Prospects}.

\end{document}

