% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Noam Chomsky}
\title{ Global economic crisis  }
\date{2009, április 11 \thanks{Demo } }
\maketitle

\begin{abstract}
  
\end{abstract}

\textbf{AMY GOODMAN: }Today, a conversation with one of the most important dissident intellectuals of our time, Noam Chomsky, on the global economic crisis, healthcare, the media, US foreign policy, the expanding wars in Afghanistan and Pakistan, and resistance to American empire. Noam Chomsky is a world-renowned linguist, philosopher, social critic, and Institute Professor Emeritus at the Massachusetts Institute of Technology.

Among his many books over the past few decades are \emph{Hegemony or Survival: America's Quest for Global Dominance}, \emph{Manufacturing Consent: The Political Economy of the Mass Media}, \emph{Profit over People: Neoliberalism and Global Order}, and \emph{Human Rights and American Foreign Policy}. There's a great collection of his work, just out now, edited by Anthony Arnove, called \emph{The Essential Chomsky}.

I spoke to Noam Chomsky earlier this month when we were on the road in Boston. This is Part II of our conversation. I began by asking him to talk about the current economic meltdown.

\textbf{NOAM CHOMSKY: } Well, let's start with G20. If you look at the Financial Times, the world's major business journal, the day before the G20 meeting, they had a section on it, and they pointed out, I think correctly, that the main purpose is to present a picture of harmony and agreement. It doesn't matter what you do, but make it look as if we're all together on this. Now, there are sharp splits about how to approach the issue, but you have to make it look as if we're all together. That's pretty much what happened.

Now, in the communiqué, which you read before, the crucial word was \textqq{voluntary.} So, the countries there are supposed to voluntarily choose to do x, y and z. Well, that means we couldn't make an agreement. So we'll call it voluntary agreement.

Now, there was one point on which they agreed: a sharp recapitalization of the International Monetary Fund; pour a lot of money into the IMF. That's a pretty dubious move. I mean, the record of the IMF hasthe IMF is more or less a branch of the US Treasury, even though it has a European director. Its past role has been extremely destructive. In fact, its American US executive director captured its role when she described it as \textqq{the credit community's enforcer,} meaning if a third world dictator incurs a huge debtpeople didn't, but the dictator did; say, Suharto in Indonesiaand then the debt defaults, the lenders, who have made plenty of money because it was a risky loan so they get high interest and so on, they have to be protected, meaning not by the dictator, by the people of Indonesia, who are subjected to harsh structural adjustment programs so that they can pay back the debt, which they didn't incur, so that we can be compensated, rich Westerners can be compensated. So that's the IMF, the credit community's enforcer, a very destructive role in the third world. Now it's to be recapitalized.

Now, there's discussion about this, and it's interesting. You can read it in the financial pages. The supporters of the recapitalization say, \textqq{Well, the IMF has changed its spots. It's going to be different from now on. We realize that it had this terrible role, but now it's going to be different.} Well, is there any reason to believe it will be different? In fact, if you look today, it's quite striking to see the advice that the Western powers are following, the programs that they're following, and compare them to the instructions given to the third world.

So, say, take Indonesia again. Indonesia had a huge financial crisis about ten years ago, and the instructions were the standard ones: \textqq{Here is what you have to do. First, pay off your debts to us. Second, privatize, so that we can then pick up your assets on the cheap. Third, raise interest rates to slow down the economy and force the population to suffer, you know, to pay us back.} Those are the regular instructions the IMF is still giving them.

What do we do? Exactly the opposite. We forget about the debt, let it explode. We reduce interest rates to zero to stimulate the economy. We pour money into the economy to get even bigger debts. We don't privatize; we nationalize, except we don't call it nationalization. We give it some other name, like \textqq{bailout} or something. It's essentially nationalization without control. So we pour money into the institutions. We lectured the third world that they must accept free trade, though we accept protectionism.

Take the \textqq{too big to fail} principle, which the House committee is discussing today. But what does \textqq{too big to fail} mean? \textqq{Too big to fail} is an insurance policy. It's a government insurance policy. Government means the public pays, which says, \textqq{You can take huge risks and make plenty of profit, and if anything goes wrong, we'll bail you out.} That's \textqq{too big to fail.} Well, that's extreme protectionism. It gives US corporations like Citigroup an enormous advantage over others, like any other kind of protection.

But we don't allow the third world to do that. I mean, they've got to privatize, so that we can pick up their assets. Now, these are happening side by side. Now, here's the instructions for you, the poor people; here's the policies for us, the rich people. Exactly the opposite. Is there any reason to think the IMF is going to change it?

\textbf{AMY GOODMAN: } Do you think President Obama is any different than President Bush when it comes to the economy? And if you were in the Congress, would you have voted for the bailouts and the stimulus packages?

\textbf{NOAM CHOMSKY: } He's different. I mean, first of all, there's a rhetorical difference. But we have to distinguish the first and the second Bush terms. They were different. I mean, the first Bush term was so arrogant and abrasive and militaristic and dismissive of everyone that they offended, they antagonized even allies, close allies, and US prestige in the world plummeted to zero. Now, the second Bush administration was moremoved more toward the center in that respect, not entirely, but more, so some of the worst offenders, like Rumsfeld, Wolfowitz and others, were thrown out. I mean, they couldn't throw out Dick Cheney, because he was the administration, so they couldn't get rid of him. He stayed, but the others, a lot of them, left. And they moved towards a somewhat more normal position.

And Obama is carrying that forward. He's a centrist Democrat. He never really pretended to be anything else. And he's moving towards a kind of a centrist position. He's very popular in Europe, not so much because of him, but because he's not Bush. So there is the kind of rhetoric that the European leaders and, in fact, the European population tend to accept. In fact, you know, even in the Middle East, where you'd think people would know better, they accept the illusions. And they are illusions, because there's nothing to back them up. So, yes, he is different from Bush.

Sameon the economy, well, you know, the current Obama-Geithner plan is not very different from the Bush-Paulson plan. I mean, somewhat different, but circumstances have changed. So, of course, it's somewhat different. But it's still based on the principle that we have tosomehow, the taxpayer has to rescue the institutions intact. They have to remain intact, including the people who, you know, destroyed the economy. In fact, they are the ones who Obama picked to fix it up.

\textbf{AMY GOODMAN: } Explain.

\textbf{NOAM CHOMSKY: } Like Larry Summers, for example, who is now his chief economic adviser. I mean, he was Secretary of Treasury under Bill Clinton. His great achievement was to prevent Congress from regulating derivatives, exotic financial instruments. Well, that's one of the main factors that led to the crisis.

His kind of senior adviser, one of the first, was Robert Rubin, who was Secretary of Treasury right before Summers. His main achievementmany achievements, like what he did to Indonesia and the third world, but here, his main achievement was to lead the way to revoke the Glass-Steagall legislation from the New Deal, which protected commercial banks from risky investments. It broke down those barriers. Immediately after having done this, he left the government, joined Citigroup as a director, and they began to make huge profits, including him, from picking up insurance companies and so on and making very risky loans, relying on the \textqq{too big to fail} doctrine, meaning if we get in trouble, the taxpayer will bail us out, which is just what's happening, taxpayers now pouring tens of billions of dollars into rescuing Citigroup.

Well, these are the advisers who were supposed to fix up the system. Tim Geithner was right in the middle of this. He was head of the New York Federal Reserve, so, yes, he was supervising these actions. Now, you know, you can argue about whether they're doing the right thing or the wrong thing, but are these the people who should be fixing up the system?

Actually, the business press just had some interesting things to say about this. Bloomberg News, you know, main business press, had an article in which they reviewed the records of the people who Obama invited to his economic summit. I think it must have been last November or December. They just reviewed the record. I think there were a couple dozen of them. People on theyou know, people like, say, Stiglitz, Krugman, they were never even allowed close to it, let alone anyone from the left or labor and so on, given token representation. So they went through the records, and they concluded that these people should not be invited to fix up the economy. Most of them should be getting subpoenas because of their record of accounting fraud, malpractice and so on, and helping bring about the current crisis.



\textbf{AMY GOODMAN: } Professor Noam Chomsky. We'll continue the conversation in a minute. If you'd like a copy of today's show, you can go to our website at democracynow.org. Stay with us.

[break]

\textbf{AMY GOODMAN: } We return now to my conversation with Noam Chomsky about the economic crisis and how the Obama administration is handling it.

\textbf{AMY GOODMAN: } Why do you think Obama chose to surround himself?

\textbf{NOAM CHOMSKY: } Because those are his beliefs. I mean, his support comes from thehis constituency is basically the financial institutions. Just take a look at the funding for his campaign. I mean, the final figures haven't come out, but we have preliminary figures, and it seems to be mostly financial institutions. I mean, the financial institutions preferred him to McCain. They are the main funders for bothyou know, I mean, core funders for both parties, but considerably more to Obama than McCain.

You can learn a lot from campaign contributions. In fact, one of the best predictors of policy around is Thomas Ferguson's investment theory of politics, as he calls itvery outstanding political economistwhich essentiallyI mean, to say it in a sentence, he describes elections as occasions in which groups of investors coalesce and invest to control the state. And he takes a look at the formation of campaign contributors, and it gives you a surprisingly good prediction of what policies are going to be. It goes back a century, New Deal and so on. So, yeah, it can predict pretty well what Obama is going to do. There's nothing surprising about this. It's the norm in what's called political democracy.

\textbf{AMY GOODMAN: } Would you have let Citibank, would you have let Citigroup, would do have let the AIG fail?

\textbf{NOAM CHOMSKY: } Well, there are other possibilities. So, the government could just take over the viable parts. And parts of them are functioning; parts are dysfunctional, like the toxicwhat they call the toxic asset parts, you know, the financial manipulations.

Well, one thing you could do, which has been suggested by a number of economists like Dean Baker, just take over the good parts, essentially nationalize them, put them under public control. And \textqq{nationalize} means public control, at least if you have a democracy. Not here, but if you had a functioning democracy, it would mean let them be under public control, and the parts that are responsible for the huge losses, just let them go off by themselves. In fact, that would be the way of taking care of the AIG bonuses that everyone's screaming about. In fact, as Baker pointed out, just spin off the parts that were involved in financial manipulations and caused the crisis, let them go bankrupt and let the executives try to get the bonuses from a bankrupt firm, OK, with no legislation necessary. That's what should be done with Citigroup.

And in fact, it's interesting, it's kind of happening. You know, after the breakdown of Glass-Steagall, they did bring inthey made use of it, under Rubin's direction, among others, to takebring in insurance companies and other risky investors. Now they're divesting them. And they're going in the direction of becoming, you know, commercial bank.

Now, incidentally, this is not the first time this has happened. Paul Volcker is on the news today, you know, saying, \textqq{Let's slow down,} and so on. Well, he's the one who, under Reagan, who helped bail out Citigroup last time they crashed. At that time they were Citibank. They had followed World Bank and IMF instructions and lent huge amounts of money to Latin America and were assured by the World Bank that it's all fine, you know, markets will take care of it, etc. Well, in a crash, Paul Volcker came in. He raised interest rates very sharply. Third world countries, whose payments are tied to US interest rates, couldn't pay their debts. The IMF moved in, took care of it, and essentially recapitalized Citibank. That's the way the system works: you make risky loans, you make a lot of money, and if you get into trouble, we're here to bail you out, namely the taxpayer.

\textbf{AMY GOODMAN: } And how do the Republicans differ from the Democrats in this? What do you make ofdo you see it as just a minor footnote that Republicans, or some of the governors like Palin, like Jindal

\textbf{NOAM CHOMSKY: } There's a difference.

\textbf{AMY GOODMAN: } are saying they're not going to take stimulus money?

\textbf{NOAM CHOMSKY: } There's a difference. I mean, we basically are a kind of a one-party state. I think C. Wright Mills must have pointed this out fifty years ago. It's a business party, but it has factionsDemocrats and Republicansand they're different. They have somewhat different constituencies and different policies. And if you look over the years, the population hasthe majority of the population has tended to make out better under Democrats than Republicans; the very wealthy have tended to make out better under Republicans than Democrats. So they're business parties, but they're somewhat different, and the differences can have an effect. However, fundamentally, they're pretty much along the same lines.

So take, say, the current financial crisis. Actually, it began under Carter. The late Carter administration is the one that beganwas pushing for financialization of the economy, you know, huge growth of speculative financial capital, deregulation, and so on. Reagan carried it much further, and Clinton continued it. And then, with Bush, it kind of went off the rails.

So there are differences, but differences within a pretty narrow spectrum. And anyone who's a little off the spectrum, like Nobel laureates in economics who are a couple of millimeters off the spectrum, they're basically on the outside. You can interview them, but they don't show up at the economic summit.

\textbf{AMY GOODMAN: } How does the global economy and our own economy relate to the issue of war and US foreign policy?

\textbf{NOAM CHOMSKY: } Well, actually, you had a pretty good interview with Joseph Stiglitz about that a couple of months ago, in which he discussed the relationship ofhe was talking about the Iraq war. And as you'll recall, he pointed out correctly that the Iraq war, which, first of all, is going to cost trillions of dollars, also had the effect of sharply increasing the price of oil, predictably. And as he pointed out, we could sort of paper that over for a while by a housing bubble, so there was a huge housing bubble which anyone with eyes open could see. I mean, for a century, housing prices had sort of tracked the economy, GDP; then, all of a sudden, they shot way beyond the trend line, which means there's a bubble, and it's going to burst, and you get into trouble. But the housing bubble, which was supervised by Alan Greenspan and with the Democratsactually, it started under Clintonit replaced the tech bubble under Clinton, and it gave an illusion of prosperity, whichso you didn't see the effects of the rise in oil prices, which went very high. But if you trace all the connections, yes, there's a clear connection, as he pointed out, between the war and the economic crisis.

And in fact, it's deeper than that. The US is just in a class by itself in military expenses. It basically matches the rest of the world, and it's far more advanced. Well, that's drawn from somewhere. You know, that's money that's not being used to develop the economy.

Now, in fact, you have to add a footnote here, because part of the very high level of US violation of free trade principles is that the economy itself is based on military spending to a substantial extent. So the modern information revolutioncomputers, the internet, fancy software and so onmost of that comes straight out of the Pentagon. My own university, MIT, was one of the places where all of this was developed under Pentagon contracts in the 1950s and the 1960s.

In fact, that's another critical part of the way the economy works. The public pays the costs and takes the risk of economic development, and if anything works, maybe decades later, it's handed over to private enterprise to make the profits. And that's a core element of the economy. Of course, we don't permit the third world to do that. That's considered a violation of free trade when they do it. But it's the way our economy works. And it's kind of complementary to the \textqq{too big to fail} doctrine of protectionism for financial institutions. But the generalwe do not have a capitalist economy. We have kind of a state capitalist economy in which the public has a role: pay the costs, take the risks, bail out if they get into trouble. And the private sector has a role: make profit, and then turn to the public if you get into trouble.

\textbf{AMY GOODMAN: } Would you extend that to healthcare?

\textbf{NOAM CHOMSKY: } Well, healthcare is a dramatic case. I mean, for decades, the healthcare issue has been right at the top of domestic concerns, for very good reasons. The US has the most dysfunctional healthcare system in the industrial world, has about twice the per capita costs and some of the worst outcomes. It's also the only privatized system. And if you look closely, those two things are related. And the privatized system is highly inefficient: a huge amount of administration, bureaucracy, supervision, you know, all kinds of things. It's been studied pretty carefully.

Now, the public has had an opinion about this for decades. A considerable majority want a national healthcare system, like other industrial countries have. They usually say a Canadian-style system, not because Canada is the best, but at least you know that Canada exists. Nobody says an Australian-style system, which is much better, because who knows anything about that? But something like what's sometimes called Medicare Plus, like extend Medicare to the population.

Well, up untilit's interesting. Up until the year 2004, that idea was described, for example, by the New York Times as politically impossible and lacking political support. So, maybe the public wants it, but that's not what counts as political support. The financial institutions are opposed, the pharmaceutical institutions are opposed, so it's notno political support. Well, in 2008, for the first time, the Democratic candidatesfirst Edwards, then the othersbegan to move in the direction of what the public has wanted, not there, but in that direction.

So what happened between 2004 and 2008? Well, public opinion didn't change. It's been this way for decades. What changed is that manufacturing industry, a big sector of the economy, has recognized that it's being severely harmed by the highly inefficient privatized health system. So, General Motors said that it costs them over a thousand dollars more to produce a car in Detroit than across the border in Windsor, Canada. And, you know, when manufacturing industry becomes concerned, then things become politically possible, and they begin to have political support. So, yes, in 2008, there's some discussion of it.

Now, you know, this is very revealing insight into how American democracy functions and what is meant by the term \textqq{political support} and \textqq{politically possible.} Again, this should be headlines. Will a proposal come that approaches what the public wants? Well, we're already getting the backlash, strong backlash. And what private healthcare systems are claiming is that this is unfair. The government is so much more efficient that they'll be driventhere's no level playing field if the government gets into it, which is true.

\textbf{AMY GOODMAN: } If you had a public and a private plan.

\textbf{NOAM CHOMSKY: } Yeah.

\textbf{AMY GOODMAN: } If it were like Medicare.

\textbf{NOAM CHOMSKY: } If you had them side by side

\textbf{AMY GOODMAN: } Most people go for Medicare

\textbf{NOAM CHOMSKY: } they will.

\textbf{AMY GOODMAN: } but if you wanted to go for a private plan, you could.

\textbf{NOAM CHOMSKY: } Yeah, if you could. But they're notthey say, \textqq{Well, we can't compete.} For good reasons. I mean, in every country exceptindustrial country except the United States, the government uses its massive purchasing power to negotiate drug prices. That's one of the reasons prices are so much higher in the United States than in other countries. Well, they couldthe Pentagon can use purchasing power to negotiate prices for, you know, paper clips or something, but, by law, the government is not permitted to do that in the case of healthcare. Well, if you had Medicare Plus, they would, and that would drive down drug prices, and the private industries can't compete.

\textbf{AMY GOODMAN: } FAIR, Fairness and Accuracy in Reporting, did a study of the week leading up to the White House healthcare summit of the networks and how they were covering single payer, the issue of like Medicare Plus, and I think they found that absolutelythat almostthere was almost no representation in the media of a single-payer advocate

\textbf{NOAM CHOMSKY: } Yeah.

\textbf{AMY GOODMAN: } and almost the only mention was someone blasting single payer.

\textbf{NOAM CHOMSKY: } Yeah, yeah. That's because it has no political support; only the majority of the public. It's the same as the media commentary in 2004. In fact, if you take a look back at the end of the last electoral campaign, Kerry-Bush campaign, in October 2004, right before the election, there was a debate on domestic issues. I think it was maybe October 28th or so. Just take a lookread the New York Times report of it the next day. It was very dramatic. It said Kerry never brought up the idea of any government involvement in healthcare, not, you know, Medicare Plus, but any government involvement, because it is not politically possible and lacks political supportjust the population. Well, that

\textbf{AMY GOODMAN: } What studies show you the population wants this?

\textbf{NOAM CHOMSKY: } I mean, there's been poll after poll, goes back, in fact

\textbf{AMY GOODMAN: } So, what do you think is going to break through?

\textbf{NOAM CHOMSKY: } Well, it's a problem of the general dysfunction of formal democracy. I mean, there's a very substantial gap between public opinion and public policy on a host of major issues. And on many of these issues, both parties are well to the right of the public, international and domestic.

Incidentally, that's one reason why elections are run the way they are. Elections are run as marketing extravaganzas, and that's not kept secret. So the advertising industry gives an award every year for best marketing campaign of the year. For 2008, they gave it to Obama. He beat out, I think, Apple Computer. And if you look at the comments of financialof advertising executives, PR executives, they were euphoric. In fact, they saidyou can read it in the Financial Times, business pressthey said, you know, \textqq{We've been marketing candidates like commodities ever since Reagan, but this is the best we've ever done. It's going to change the atmosphere in corporate boardrooms. We have a new style of selling things, you know, the Obama style, you know, soaring rhetoric, hope and change, and so on.} Yeah, that's true.

And if you look at the campaigns themselves, they're designed essentially by the advertising industry to sell the commodityit happens to be a candidateand they're pretty carefully designed so that you marginalize issues and you focus on what are called \textqq{qualities.} In Obama's case, you know, soaring rhetoric and so on; in Bush's case, a nice guy and like to have a beer with him and so on. That's the kind of thing you focus on. Where do they stand on issues? Well, the public is mostly uninformed. I haven't seen current polls on 2008, but the 2004 election, where there were polls shortly after, showed the public had almost no idea what Bush's stand was. In fact, a majority of Bush voters thought that he supported the Kyoto Protocol, because they support it, and he's a nice guy, so he must support it.

And elections are designed that way, and it makes good sense. I mean, the people who run the elections, they read the polls, and very carefully, in fact. In fact, they mostly the design them for their own interest. And they know that the parties are to the right of the public, so you betteron a large number of issues, including crucial ones like Iran and othersso you better keep issues off the table, which is what's done. So what thehealthcare is a dramatic case of it, but it's only one instance.



\textbf{AMY GOODMAN: } Renowned linguist Noam Chomsky, speaking to me in Boston last week. We will return to the last part of our conversation after this break. You can get a copy of the full two parts by going to democracynow.org. Stay with us.

[break]

\textbf{AMY GOODMAN: } We return now to the last part of my conversation with leading American intellectual and anti-imperialist critic Noam Chomsky.

\textbf{AMY GOODMAN: } The whole issue of populist rage, Noam Chomsky, actually, do you think that this rage is going to boil over as the unemployment figures rise?

\textbf{NOAM CHOMSKY: } It's very hard to predict those things. I mean, it has a potentially positive side, like it could be like the activism of the 1930s or the 1960s, which ended up making it a more civilized society in many ways, or it could be like an unfortunate precedent that quickly comes to mind. I've written about it.

Take a look at Germany. In the 1920s, Germany was the absolute peak of Western civilization, in the arts and the sciences. It was regarded as a model of democracy and so on. I mean, ten years later, it was the depths of barbarism. That was a quick transition. \textqq{The descent into barbarism} it's sometimes called in the scholarly literature.

Now, if you listen to early Nazi propaganda, you know, end of the Weimar Republic and so on, and you listen to talk radio in the United States, which I often doit's interestingthere's a resemblance. And in both cases, you have a lot of demagogues appealing to people with real grievances.

Grievances aren't invented. I mean, for the American population, the last thirty years have been some of the worst in economic history. It's a rich country, but real wages have stagnated or declined, working hours have shot up, benefits have gone down, and people are in real trouble and now in very real trouble after the bubbles burst. And they're angry. And they want to know, \textqq{What happened to me? You know, I'm a hard-working, white, God-fearing American. You know, how come this is happening to me?}

That's pretty much the Nazi appeal. The grievances were real. And one of the possibilities is what Rush Limbaugh tells you: \textqq{Well, it's happening to you because of those bad guys out there.} OK, in the Nazi case, it was the Jews and the Bolsheviks. Here, it's the rich Democrats who run Wall Street and run the media and give everything away to illegal immigrants, and so on and so forth. It sort of peaked during the Sarah Palin period. And it's kind of interesting. It's been pointed out that of all the candidates, Sarah Palin is the only one who used the phrase \textqq{working class.} She was talking to the working people. And yeah, they're the ones who are suffering. So, there are models that are not very attractive.

\textbf{AMY GOODMAN: } And she very much is being talked about as a leader, really, of the Republican Party.

\textbf{NOAM CHOMSKY: } Well, she was kind of a model. You know, the talk radio mob went crazy over her. And one shouldn't demean it. You know, they describe themselvesit's really worth listening to: \textqq{We're fly-by country. You know, they don't care about us, those rich Democrats on the East Coast and the West Coast who are all, you know, interested in gay rights and giving things away to illegal immigrants and so on. They don't care about us, the hard-working, God-fearing people, so we've got to somehow rise up and take over and elect Sarah Palin or Rush Limbaugh or someone like that.}

As I say, the precedents are not attractive. Now, ifnow even before the next presidential, if in the next congressional election the economy has not begun to recover, this kind of populist rage could boil over and could have very dangerous consequences. This country has a long history of being kind of ridden by fear. It's a very frightened country. This goes back to colonial times.

I mean, we're very lucky that we have never had an honest demagogue. I mean, the demagogues we've had are so corrupt that they never got anywhereyou know, Nixon, McCarthy, you know, Jimmy Swaggart and others. So they were kind of destroyed by their own corruption.

But suppose we had an honest demagogue, you know, a Hitler type, who was not corrupt. There's probablyit could be unpleasant. There's a background of concern and fear, tremendous fear, and searching for some answer, which they're not getting from the establishment. \textqq{Who's responsible for my plight?} You know, and that can be exploited. And unless there's active, effective organizing and education, it's dangerous.

\textbf{AMY GOODMAN: } Your assessment of President Obama so far?

\textbf{NOAM CHOMSKY: } Frankly, I never had any expectations. I wrote about it over a year ago. I thought then, and I think it's been confirmed, that he's essentially a centrist Democrat. He's moving backI mean, the Bush administration was kind of off the spectrum, especially the first term. So he's moving things back toward the center with a kind of a public posture, which was recognized by the advertising industry. That's why they gave him the award for best marketing campaign, whichbut as far as policy is concerned, unless he's under a lot of pressure from activist sectors, he's not going to go beyond what he's presented himself as in actual policy statements or cabinet choices and so on: a centrist Democrat, going to basically continue Bush's policies, maybe in a more modulated way.

\textbf{AMY GOODMAN: } Do you see Afghanistan becoming an ever-expanded war in the next decade or so? Do younow we're talking about doubling the US troops there.

\textbf{NOAM CHOMSKY: } No, that's the way Obama and the Pentagon see it. In fact, they say so: this is going to be a long war, it's going to be extended, the US is going to take over the military side, and it's going to expand it, it's going to expand into Pakistan. And, I mean, we'll talk about development, but the focus will be on the military. Obama, right now, is trying to get NATO to cooperate, but recognizing that they're not going to send military forces. The populations are opposed.

\textbf{AMY GOODMAN: } Canada is pulling out.

\textbf{NOAM CHOMSKY: } Yeah, Canada's pulling out, and the othersmaybe Holland has made a termination date, but we'll at least ask them to come in and sort of help out on the civilian side. That's their job. It's the famous line of, I guess it was Robert Kagan: you know, \textqq{they're Venus, we're Mars.} So we'll move in like Mars and take care of the military side. You know, we're good at killing people. And they can come in and sort of put on the band-aids and make it look like something good is happening. It's not the right direction.

\textbf{AMY GOODMAN: } The unmanned drones bombing Pakistan?

\textbf{NOAM CHOMSKY: } Yeah, drones. And that has effects. So a lot of the worst fighting recently has been in the Bajaur province, right on the border. It's in Pakistan's side. And militants in the area have reported to the press that part of the reason is that an American drone attack hit a madrasa, a school, and killed about eighty people. Well, you know, they're \textqq{uncivilized barbarians}; they sort of don't like that. So they reacted. And now, one of the militants has said, \textqq{OK, we're going to bomb the White House,} which is considered totally outrageous. But, you know, if we kill as we like, there's going to be a reaction.

\textbf{AMY GOODMAN: } Where do you see American empire in ten, twenty, thirty years?

\textbf{NOAM CHOMSKY: } Prediction in human affairs is a very lowhas very little success, too many complications. The United States, I think, will come out of the economic crisis, very likely, as the dominant superpower. There's a lot of talk about China and India, and it's real, they're changing, but they're just not in the same league. I mean, both China and India have enormous internal problems that the West doesn't face.

You get kind of a picture of this by looking at the Human Development Index of the United Nations. The last time I looked, India was about 125th or something. And I think China was about eightieth. And China would be worse, I think, if it wasn't such a closed society. In India, you sort of get better data, so you can see what's happening. China is kind of closed. You don't see what's going on in the peasant areas, which are in turmoil, you know. They have environmental problems. They have hugehundreds of millions of people are kind of like at the edge of starvation.

We don't haveyou know, we have problems, but not those problems. And even the industrial growth, which is thereyou know, for part of the population, there's been improvement. But when you take, say, India, where we know more, in the areas where high-tech industries developedand it's pretty impressive. I've visited some of the labs in Hyderabad. You know, it's as good or better than MIT. But right nearby, the rate of peasant suicides is going up, very sharply, in fact. And it's the same source. It's the neoliberal policies, which privilege a certain sector of the population and a certainand let the rest take care of themselves.

\textbf{AMY GOODMAN: } And yet, the rise of progressives in Latin America?

\textbf{NOAM CHOMSKY: } That's important. I mean, Latin America, for the first time in 500 years, is moving towards a degree of independence and a kind of integration, which is a prerequisite for independence, and also at least is beginning to face some of its massive internal problems. I mean, Latin America has probably the worst inequality in the world. There's a wealthy sector, small wealthy sector, which is extremely rich, but they havetheir tradition is that they have no responsibility to the country, so they send their capital to Zurich. You know, they have their second homes in the Riviera, and their children study in Oxford or whatever. This is beginning to be faced in different ways, but it's sort of happening all over the continent. And they are beginning to integrate. The United States obviously doesn't like it. In fact, it's barely reported most of the time.

So there was a very interesting case last September, when President Morales in BoliviaBolivia is, in my opinion at least, probably the most democratic country in the world. Nobody says that, but if you look at what happened in the last couple of years, there were huge, popular, mass organizations of the most repressed population in the hemisphere, the indigenous population, which for the first time ever has entered the political arena significantly and were able to elect a president from their own ranks and one who doesn't give instructions to his army, but who's following policies that were largely produced by the population. So he's their representative, in a sense in which democracy is supposed to work.

And they know the issues. It's not like our elections. They know the issues. They're serious issues: control over resources, economic justice, cultural rights, and so on. You can say they're right or wrong, but at least it's functioning.

Now, the elites that have traditionally ruled the country, of course, don't like it. And they're threatening virtual secession. And, of course, the United States is backing them, as the media are. And it got to the point last summer, I suppose, where it led to real violence.

Well, there was a meeting of UNASUR, the Union of South American Republicsthat's all of South Americaa meeting in Chile, Santiago, Chile. And it came out with a declaration, important declaration, in which it supported President Morales and opposed thecondemned the violence being led by the quasi-secessionist forces. And Morales responded, thanking them for their gesture of support, but also saying, correctly, that this is the first time in 500 years that South America is beginning to take its affairs in its own hands without the intervention of foreign powers, primarily the US.

Well, that was so important that I don't think it was even reported here. I mean, the meeting was known, so you see vague references to it. But it's an indication of developments that are taking place in various ways.

\textbf{AMY GOODMAN: } Noam Chomsky, you've just hit eighty. We just have a few minutes to go. And how does it feel?

\textbf{NOAM CHOMSKY: } I have a few years to go. I don't think about it much.

\textbf{AMY GOODMAN: } But as you reflect, talking about these huge social movements, cataclysmic times in the world, your life experience, what gives you hope?

\textbf{NOAM CHOMSKY: } Well, there's both hope and fear. I mean, I'm old enough to have grown up in the Depression. And some of my memoriesI didn't understand that much at the timechildhood memories, are listening to Hitler's speeches. I didn't understand them, but I could sense the reaction of my parents, you know, and had a feeling of fear, you know, a tremendous fear. In fact, the first article I wrote was in 1939, when I was in fourth grade, and it was about the expansion of fascism over Europe, a kind of a dark cloud that may envelop everything. And as I mentioned before, I have some of those same concerns now.

On the other hand, there's been tremendous progress. The country is far more civilized than it was, say, forty years ago, thanks to the activism of the '60s and its aftermath. And some of the most important developments were after the '60s, like, say, the feminist movement, which has probably had more of an impact on this society than any other. It's mostly post-'60s. The solidarity movements, which are unique in the history of imperialism, there's never been anything like them. That's from the '80s. The global justice movements, what's called anti-globalizationshouldn't bethat's, you know, the '90s and this century. These were all very positive developments.

They haven't changed the institutions. In fact, the institutions have reacted by becoming harsher, not surprisingly. But they've changed the culture. I mean, take, say, the 2008 election. I mean, I didn't like the candidates, as I've made clear. On the other hand, forty years ago, or maybe ten years ago, you couldn't have imagined that the Democratic Party would have two candidates, an African American and a woman. OK, that's a sign of the civilizing effect of the activism of the '60s and everything that followed.

Well, that can be mobilized. In fact, it's already. If you count the number of activists in the country, it's, I suspect, well beyond the '60s, except maybe for a very brief moment at the peak of the antiwar movement. OK, that can be a basis for proceeding onward. So, that's a reason for hope.

\textbf{AMY GOODMAN: } And finally, our condolences on the loss of Carol.

\textbf{NOAM CHOMSKY: } Thanks.

\textbf{AMY GOODMAN: } Your life partner, someone you knewwell, you're eightywhat, for seventy-seven years?

\textbf{NOAM CHOMSKY: } Yeah, actually. Not easy to face.

\textbf{AMY GOODMAN: } What gives you the strength to go on after Carol?

\textbf{NOAM CHOMSKY: } Well, the kind of thing you do, for example. That makes a difference.

\textbf{AMY GOODMAN: } And you have a wonderful family.

\textbf{NOAM CHOMSKY: } Yeah.

\textbf{AMY GOODMAN: } So, our condolences to you

\textbf{NOAM CHOMSKY: } Thanks.

\textbf{AMY GOODMAN: } and your kids. Noam Chomsky, thanks so much.

\textbf{NOAM CHOMSKY: } Thanks.

\textbf{AMY GOODMAN: } Noam Chomsky, Professor Emeritus at MIT, leading public intellectual of our day. If you'd like to get a copy of the full interview, part one and today's part two, with Noam Chomsky, you can go to our website at democracynow.org.



\end{document}
