% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Noam Chomsky}
\title{ Noam Chomsky on Civil Rights, Obama, Latin America, and the History of the U.S. in the Middle East  }
\date{2011, február 20 \thanks{Demo } }
\maketitle

\begin{abstract}
\url{http://www.democracynow.org/2011/2/17/democracy_uprising_in_the_usa_noam}
\end{abstract}


\textbf{NOAM CHOMSKY:} We were talking about unions before. Union busting is criminal activity by the government, because they're saying, "You can go ahead and do it; we're not going to apply the laws," effectively. And the COINTELPRO, which you mentioned, is actually the worst systematic and extended violation of basic civil rights by the federal government. It maybe compares with Wilson's Red Scare. But COINTELPRO went on from the late '50 right through all of the '60s; it finally ended, at least theoretically ended, when the courts terminated it in the early '70s. And it was serious.

It started, as is everything, going after the Communist Party, then the Puerto Rican Independence Party. Then it extended---the women's movement, the New Left, but particularly black nationalists. And it ended up---didn't end up, but one of the events was a straight Gestapo-style assassination of two black organizers, Fred Hampton and Mark Clark, literally. The FBI set up the assassination. The Chicago police actually carried it out, broke into the apartment at 4:00 in the morning and murdered them. Fake information that came from the FBI about arms stores and so on. There was almost nothing about it. In fact, the information about this, remarkably, was released at about the same time as Watergate. I mean, as compared with this, Watergate was a tea party. There was nothing, you know?

\textbf{JUAN GONZALEZ:} I'd like to ask you---we're obviously entering very soon a new presidential season, and for many of the progressives and liberals who had placed some much hope in the Obama administration, they're now going to be faced with the quandary of what to do as they move into a new administration. On the one hand, they feel betrayed by many of the things the administration has done; on the other hand, they see this extreme right that is attempting to paint Obama as a socialist, as destroying the Constitution and freedom in America. And they're going to have to figure out how they're going to maneuver in this new reality, especially with the \emph{Citizens United} case, the enormous amount of money that's going to be poured into. Your thoughts on what progressives who are still glued to the ground and understand the reality of what's happening in the country should be doing?

\textbf{NOAM CHOMSKY:} Well, my feeling---actually, I had the same feeling in 2008. I'm not disillusioned, because I didn't have any expectations, just looking at the funding, looking at his background. Actually, I wrote about it before the primaries even. But nevertheless, you know, when I was asked in 2008, "Who should you vote for?" my own feeling was---and it will be next time---that if you're in a swing state, you better vote against the prehistoric monsters, because they're going to cause much more trouble. Well, in our system, the only choice you have would be to vote for Obama. Hold your nose and vote, but don't expect anything.

Just take a look at where he's coming from, where his funding is coming from. Over a long period, like a century, you can pretty well predict policies by just looking at concentration of campaign funding. Thomas Ferguson, very outstanding political scientist, has done the main work on this, and it's convincing. So, when you find that the core of the funding is the financial institutions, you can pretty well expect that the major policies will be to reward them. Yeah, OK, it's pretty much what happened. You shouldn't be disillusioned. But if you have to make a choice between that and, you know, Newt Gingrich, well, OK, you have to make that choice. Don't expect anything.

What has to be done is what's happening in Madison, or what's happening in Tahrir Square in Cairo. If there's mass popular opposition, any political leader is going to have to respond to it, whoever they are.

\textbf{AMY GOODMAN:} Noam Chomsky, I wanted to ask you about the situation in Haiti. The country is preparing to hold a controversial runoff presidential vote next month. The U.S. has resumed deportations to Haiti despite the earthquake-ravaged, cholera-ravaged country. And former Haitian president Jean-Bertrand Aristide has been given a passport that would allow him to return home seven years after he was ousted in a U.S.-backed coup. I wanted to go back just to a brief clip. I spoke to President Aristide at the time of the coup in 2004, and he talked about the role of the United States in Haiti and in the world.

\begin{quote}
\textbf{PRESIDENT JEAN-BERTRAND ARISTIDE:} They went to Iraq. We see how is the situation in Iraq. They went to Haiti. We see how is the situation in Haiti. Pretending imposing democracy, we saw people killing people. Why don't they change their approach to let democracy and the constitutional order flourish, slowly, but surely? After imposing an economic embargo on us, being, from the cultural point of view, very rich, from an historic point of view, very rich, but from an economic point of view, very poor, because we are the poorest country of the Western hemisphere, after imposing their economic embargo upon us, because the people wanted one man, one vote, so equality among us.
\end{quote}

\textbf{AMY GOODMAN:} That was President Aristide in 2004.

\textbf{NOAM CHOMSKY:} That's right. I mean, in 2004---I don't have to tell you---the United States and France, the two traditional torturers of Haiti for hundreds of years, joined by Canada, you know, plodding along, carried out a military coup. They kidnapped the president, sent him off to Central Africa. The United States has tried hard to keep him out of the hemisphere ever since, blocked him from coming back to Haiti.

This last election, which is a complete farce, I think about less than a quarter of the population voted even. By most accounts, Aristide is the most popular figure in Haiti. He's kept out of the election. His party, Fanmi Lavalas, has easily won every election in which there was even a modicum of sort of honesty. They were kept out of the election by the United States. So we have an election in which the most popular political figure is out, most popular political party is out, country is a total wreck, people can't get registration cards. I mean, that's a total ruin. A lot of money was pledged; very little of it has actually been allocated. Having an election under those conditions, it doesn't rise to the level of a joke. There was an OAS, Organization of American States, investigation, but if you look at the people on the commission, it's mostly the United States or its puppets, totally unserious. You can't even laugh about it. I mean, Haiti, once again, is being denied the possibility of having a democratic election.

Now, it's not the first time. The first real democratic election in Haiti was in, 20 years ago, 1990. To everyone's amazement, Aristide won. Everyone assumed---me, too---that the U.S. candidate would win. Former World Bank official, he had all the money, all the elite support.

\textbf{AMY GOODMAN:} Marc Bazin.

\textbf{NOAM CHOMSKY:} He got 14 percent of the vote. You know, nobody was---it's kind of like Egypt and what Marwan was saying about the Middle East. Nobody is paying attention to what's going on in the slums and the hills, which happens to be where the population is. They're just paying attention to what's happening up in the rich sectors of Pétionville, you know, where the rich people live. Well, it turns out a lot of popular organizing was going on, a really impressive democratic achievement. It's something that I wish we could even come close to here: actual, real, live democracy.

And they swept into office, with a big majority, a populist priest who immediately initiated programs which were in fact pretty constructive. They were in fact highly praised, even by the international financial institutions, you know, which don't usually go for this. He cut back corruption. He fixed up the budget. Well, you know, just kind of waiting, and it took seven months for the military coup to come, which threw him out.

The OAS declared an embargo. The U.S. technically joined the embargo, but within weeks the government, that was Bush number one, announced that U.S. firms would be exempt from the embargo. I remember the \emph{New York Times}, that report, saying this is a very humanitarian gesture: the embargo is being fine-tuned for the interests of the people of Haiti, namely by exempting U.S. firms. It turned out---and trade increased. Actually, I was there during---it was a horrible terror. I was right during it. Maybe you were, too. It was just awful. The CIA was reporting that all---to Congress, that no oil is coming in. You could see the oil farms being built by the rich families. And in fact, it later turned out that first Bush, then Clinton, had authorized the Texaco Oil Company to ship oil to the military junta and to the elite in violation of presidential orders. Barely mentioned. The \emph{Wall Street Journal} had an article on it. And so it went on.

You know, every time there has been an effort by the Haitian people to overcome the misery and poverty that comes from 200 years of bitter attacks, really bitter, the U.S. steps in and blocks it. And that's what's happening now with this so-called election.

\textbf{JUAN GONZALEZ:} I'd like to ask you one final question on the U.S. situation. Yesterday the Federal Reserve Bank upped its prediction for growth in the United States. Corporations are getting record profits. The banks are back in great shape. The Dow is back up before it was in the crisis. And yet, we still have massive unemployment, conservatively estimated at nine percent, and we still have a huge mortgage crisis in the country, more and more people losing their homes. The disconnect between what the indicators are saying and the reality of what the American people are facing?

\textbf{NOAM CHOMSKY:} Well, actually, it depends how you look at the indicators. I mean, for the last 30 years, for a majority of the population, real incomes have pretty much stagnated. I mean, there's growth. And the growth is going to---and the wealth is going into very few pockets. That's by design. Tax laws are designed that way.

Take, say, the Bush tax cut. It was very cleverly done. It's devastating the economy, but not the rich. The tax cut was done so that at the beginning, the first---right off, everyone got a check, you know, so it looks, "Oh, great. We're getting a tax cut, couple hundred dollars." But it was designed so that over the years the cut would shift. By the end, over half the tax cut, I think, was going to maybe one percent of the population. But by then, nobody notices anymore. Well, that's the way fiscal policy has been designed. It's the way corporate governance rules have been designed. They come from the federal government. And they effectively give the CEO the right to pick the panel that gives him the salary, and all sorts of things like this, along with the deregulation, which---bipartisan, incidentally---which has led to a situation where, you know, maybe you can talk about growth on the average, but for most of the population it's not there.

For a large part, especially maybe the lower half of the population, they're basically living in the Depression. Not quite. I mean, I'm old enough to remember the Depression. My family was mostly unemployed working class. It was objectively worse than now, if you count, you know, objective standards. On the other hand, it was hopeful. There was a sense that something is going to happen. You had a government which was doing things that helped the population, because they were under pressure. In fact, Roosevelt famously talked to the labor leaders and said, "Make me do this. You know, so you go have sit-down strikes and you protest and so on, then we'll push this legislation through." Well, it happened. So you had WPA. You had---Social Security was coming in. There was a sense that we're going to get out of this somehow. There was hope for the future. Now there isn't. The industrial workforce is living in the Depression. Unemployment is at Depression levels.

And the jobs aren't coming back, because policy is designed, by the man in charge of jobs for the Obama administration and others like him, to send production abroad. It's cheaper. It's more profitable for the banks and the management. Or to move from investment in production to investment in finance, which does nothing for the economy, probably harms it, but it is very profitable and has the nice feature that when it crashes, as it's going to do, the taxpayer will come in and bail you out. It's a great system. It's a real racket. We will---the regulations are such so that we can take very risky transactions, make a lot of money, it's going to crash, but no problem, there's that nice taxpayer. They will come in and bail us out. We'll be richer than before. And each time it gets worse than it was the last time. Now, this one is really bad. So whatever the growth figures show, for the population, that's not happening, except for a small sector. So the numbers could be right, but that's not what it means for people's lives.

\textbf{AMY GOODMAN:} I wanted to ask you quickly about Vermont. You have Madison, Wisconsin. You have all of Wisconsin now, Scott Walker saying they'll break the unions, bring out the National Guard if the teachers and other union workers protest. In Vermont, the new governor, Peter Shumlin, has run on a platform of instituting single-payer healthcare immediately. And in January, a landmark measure was introduced to revoke the granting of personhood rights to U.S. corporations. The bill calls for a constitutional amendment declaring corporations are not persons under the laws of the United States. you live next door in Massachusetts. How significant is this whole movement?

\textbf{NOAM CHOMSKY:} The movement is significant, but of course it has to take root and spread. I mean, take, say, the personhood goes back a century, and it was not by law. No legislation saying corporations are persons. That was by court decisions, a series of court decisions over time which have given these fictitious legal entities---established by the state, incidentally, and protected by the state; they're basically state-based organizations---giving these entities more and more rights. It was bitterly attacked by conservatives when---because it was a big attack on the classical liberal ideals a century ago. \emph{Citizens United}, which you mentioned, is just the last state of it. So that's quite right.

The other thing about single payer is extremely significant. I mean, you know, we're supposed to be upset about the deficit. Whether we should be or not is another question. You should have a deficit in a recession. But let's say we're worried about the deficit. Where is the deficit coming from? About half of it is military spending, which is out of sight. You know, it's as much as the rest of the world combined. It's not for defense. In fact, it probably increases dangers to the United States. But it's there. The other half is our totally dysfunctional healthcare system. I mean, it costs about twice as much per capita as comparable countries and doesn't have---has pretty poor outcomes, plus 50 million uninsured and other scandals. And it's the only privatized, virtually unregulated healthcare system. So costs are out of sight. Administrative costs are very high. You have profits. You have cherry picking, all sorts of things that cost money. That's about half the deficit. In fact, if we had a healthcare system like comparable countries, we'd problably have a surplus.

Well, if you look at the debate that's going on---you know, you read \emph{New York Times}, anybody---they say the big problem is entitlements. Entitlements means Social Security, Medicare and Medicaid. Well, Social Security is just pure lies. I mean, Social Security doesn't even add to the deficit. It's funded by payroll taxes. And furthermore, it's in quite good shape for decades. So that's just mentioned in order to try to destroy Social Security. Social Security does nothing for the wealthy. It's a means of survival for working people and poorer people, so therefore let's get rid of it. Also, Social Security is dangerous. Social Security is based on a principle which is frightening: namely, we care about each other. So, Social Security is based on the idea that you care if a disabled widow across town has food to eat, and you have to drive that out of people's heads. They're supposed to care only about themselves, not anybody else, like part of the reason for attacking unions. So you've got to get rid of Social Security, so therefore lie about it. But what about Medicare and Medicaid? It's true, those expenses are going through the roof, and they're going to tank the federal budget. But that's because of the healthcare system. I mean, Medicare---

\textbf{AMY GOODMAN:} So, what could Obama do right now?

\textbf{NOAM CHOMSKY:} He could do what the population has wanted for years: put in a national healthcare system like every other industrial country has in one form or another. That was just given up during the healthcare---they didn't talk about it. There was one last residue of it in the healthcare reform: namely, the public option. The public was in favor of that by I think about five to three or something, substantially. That was just given away, you know? We've got to make sure that the rich---financial institutions are richer and richer---insurance companies, in this case.

Same with pharmaceutical corporations. Drug prices in the United States are much higher than comparable countries, with one exception: Veterans Administration. Veterans Administration has reasonable prices, and there's a reason. The government is allowed to bargain with pharmaceutical corporations for the VA, but not for the rest of the population. So, of course, the prices are out of sight. Well, yeah, the public has views on this, too. In fact, actually, it's only one poll. It showed about 85 percent opposition. But it's not even discussed.

So, yes, entitlements are a problem, but not the entitlements. What's a problem is paying off the insurance companies and paying off Big Pharma. That's a problem. And unless we do something about that, that problem is going to get worse and worse, and you'll have a bigger and bigger deficit, plus the military. So what Vermont is doing is picking the right problem. But, you know, it's a small state. What they can do depends on how---if we have a popular uprising like, say, Tunisia or Egypt or Bahrain, yeah, then you could get somewhere.

\textbf{AMY GOODMAN:} We just had this breaking news that the Obama administration is making a call to Bahrain to use restraint. I mean, again, we have the U.S. military base there. But did you ever think you'd see, Noam, in your lifetime, what we are seeing now in the Middle East, this rolling revolution?

\textbf{NOAM CHOMSKY:} No, not really. But then, I never expected to see what's happened in Latin America for the past 10 years. Over the past 10---what's happened in Latin America is very dramatic. There's 500 years of history here. And this is the first time, since the Spanish and Portuguese conquerors came, that Latin America has started to take its fate into its own hands. They've kicked out all the American military bases. The countries are integrating. They're beginning to deal with the absolute scandal, that's internal to each country, a tiny sector of extreme wealth, Western-oriented---

\textbf{JUAN GONZALEZ:} The wealth gap is shrinking.

\textbf{NOAM CHOMSKY:} And it's shrinking, and they're doing something about it. A long way to go, but at least being faced. And there also integration of the countries, which is a prerequisite for independence. Now, that's dramatic. It's far more significant than what happened in Eastern Europe. I mean, the comparisons to Eastern Europe I don't think are very convincing. First of all, in Eastern Europe, remember, you had Gorbachev. Now, the person who was in charge, basically, and had the guns was saying, "Go ahead." You don't have any Gorbachev in the West, nothing like him. Furthermore, in the case of Eastern Europe, the major power sectors in the world---the United States and Western Europe---were supporting the uprising, of course, because it was undermining an enemy. There's nothing like that here.

In fact, about the only comparison to Eastern Europe that isn't sort of ridiculous is the one that's never talked about: Romania. Romania, which had the worst dictator in Eastern Europe, Ceausescu, he was a darling of the West. The United States and Britain loved him. He was supported until the last minute. They couldn't support him anymore, so, you know, turned against him---the usual game plan. But that's about the only analogy.

On the other hand, what happened in Latin America is extremely significant, and what's happening in the Middle East could turn into something similar.

\textbf{AMY GOODMAN:} And the role of the U.S. in the Middle East, what it should play right now? And I know you have to go after that.

\textbf{NOAM CHOMSKY:} Well, what it should do is say, "OK, we're out of here. This is your country. You live there. You do what you want. We're going to support democracy. Or we'll support whatever comes out." But think what's involved in that. Just take a look at Arab public opinion. You take a look at Arab public opinion, you see democracy would be a disaster for the U.S. leadership. I mentioned the figures, but the whole of U.S. policy in the region would instantly collapse if you had democracy.

Well, you know, the Middle East is an important area. It goes back to almost 90 years, since oil was discovered, especially since the Second World War. Take a look at internal documents. The Middle East oil was regarded as the most important and strategic---the strategically most important area of the world, because it's got the major oil reserves. And if you think what's happened in the Middle East over the years, the big---the United States and Britain have traditionally supported radical Islamic fundamentalism. The core of radical Islamic fundamentalism is Saudi Arabia. That's also the main fundamental of jihadi terror. That's our main ally. In fact, in 1967, when U.S. relations with Israel took on their current form, the primary reason was that there was a war going on, literally, between Saudi Arabia and Egypt in Yemen. And Saudi Arabia is the center of radical Islamic fundamentalism. Egypt was the center of secular nationalism. And secular nationalism is frightening, because---it wasn't democratic, it was autocratic, but it was secular and it was nationalist. And Nasser was talking about using the resources of the region for their own populations, not to enrich Western oil companies and, you know, the Saudi elite. Well, that's frightening. So there was this conflict going on, the U.S. and Britain of course supporting radical Islamism. Israel won that battle for them. That's when relations were established in their current form, and that continues. It's not 100 percent, but substantially the U.S. and Britain have supported and continue to support radical Islam, because it's a barrier against democracy. If it goes the wrong way, they don't like it, but as long as it's going your way, fine.

Actually, during this entire crisis, I thought one of the most astute comments was a two-sentence comment by Marwan Muasher. He's a former high Jordanian official who's head of Middle East research for the Carnegie Endowment. He said, "There's an operative principle in the Middle East." He said, "The principle is, as long as people are quiet and passive, we'll do whatever we like." That's a general principle of statesmanship that applies here, too. As long as people are quiet and passive, we'll do whatever we like. Now, of course, if they stop being quiet and passive, we'll have to adjust somehow. Maybe they'll even throw us out, but we'll try to hang on as much as we can. And that's what we see going on in the Middle East. That's what we saw going on in Latin America. It's what we see right here.

\textbf{AMY GOODMAN:} Noam Chomsky, thank you so much for spending this time.

\end{document}

