% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Noam Chomsky}
\title{ U.S. Should Back Chemical Weapons Ban  }
\date{2013, szeptember 18 \thanks{Demo} }
\maketitle

\begin{abstract}
  Chomsky: Instead of "Illegal" Threat to Syria, U.S. Should Back Chemical Weapons Ban in All Nations
\end{abstract}



In a national address from the White House Tuesday night, President Obama announced he is delaying a plan to strike Syria while pursuing a diplomatic effort from Russia for international monitors to take over and destroy Syria's arsenal of chemical weapons. However, Obama still threatened to use force against Syria if the plan fails. We get reaction to Obama's speech from world-renowned political dissident and linguist, MIT Professor Emeritus Noam Chomsky. "The Russian plan is a godsend for Obama," Chomsky says. "It saves him from what would look like a very serious defeat. He has not been able to obtain virtually any international support, and it looked as though Congress wasn't going to support it either, which would leave him completely out on a limb. This leaves him a way out: He can maintain the threat of force, which incidentally is a crime under international law. We should bear in mind that the core principle of the United Nations Charter bars the threat or use of force. So all of this is criminal, to begin with, but he'll continue with that."

TRANSCRIPT

This is a rush transcript. Copy may not be in its final form.

\textbf{NERMEEN SHAIKH}: In a nationally televised address, President Obama announced he was putting off a plan to strike Syria while pursuing a diplomatic effort from Russia for international monitors to take over and destroy Syria's arsenal of chemical weapons. The speech came just 10 days after he told the nation he would ask Congress to authorize using military force. On Tuesday night, Obama asked congressional leaders to put off a vote on his request to authorize the use of military strikes, but he said the military would remain ready if diplomacy fails. \#

\textbf{PRESIDENT BARACK OBAMA}: America is not the world's policeman. Terrible things happen across the globe, and it is beyond our means to right every wrong. But when, with modest effort and risk, we can stop children from being gassed to death, and thereby make our own children safer over the long run, I believe we should act. That's what makes America different. That's what makes us exceptional. With humility, but with resolve, let us never lose sight of that essential truth. \#

\textbf{NERMEEN SHAIKH}: President Obama offered a qualified endorsement of the Russian proposal to secure Syria's chemical weapons arsenal. \#

\textbf{PRESIDENT BARACK OBAMA}: Over the last few days, we've seen some encouraging signs, in part because of the credible threat of U.S. military action, as well as constructive talks that I had with President Putin. The Russian government has indicated a willingness to join with the international community in pushing Assad to give up his chemical weapons. The Assad regime has now admitted that it has these weapons, and even said they'd join the chemical weapons convention, which prohibits their use. It's too early to tell whether this offer will succeed, and any agreement must verify that the Assad regime keeps its commitments. But this initiative has the potential to remove the threat of chemical weapons without the use of force, particularly because Russia is one of Assad's strongest allies.

I have therefore asked the leaders of Congress to postpone a vote to authorize the use of force while we pursue this diplomatic path. I'm sending Secretary of State John Kerry to meet his Russian counterpart on Thursday, and I will continue my own discussions with President Putin. I've spoken to the leaders of two of our closest allies---France and the United Kingdom---and we will work together in consultation with Russia and China to put forward a resolution at the U.N. Security Council requiring Assad to give up his chemical weapons and to ultimately destroy them under international control. We'll also give U.N. inspectors the opportunity to report their findings about what happened on August 21st. And we will continue to rally support from allies from Europe to the Americas, from Asia to the Middle East, who agree on the need for action. Meanwhile, I've ordered our military to maintain their current posture, to keep pressure on Assad and to be in a position to respond if diplomacy fails. \#


\textbf{AMY GOODMAN}: To talk more about President Obama's speech and the crisis in Syria, we're joined by the world-renowned political dissident, linguist, author, MIT Professor Noam Chomsky. He has authored numerous books. His latest is On Western Terrorism: From Hiroshima to Drone Warfare, that's out next week. He joins us via Democracy Now! video stream from his home in Massachusetts.

Noam, welcome to Democracy Now! First, let's get your response to President Obama announcing last night in a nationwide address, which I'm sure was watched worldwide, that for the moment there would be no strike on Syria, as the U.S. supports the Russian plan to deal with the chemical weapons stockpile of Syria? \#

\textbf{NOAM CHOMSKY}: Well, the Russian plan is a godsend for Obama. It saves him from what would look like a very serious political defeat. He has not been able to obtain virtually any international support for this---the action he's contemplating. Even Britain wouldn't support it. And it looked as though Congress wasn't going to support it either, which would leave him completely out on a limb. This leaves him a way out.

He can maintain the threat of force, which incidentally is a crime under international law, that we should bear in mind that the core principle of the United Nations Charter bars the threat or use of force, threat or use of force. So all of this is criminal, to begin with, but he'll continue with that. The United States is a rogue state. It doesn't pay any attention to international law.

He---it was kind of interesting what he didn't say. This would be a perfect opportunity to ban chemical weapons, to impose the chemical weapons convention on the Middle East. The convention, contrary to what Obama said, does not specifically refer just to use of chemical weapons; it refers to production, storage or use of chemical weapons. That's banned by the international norm that Obama likes to preach about. Well, there is a country which happens to be---happens to have illegally annexed part of Syrian territory, which has chemical weapons and is in violation of the chemical weapons convention and has refused even to ratify it---namely, Israel. So here's an opportunity to eliminate chemical weapons from the region, to impose the chemical weapons convention as it's actually formulated. But Obama was very careful not to say that he---for reasons which are too obvious to go into---he---and that gap is highly significant. Of course, chemical weapons should be eliminated everywhere, but certainly in that region.

The other things that he said were not unusual, but nevertheless kind of shocking to anyone not familiar with U.S. political discourse, at least. So he described the United---he said that for seven decades the United States has been "the anchor of global security." Really? Seven decades? That includes, for example, just 40 years ago today, when the United States played a major role in overthrowing the parliamentary democracy of Chile and imposing a brutal dictatorship, called "the first 9/11" in Latin America. Go back earlier years, overthrowing the parliamentary system in Iran, imposing a dictatorship; same in Guatemala a year later; attacking Indochina, the worst crime in the postwar period, killing millions of people; attacking Central America; killing---involved in killing---in imposing a dictatorship in the Congo; and invading Iraq---on and on. That's stability? I mean, that a Harvard Law School graduate can pronounce those words is pretty amazing, as is the fact that they're accepted without comment.

So what he said is I'm going to lie like a trooper about history; I'm going to suppress the U.S. role, the actual U.S. role, for the last seven decades; I'm going to maintain the threat of force, which is of course illegal; and I'm going to ensure that the chemical weapons convention is not imposed on the region, because our ally, Israel, would be subjected to it. And I think those are some of the main points of his address. \#

\textbf{AMY GOODMAN}: Noam Chomsky. Noam Chomsky, the world-renowned linguist, political dissident. We're going to go to break and then spend the hour with him on President Obama's policy and what's happening in the Middle East. This is Democracy Now!, democracynow.org, The War and Peace Report. Back in a minute.

\textbf{AMY GOODMAN}: Our guest for the hour is Professor Noam Chomsky. We're going to turn again back to President Obama, who addressed part of his speech to the nation last night to opponents of military action on the right and left. \#

\textbf{PRESIDENT BARACK OBAMA}: My fellow Americans, for nearly seven decades, the United States has been the anchor of global security. This has meant doing more than forging international agreements; it has meant enforcing them. The burdens of leadership are often heavy, but the world is a better place because we have borne them. And so, to my friends on the right, I ask you to reconcile your commitment to America's military might with a failure to act when a cause is so plainly just; to my friends on the left, I ask you to reconcile your belief in freedom and dignity for all people with those images of children writhing in pain and going still on a cold hospital floor---for sometimes resolutions and statements of condemnation are simply not enough. Indeed, I'd ask every member of Congress and those of you watching at home tonight to view those videos of the attack and then ask: What kind of world will we live in if the United States of America sees a dictator brazenly violate international law with poison gas and we choose to look the other way? \#

\textbf{AMY GOODMAN}: That was President Obama addressing the nation last night. Professor Noam Chomsky, your response to his description of those who oppose military strike against Syria for a chemical weapons attack? \#

\textbf{NOAM CHOMSKY}: Well, once again, what's particularly interesting is what he didn't say. So, yes, a good idea to look at the videos of the gas attack in Syria. But then we could also look at the photos of deformed fetuses in Saigon hospitals still appearing decades after John F. Kennedy launched a major chemical warfare attack against South Vietnam, 1961, dousing the country with poisonous dioxin-laced Agent Orange. Dioxin is one of the major carcinogens. The attack was aimed at food crops, in an effort---and at ground cover, part of a general assault against the country---a huge number of atrocities, millions of people killed. The chemical---the effects of chemical warfare are felt until today, partially by American soldiers, too. Or we could look at the photos of other deformed fetuses coming regularly in Fallujah, attacked by U.S. Marines in November 2004, killing several thousand people, destroying much of the town, using weapons which---of unknown character, but which left radiation levels that epidemiologists have estimated are comparable to Hiroshima. And the effects of that on high cancer rates, on deformed fetuses, on children devastated by horrifying deformities, that we could look at, too. Now, those are the ways in which the U.S. has brought---has been the anchor for global security for seven decades. Can run through the record, if there were time, but everyone should know it. These, of course---that's not said.

The U.S.---the idea that the U.S. has introduced and imposed principles of international law, that's hardly even a joke. The United States has even gone so far as to veto Security Council resolutions calling on all states to observe international law. That was in the 1980s under Reagan. No state was mentioned, but it was evident that the intention was to request the United States to observe international law, after it had rejected a World Court judgment condemning it for what was called unlawful use of force---it means international terrorism---against Nicaragua. In fact, the U.S. has been a rogue state, the leading rogue state, radically violating international law, refusing to accept international conventions. There's hardly any international conventions that the U.S. has accepted, and those few that it has accepted are conditioned so as to be inapplicable to the United States. That's true even of the genocide convention. The United States is self-authorized to commit genocide. In fact, that was accepted by the International Court of Justice. In the case of Yugoslavia v. NATO, one of the charges was genocide. The U.S. appealed to the court, saying that, by law, the United States is immune to the charge of genocide, self-immunized, and the court accepted that, so the case proceeded against the other NATO powers but not against the United States. In fact, the United States, when it joined the World Court---it helped introduce the modern World Court in 1946, and joined the World Court, but with a reservation. The reservation is that international agreements, laws, do not apply to the United States. So the U.N. Charter, the charter of the Organization of American States, the U.S. is immune to their---self-immunized to their requirements against the threat and use of force, intervention and so on.

It's kind of astonishing. I mean, by now it's hard to be astonished, but it should be astonishing that a president of the United States, who is furthermore a constitutional lawyer or a graduate of Harvard Law School, can say things like this, in the full knowledge that the facts are exactly the opposite, radically the opposite. And there are millions and millions of victims who can testify to that. Right today is---happens to be an important date, the 40th anniversary of the overthrow of the parliamentary democracy of Chile, with substantial U.S. aid, because we insisted on having a vicious dictatorship, which became a major international terror center with our support, rather than allowing a Democratic Socialist government. Well, that's---these are some of the realities of the world. Now, the picture that the president presented is---it doesn't even merit the name fairy tale. \#

\textbf{NERMEEN SHAIKH}: Well, Professor Noam Chomsky, why do you think that the U.S. so quickly started to push for military strikes? And what do you think the U.S. or the international community should do to deal with this alleged use of chemical weapons in Syria? What do you think the appropriate response would be? \#

\textbf{NOAM CHOMSKY}: The appropriate response would be to call for imposing the chemical weapons convention in the Middle East---in fact beyond, but we'll keep to the Middle East---which would mean that any country that is in violation of that convention, whether it has accepted it or not, would be compelled to eliminate its chemical weapons stores. Just maintaining those stores, producing chemical weapons, all of that's in violation of the convention, and now is a perfect opportunity to do that. Of course, that would require that U.S. ally Israel give up its chemical weapons and permit international inspections. Incidentally, this should extend to nuclear weapons, as well. The further step would be to move towards the kinds of negotiations, Geneva negotiations, that the U.N. negotiator, Lakhdar Brahimi, has been calling for, with Russian support and with the United States kind of dragging its feet. Obama misstated that, too, last night. That's the one thin hope, and it's pretty thin, for some way to allow Syria to escape what is in fact a plunged, virtual suicide. \#

\textbf{NERMEEN SHAIKH}: And why do you think the U.S. started to push for military action so swiftly? \#

\textbf{NOAM CHOMSKY}: As it always does. The United States is a violent military state. It's been involved in military action all over the place. It invaded South Vietnam, practically destroyed Indochina, invaded Iraq, elicited a Sunni-Shia conflict, which is now tearing the region to shreds. I don't have to run through the rest of the record. But the United States moves very quickly to military action, unilaterally. It can---sometimes can get some allies to go along. In this case, it can't even do that. And it's just a routine. The United States is self-immunized from international law, which bans the threat or use of force. And this is taken for granted here. So, for example, when President Obama repeatedly says all options are open with regard to Iran, that's a violation of fundamental international law. It says we are using the threat of force, in violation of international law, to which we are self-immunized. There's nothing new about this. Can you think of any other country that's used military force internationally on anything remotely like the scale of the United States during these seven decades when, according to Obama, we've been the anchor of global security? \#

\textbf{NERMEEN SHAIKH}: Well, Noam Chomsky, supporters of the U.S. plan say that the only reason that Assad agreed to hand over, relinquish control over chemical weapons was because of the threat of military force, of U.S. military force. And what interest does the U.S. have in striking Syria militarily? \#

\textbf{NOAM CHOMSKY}: The first comment is correct. The threat and use of force can be effective. So, for example, Russia was able to control Eastern Europe for 50 years with the threat and occasional use of force. Hitler was able to take over Czechoslovakia with the threat of force. Yes, it often works, no doubt. That's one of the reasons it's banned under international---under international law.

The reason---the pretexts for imposing---for carrying out a forceful act have generally declined, to the point that even the British government hasn't accepted them, and the Congress was apparently going to reject them, and the United States, the government, resorted to the---what is usually the last---the last resort, when everything else fails, saying our credibility is at stake. That's correct. U.S. credibility is at stake. Obama issued an edict, and it has to be enforced. That's a familiar doctrine. It's one of the leading doctrines of world affairs. Credibility of powerful, violent states must be maintained. It's---occasionally called it the Mafia doctrine. It's essentially the doctrine by which the godfather rules his domains within the Mafia system. That's one of the leading principles of world order: Credibility has to be maintained.

But that has many variants. Sometimes it's called the domino theory. If we don't impose our will here, the dominos will start to fall, others will begin to be disobedient. In the case of Chile 40 years ago, to go back to that, what Latin Americans called the first 9/11, Henry Kissinger explained that Chile, under Allende, he said, is a virus that might spread contagion elsewhere, all the way to southern Europe. And he wasn't saying that Chilean troops were going to land in Rome. He was concerned, rightly, that the model of peaceful, parliamentary democracy might spread, in which case the contagion would spread beyond, and the U.S. system of domination would erode.

Just earlier on the program, you had an interview with Saul Landau, the late Saul Landau, with regard to {[}Cuba{]}, and exactly the same doctrine applies there. The U.S. carried out---invaded Cuba, Bay of Pigs invasion. When that failed, Kennedy launched an enormous terrorist campaign, murderous terrorist campaign. The goal was to bring "the terrors of the earth" to Cuba, as Arthur Schlesinger described it, Kennedy's adviser, Latin American adviser. It was in the hands of Robert Kennedy, and it was no joke. It was very serious. Now, that's been followed by 50 years of economic warfare, very harsh economic warfare, all unilateral. The world was overwhelmingly opposed to it. But it doesn't matter: We, as a rogue state, we do what we like. And the reasons are explicit in the internal record. The reasons, you go back to the early '60s, the internal government record explains that Castro is guilty of what they called "successful defiance" of the U.S. principles going back to the Monroe Doctrine, 1823---no Russians, just the Monroe Doctrine, which established, in principle, our right to dominate the hemisphere. The U.S. wasn't powerful enough to do it then, but that was the principle, and Castro is carrying out "successful defiance" of that principle, therefore he must---Cuba must be subjected to massive terrorism, economic warfare and strangulation. That's been going on for 50 years. Same principle, the Mafia principle.

The same was true in Vietnam. The primary motive for the Indochina wars, going back to the early 1950s, was presented here as the domino theory. But what that meant was, if you read the internal records, that there was a fear, a justified fear, that successful independent development in Vietnam might spread through the region, might spread contagion through the region. Others would attempt the same path, that itself was of no great significance, but it might spread as far as Indonesia, which has rich resources, and there, too, there might be a move towards independent development, independent of U.S. domination. And it was even feared that that might bring in Japan. John Dower, the famous Asia historian, described Japan as the "superdomino." The U.S. was concerned, deeply concerned, that if Southeast Asia moved toward independent development, Japan would "accommodate," the word that was used, to East and Southeastern Asia, becoming its technological industrial center and creating a system, an Asian system, from which the U.S. would maybe not be excluded, but at least which it wouldn't control. Now, the U.S. had fought the Second World War to prevent that. That's Japan's new order, and it was in danger of being reconstituted if Indochina gained independence. That's the domino theory. And that was understood. McGeorge Bundy, Kennedy-Johnson national security adviser, in retrospect, observed that the Vietnam War---the United States should have called off the Vietnam War in 1965. Why 1965? Well, because in 1965 a U.S.-backed military coup took place in Indonesia, slaughtering hundreds of thousands of people, wiping out the only mass-based political party and instituting a regime of torture and terror, but opening the country up to Western exploitation, with its rich resources, and that meant that the Vietnam War was essentially over. The U.S. had won its main objectives. It was pointless to continue it.

Now, this policy is---these are major principles of world affairs, and they're understandable, and they're understood. So, go back to Cuba again. When Kennedy came into office, he was concerned with changing Latin American policy. He developed the---set up a Latin American research commission. It was headed by Arthur Schlesinger, his historian who was his adviser, and they came out with a report. It was presented by Schlesinger to the president. And in it, Schlesinger described the problem of Cuba. He said the problem of Cuba is the Castro idea of taking matters into your own hands, an idea which may have resonance in other parts of Latin America, where the mass of the population is subjected to the same kind of harsh repression that they are in Cuba. And if this idea spreads, the U.S. system of control erodes. Well, going back to the Middle East, it's the same. \#

\textbf{AMY GOODMAN}: Noam, we're going to go back to the Middle East just when we come back from break. We want to ask you about Syria in the larger Middle East context, particularly looking at Iran and looking at Israel. And, of course, as you point out, this is major date in history. Forty years ago today, September 11, 1973, in Chile, Salvador Allende died in the palace as the Pinochet forces rose to power. And it is also the 12th anniversary of the September 11th attacks. This is Democracy Now! We'll be back in a moment.

\end{document}

