% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Elizabeth Rush}
\title{ Down on the disappearing bayou  }
\date{2015, október 16 \thanks{LMD} }
\maketitle


\textbf{Rising sea levels, river dams, and the hydrocarbon industry are daily diminishing the once-rich wetlands that the Mississippi delta created along the Louisiana coast.}


In late August the evening sky in southern Louisiana is often a deep bruised blue. Tonight the horizon threatens thunder. A couple of mullet throw themselves out of the water, their belly-flopping bodies make noises when they land, doip! doip! doip! Jake Billiot noses~\emph{Sitting Bull,}~his shrimping boat, up against the sidewalk that wraps around the Pointe-aux-Chenes launchway. As the name suggests, Pointe-aux-Chenes is the end of the line --- the very southern tip of a long finger of land ever more precariously connected to North America's immense solidity by a single road. The closer you get to the end of highway 665 the more water you see, until, at the place where the road stops, the land drops away.

Past Billiot's boat, the land and water are only two colours --- the occasional green cordgrass and the near-black of the bayou. This expanse used to be a diverse estuarial marshland where shrimp spawned in spring and blackpoll warblers paused on their great transcontinental migration. But over the past 50 years, roughly 90\% of the Isle de Jean Charles, Pointe-aux-Chenes, and the surrounding area have disappeared. Compare aerial images of then and now, and the change is dramatic. Blue replaces green. What was once one of the largest and most productive coastal wetlands in the world is currently losing a football field of land every hour~(\href{http://mondediplo.com/2015/11/11louisiana\#nb1}{\emph{1}}).

The Louisiana bayou is disappearing --- because of rising sea levels, coastal erosion, land subsidence and an estimated 10,000 miles of canals dug by oil companies~(\href{http://mondediplo.com/2015/11/11louisiana\#nb2}{\emph{2}}) --- and with it goes a way of life. Billiot is 70 and was born and raised on the bayou: \textqq{I have been fishing on my own for 55 years. When I started, I caught crab and shrimp in the summer and there was land left that you could trap in the winter. There was always more than enough, but now the muskrat are all gone and I have to motor for hours sometimes just to catch a little something.}

Today we think of sea rise as linked to global warming, but the deltaic lands at the mouth of the Mississippi have been losing ground to the sea since the 1930s when the US Army Corps of Engineers dammed the river that created them. For much of the past 10,000 years, the mighty Mississippi drained a vast watershed from Wyoming to Pennsylvania, from the border of Canada to the Gulf of Mexico. The world's third-longest river built up the Louisiana coast by carrying silt and sediment from the far reaches of the continent and spitting it out where the river emptied into the sea.

Pre-Columbian Native American societies understood that a healthy river goes through cycles of flood and drought, and they shaped their civilisations around the Mississippi's ebb and flow. Their villages were not on the banks, but nearby, and most weren't permanent settlements but camps that could be relocated if the waters rose. In~1543 Spanish conquistador Hernando de Soto was stopped in his march westward across present-day Tennessee by a swollen Mississippi. His chronicler, Garcilaso de la Vega, wrote about it in his journals~(\href{http://mondediplo.com/2015/11/11louisiana\#nb3}{\emph{3}}). This was the first time the river's regular high waters and sediment-delivering surges were interpreted as a deterrent to human progress. The second recorded instance of the river's \textqq{wrath} was in 1734 when the Mississippi inundated a fledgling New Orleans for several months. It wasn't until the river got in the way of, or ruined, European prospects that its irregular flow was deemed a problem.

In 1927 the river inundated towns along the banks, flooding an area the size of Massachusetts, Connecticut, New Hampshire and Vermont for six months. In an effort to cut down the destruction that the regular flooding brought to the delta, the Army Corps of Engineers began to dam the river that had made the land they were trying to protect. But river controls had the reverse effect. \textqq{When the silt backed up behind those dams, the floods came stronger and didn't carry any sediment with them, and that's when the land started to go,} says Lora Ann Chaisson, vice-principal chief of the United Houma Nation. Suddenly the long tongues of alluvial land that defined the Mississippi delta started to withdraw. Chaisson doesn't live on the Isle de Jean Charles, the seat of the local Native American community, any more. Like many others she moved in a couple of miles: \textqq{When I came here, to Pointe-aux-Chenes, all I wanted to know was, does it flood? `No, it don't flood here,' they told me. And do you know what? You can't stop the water; it is coming in here now too. Of course it only got worse when they discovered oil on the baya.}

\textbf{Oil country}

In 1948 the first oilrig was installed about five miles from her childhood home. With the rigs came channelisation, where routes are dug through the marsh and swamp for easy access to the rigs. The oil companies were supposed to \textqq{rock} each channel --- to backfill them when the well came out, reducing the movement of water through the fragile marshland that surrounds and supports the bayous. \textqq{But they didn't do that, they didn't maintain the bayalike they said they would and now the Gulf is at our back door,} says Billiot.

From the air, the fine lace of Louisiana's wetlands is slashed with razor cuts --- channels for the rigs, dug by many hired hands (including Billiot's father) so that extracting oil might turn a greater profit. Each of these thousands of channels gives the salt water another way in. The more water moves through the wetlands, the wider the cuts become, and the faster the wetlands disappear~(\href{http://mondediplo.com/2015/11/11louisiana\#nb4}{\emph{4}}). As a result, 30~square miles of these marshes vanish annually. This might not seem a lot, but Louisiana contains about 50\% of US wetlands, and they are some of the best defences against ever-more-frequent offshore storms. Wetlands act like giant sponges, absorbing the water that often accompanies storm surges. Instead of the surge hitting the coast all at once, wetlands traps the water, then slowly release it, saving inland areas from flooding.

Out in the bayou turned channel behind Pointe-aux-Chenes, Billiot calls on the radio to find out if anyone else is catching. The results don't sound promising but he decides to try anyway. He is one of the wealthier shrimpers in the area because a couple of his boats were hired by BP to help clean up after the 2010 Deepwater Horizon oil spill. Unlike many of his peers he can afford to burn a little extra fuel.

During the 87 days it took to stop the oil spill, the Food and Drug Administration ordered all local fisheries closed, putting thousands out of work. That's what the newspapers said, yet it might be more accurate to say that the spill temporarily ended not just jobs but an entire way of living and getting by. Shrimping, the third largest local industry after oil and natural gas, is less like work and more like a religion. Fishers link shrimp numbers to fate and God's good favour.

\textbf{`Just break the eggs'}

Limited shrimp have returned since the spill, some of them with small tumours and others born without eyes~(\href{http://mondediplo.com/2015/11/11louisiana\#nb5}{\emph{5}}). The explosion of the rig released 210m US gallons of oil into the Gulf of Mexico, and BP sprayed 1.9m~gallons of the dispersant Corexit, designed to break apart the oil and suspend it underwater, getting rid of surface slicks. Corexit is heavy and sinks to the ocean floor, pulling oil with it and contaminating surfaces where shellfish spawn. Fish in their egg and larval phases have none of the defences of more mature animals, making them particularly susceptible to toxins like those in Corexit. Even small doses can cause death or mutation~(\href{http://mondediplo.com/2015/11/11louisiana\#nb6}{\emph{6}}). \textqq{You don't have to kill all the chickens, just break all the eggs,} says Burt Knight, a lifelong shrimper. \textqq{I am catching only a tenth of what I did years ago.}

Billiot chugs silently up the tea-coloured channel behind Pointe-aux-Chenes. The incoming tide is running strong, a good thing since the faster the water moves the less likely the netted shrimp are to swim out. Billiot turns a switch and the winches start to unspool metal cables. Suddenly his little boat looks like a cormorant unfurling its wings. \textqq{When I was a boy, it was like the trees shook hands across this little baya,} Billiot says. It is hard to imagine. The shore is a good hundred feet away in both directions and the only trees are dead. We pass a huge cypress stripped bare; the cause of its untimely demise is deep in the ground, where, sometime during the last half century, its roots started to suck up salt water instead of fresh.

Only storytelling can take you back to Pointe-aux-Chenes as it was. Where there is water today there once was a cypress grove, a long wetland and plenty of waterfowl to shoot, a general store, a home. Everyone has a storehouse of memories from what seems like an entirely different place. Billiot draws a map on an envelope, using little waves to cover over where wetland and islands used to be. \textqq{Whiskey Pass, Cat Island Pass, Wine Island Pass, those little channels to the Gulf grew bigger, and the salt water started to push in.}

\textbf{The problem with magical thinking}

The US is eleventh on the list of countries most at risk from sea level rise, just behind the Netherlands, Bangladesh and the Philippines, so we are going to have to figure out what to do with our densely populated coasts, and soon. Currently 18~million live in areas at high risk of a 100-year-flood --- roughly 6\% of the population of the US --- and this is expected to double by the century's end~(\href{http://mondediplo.com/2015/11/11louisiana\#nb7}{\emph{7}}). The Intergovernmental Panel on Climate Change (IPCC) predicts that the Greenland ice sheet's ever-accelerating melt (during the summer of 2014 it lost, on average, 40,000~square miles of ice each day)~(\href{http://mondediplo.com/2015/11/11louisiana\#nb8}{\emph{8}}) coupled with warmer sea temperatures, is likely to heighten global sea levels between 11 and 38~inches by 2100. By then, many low-lying communities will be underwater --- including most of Louisiana's bayou country --- unless we can figure out a way to protect them.

Nine out of 10 of the costliest hurricanes have impacted the Gulf coast, and six of those hit in the last decade~(\href{http://mondediplo.com/2015/11/11louisiana\#nb9}{\emph{9}}). In 2012 Louisiana, reeling from the increased cost of coastal rebuilding, unanimously approved a 50-year coastal protection plan for reducing flood risk and land loss. The Master Plan is one of the first integrated sea rise adaptation plans, using science to investigate multiple adaptation strategies to improve decision-making; its creation brought together stakeholders from a wide range of communities, including the oil and natural gas industry, policy makers and indigenous groups.

The Master Plan will cost \$50bn but, according to Jordan Fischbach of the RAND Water and Climate Resilience Centre, the price of doing nothing would be significantly higher. \textqq{The implementation of the Master Plan is projected to dramatically decrease estimated annual damages (EAD) from coast-wide flooding from \$2.2-2.8bn today to \$4.8bn in 2050. {[}...{]} Without the Master Plan in place, EAD could exceed \$20bn under the less optimistic sea rise scenario.}

Creating resilience is a moving target, and funds are difficult to come by. In Louisiana a significant portion of the settlement from the Deepwater Horizon spill is slated to go towards wetlands repair and coastal buffering --- the comprehensive infrastructure initiatives that the Master Plan calls for. On the list are an even more extensive levee system, structural elevation, and largely untested restoration projects, such as large-scale sediment diversion schemes that would pump and distribute dammed river silt throughout the bayou. These sediment-delivering pipes would crisscross the delta, perhaps even taking advantage of the oil industry's distribution channels.

But the plan does not attempt to significantly reduce the oil and gas industries' footprint. While the Interior Department has proposed tougher standards for offshore oil development, lawmakers have yet to approve, and without such measures, spending billions on sediment redistribution puts good work behind bad. The Master Plan embodies the collaborative approach to combating climate change that a planetary crisis demands, but any serious discussion of coastal resilience must also look at mitigation, by reducing the stressors.

\textbf{Low-lying and left out~}

Albert Naquin, chief of the Isle de Jean Charles (two miles south of Pointe-aux-Chenes), has been trying to get residents to agree to leave en masse, the only way he can secure relocation funds from the Army Corps of Engineers. His village was left out of the state's previous large-scale infrastructure scheme (the Morganza to the Gulf Protection Plan), and he has little reason to believe that the new Master Plan will include them: \textqq{We were chased out here by the whites} --- the area has a long history of hosting persecuted Native Americans --- \textqq{and now Mother Nature is chasing us back. But not everyone wants to leave.}

Edison Dadar, one of the few remaining residents of the isle, says: \textqq{My dad lived 91 years on the island, and his dad was born and raised and died here too. We have been down here a long time.} He pulls a five-gallon bucket out of his refrigerator. There can't be more than 50~shrimp in its bottom, two big whites and the rest brown, while it isn't much, it is enough to feed him, his wife and son, who still lives at home.

He shows off his garden, which has a bathtub and a few raised beds to keep the saline out of the roots. There are three cantaloupe melons and a couple of cucumbers. Out front, the two persimmon trees Dadar planted after Hurricane Andrew catch the wind: \textqq{Each tree you have is good protection. Plus the fruit is pretty tasty too.}

Perhaps the solution is not a multi-billion-dollar plan, but to live like Dadar --- to make do with less. Quit building bigger levees and learn to live smaller, slighter than before. According to Naomi Klein, whose recent book chronicles the ways in which deregulated capitalism has rapidly warmed the planet, we need not move back into the Stone Age in order to live within our planetary means: \textqq{The truth is that if we want to live within ecological limits, we would need to return to a lifestyle similar to what we had in the 1970s, before consumption rates went crazy}~(\href{http://mondediplo.com/2015/11/11louisiana\#nb10}{\emph{10}}).

Jake Billiot and I trawl back and forth across the bayou for hours. We catch next to nothing and burn a couple of gallons of gasoline. \textqq{Maybe, if we are lucky, tomorrow will be better,} says Billiot. He looks up at me, and the thin wire of his smile fades. We both know it has nothing to do with luck.

\textsc{Russia returns to the Middle East}

\textbf{Putin's Syrian bet}

\textbf{Putin has taken an enormous risk in backing Syria's Assad with military force. If his gamble pays off, Russia could be the peacemaker with only marginal military losses. If not, he invites destabilisation and threats from ISIS back home.}

\textsc{by Alexei Malashenko}

Russian Sukhoi Su-34 fighter-bombers and Kalbir cruise missiles fired from the Caspian Sea have~changed the balance of power on the Syrian battlefield, at least in the short term. The intensity of the bombing has allowed the Syrian government forces to take the offensive again. It's unsurprising that Putin is giving Bashar al-Assad this tactical support: Syria is the last vestige of Russia's Middle East presence, a symbol of past glory. Russia's steadfast support of the Assad regime led it to play a decisive role, in~2013, in dismantling Syria's chemical arsenal, to avoid western intervention~(\href{http://mondediplo.com/2015/11/02russiandiplo\#nb1}{\emph{1}}). It has proved wrong those who claimed Russia was now only a regional power, with no interests outside the post-Soviet space.

From the first armaments contracts in 1956, Syria maintained very close relations with the Soviet Union, strengthened by Syria's short-lived political union with Egypt (United Arab Republic, 1958-61) and the rise to power in 1963 of the Baath Party, inspired by Arab socialism. President Hafez al-Assad, before his death in 2000, urged his son Bashar to maintain the relationship as vital to keeping his clan at the head of the country.

After the breakdown of Russia's alliance with Egypt and the loss of support facilities at Alexandria and Mersa Matruh in 1977, the Syrian port of Tartus was the only base for Russian warships in the Mediterranean. Over the past few months, these have become a noticeable presence off the Syrian coast; in September the huge Typhoon-class nuclear submarine~\emph{Dmitry Donskoy}~was in the area.

Russian aid to Syria has grown since the Arab Spring. The collapse of the Tunisian, Egyptian and Libyan regimes, the dismantling of Iraq and the emergence of ISIS in 2014 convinced Russia that it must continue to support Assad and strengthen its position in the region. Widespread instability and difficulty in reading the policies of the West, especially those of the US, have driven governments to diversify their alliances. France has sold arms to the Gulf states; Russia has recently signed trade, military and technical agreements with Egypt, Iraq and Jordan; Saudi Arabia finances Egyptian purchases of Russian arms, and the Saudi sovereign wealth fund decided in July to invest \$10bn in Russia.

\textbf{Nostalgia for Nasser}

Some Arab politicians and military officers are nostalgic for the days of Gamal Abdel Nasser, in the 1950s and 60s, when ideological competition between the Soviet Union and the West gave Arabs room for manoeuvre. It was not by chance that Egypt's president Abdel Fatah al-Sissi praised Nasser during his electoral campaign in 2014. Putin has renewed old ties and secured an arms contract worth \$3.5bn by supporting Sissi --- who visited Moscow in August --- without hesitation.

Russia hopes to strengthen its influence under cover of international law. In a statement at the UN General Assembly on 28~September, Putin proposed the idea of \textqq{a resolution aimed at coordinating the actions of all the forces that confront the Islamic State and other terrorist organisations. {[}...{]} Naturally, any assistance to sovereign states can and must be offered rather than imposed, exclusively and solely in accordance with the UN Charter. {[}...{]} I believe it is of the utmost importance to help restore government institutions in Libya, support the new government of Iraq, and provide comprehensive assistance to the legitimate government of Syria}~(\href{http://mondediplo.com/2015/11/02russiandiplo\#nb2}{\emph{2}}).

Russia is returning to the Middle East, though without claiming that it will regain the influence the Soviet Union once had on international relations. But it also faces a paradox: Assad is far from popular in many Arab countries, and this alliance aligns Russia with Iran, Hizbullah in Lebanon and the Shia militias in Iraq, in a regional confrontation that their Sunni adversaries increasingly present as sectarian. If Putin wants to demonstrate Russia's strength and ability to help its friends --- both to public opinion at home and to his regional partners --- he can't afford to show weakness by giving up Assad.

Can the conflict be resolved by an accord? From the Russian viewpoint, this would be possible if the West accepted that Assad should remain in power, at least for a defined period. Putin told the Collective Security Treaty Organisation (CSTO, the Russian NATO)~(\href{http://mondediplo.com/2015/11/02russiandiplo\#nb3}{\emph{3}}) summit in Dushanbe, Tajikistan, in September that it was vital to consider political compromise in Syria, saying that Assad was ready to involve the \textqq{healthy} part of the opposition in government. This would make it possible to form a Syrian coalition including that part of the opposition able to break with the jihadist movements. The president could then hand over \textqq{of his own free will} to a successor who would be approved both by the major political forces in Syria and by external players.

This may seem unlikely at present, but has already been discussed by foreign ministers --- albeit reluctantly. If it comes about, Russia could look like an agent for peace. Intervening at the request of the Syrian government would give it the status of geopolitical counterweight, like the Soviet Union during the cold war, and protector of minorities in the region --- a role that tsarist Russia assumed with regard to Eastern Christians.

\textbf{Trading Syria for the Donbass?}

But Russia is playing on a far bigger chessboard. There are indications that it may be planning to \textqq{exchange} Syria for the Donbass region of Ukraine, torn between those who want to rejoin Russia and those faithful to Kiev. If the US and its allies were willing to take greater account of Russia's interests in Syria, Russia could be more understanding about Ukraine. In September, Russia and Ukraine, with the European Union as mediator, agreed a price for gas to suit both Ukraine and Russia's Gazprom, which has been finding it difficult to secure new contracts~(\href{http://mondediplo.com/2015/11/02russiandiplo\#nb4}{\emph{4}}). Though it is uncertain that the Minsk I and II agreements of 2014 and 2015 will be applied in full, a meeting in October raised hopes of a lasting truce in the Donbass, with the effective withdrawal of heavy weaponry and both sides accepting that local elections should be postponed to allow an institutional solution.

Russia is at an impasse. At the UN General Assembly, Putin accused some members of NATO of having orchestrated \textqq{a military coup {[}...{]} from outside that triggered a civil war.} But he defended the Minsk agreements:\textqq{Ukraine's territorial integrity cannot be ensured by threats and force of arms. What is needed is a genuine consideration for the interests and rights of the people in the Donbass region, and respect for their choice.}

In Syria, the post-Assad future, and whether a coalition may be possible, is uncertain. Every player in the conflict sees a different solution. Russia continues to increase military and technical aid. The airstrikes by Russian planes are a major logistical exercise: since the enemy is quite near, the airbase at Latakia needs to be protected by Mil Mi-24 assault helicopters and tanks. Official sources talk of 2,000~men, which the military apparently consider sufficient. But the Crimean crisis has already shown how hard it is to determine the real numbers of troops involved. There is however nothing new about the presence of military advisors: Russia has been sending them to the Middle East since the~1950s.

The role of Russian troops may be limited to protecting the Syrian army's principal bases, and special operations. Russians remember the \textqq{limited contingent} initially sent to Afghanistan and the stalemate (1979-89) during which, according to official figures, 14,000 Soviet troops died and 50,000 were wounded, while 1.5 million Afghans were killed. ISIS ideologues have pointed out that this disastrous war contributed to the fall of the Soviet Union. A ground intervention in Syria would do little for Putin's popularity.

\textbf{Keeping Assad in power}

According to western leaders, Russia's main objective is not to crush ISIS but to keep Assad in power. The airstrikes have targeted different opposition groups, including the Al-Nusra Front, affiliated to Al-Qaida. Russia hopes that confronting ISIS will encourage other countries involved to join it in a \textqq{war on extremism}. However, a worldwide coalition is unlikely; at best, there will be some technical coordination to prevent air accidents. Also unlikely is a coalition of Russia, Iran and China, as discussed in the Russian media, to rival the one led by the US. China refuses to intervene outside its own borders, and Iran is pursuing its own ends --- although the powerful General Qassem Suleimani, commander of the Revolutionary Guards' Al-Quds force which is fighting in Iraq and Syria, has been seen in Moscow. Israel's prime minister Binyamin Netanyahu, who is worried about such rapprochements and fears that an alliance between a Shia axis and Russia will allow Hizbullah to obtain arms, has sought assurances from Putin.

ISIS has served Russia's interests by allowing Russia to show its friends in the region that it can still take a decisive role. Russia can present itself as protector of the Muslim countries of Central Asia, through the CSTO. More than ever since the breakup of the Soviet Union, Russia talks of maintaining stability in \textqq{Eurasia} and protecting its allies from external threats. For many years, this meant the Afghan Taliban; now it includes ISIS, which is gradually penetrating southern Central Asia.

ISIS identified Russia as an opponent from the start. Last year it released a video warning Putin: \textqq{These are the planes {[}captured by ISIS{]} that you sent to Bashar {[}al-Assad{]}. Allah willing, we will send them back to you. {[}...{]} Your throne has already been shaken {[}...{]} and will fall with our arrival {[}in Russia{]}.} ISIS also vowed to free Chechnya and all of the Caucasus. This may seem ambitious, but the jihadists are in a position to strengthen their influence in the North Caucasus, still beset by protests against central authorities, as well as economic and social problems. There will probably be more attacks like those on the Russian embassy in Damascus (in May, September and October), and they could spread to Russian soil.

The decision to invade Afghanistan in 1979 was made at a meeting of the Politburo, within the Central Committee of the Communist Party. Leonid Brezhnev allowed his colleagues to convince him. The head of the KGB, Yuri Andropov, future general secretary of the Central Committee, initially opposed the invasion, but later came round.

Today, we know who makes the decisions, and how: though the senate unanimously approved sending Russian troops to Syria in September, Putin makes all the decisions himself, based on his own understanding of the situation. Some decisions seem to have a strong emotional element, and to be made without sufficient thought as to consequences.

Russians remember the Cuban missile crisis of 1962: Soviet leader Nikita Khrushchev had suddenly decided to send nuclear missiles to Cuba. He claimed to have discussed this with his colleagues, but we now know that he took the decision alone, and that his entourage was only informed of it~(\href{http://mondediplo.com/2015/11/02russiandiplo\#nb5}{\emph{5}}). The US's sharp reaction, blockading Cuba and threatening to invade it~(\href{http://mondediplo.com/2015/11/02russiandiplo\#nb6}{\emph{6}}), forced Russia to back down, and the incident seriously damaged Khrushchev's credibility with his colleagues.

Many see Putin as more decisive and effective than President Obama. They claim he has demonstrated this in his management of the Syrian crisis, and is one step ahead of his partners/rivals. But providential military victories are often followed by a stalemate, or even a retreat. All the bombing over the past year has not forced ISIS to retreat. Russia's attempt to return to the Middle East will only succeed if it is able to create conditions favourable to an international political solution.

\end{document}

