% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Philip S Golub}
\title{  China rewrites the global rules }
\date{2016, január 13 \thanks{LMD } }
\maketitle

\begin{abstract}
  \textsc{`A historic transition that will unfold over decades'}

\textbf{China rewrites the global rules}

\textbf{Despite worries over the end of the Chinese economic miracle and harder times because of the aftershocks of the 2008 financial crisis, global rebalancing is happening --- and quickly.}

\textsc{by Philip S Golub}

\end{abstract}


The International Monetary Fund's executive board announced on~30~November that the Chinese currency, the renminbi (RMB) or yuan, would soon be included in the basket of currencies that make up its Special Drawing Rights (SDR), beside the US dollar, the euro, the British pound and the yen.

The decision recognised the growing role of the Chinese currency in international trade settlements. Since the launch in~2009 of a pilot programme for cross-border RMB trade settlements in Asia, the share of Chinese trade settled in RMB has risen from~3.2\% to~25\%, and should, according to the~\emph{Financial Times,}~soon reach~46\%~(\href{https://mondediplo.com/2016/02/08china\#nb1}{1}). Offshore centres clearing the currency have been set up in Hong Kong, Singapore, Taiwan, London and Frankfurt. The RMB is still far from world reserve status, requiring full convertibility and complete capital market liberalisation, which China is not ready for and carries significant risks. But the RMB's rise is a \textqq{momentous event in {[}...{]} international finance}, according to a former IMF economist~(\href{https://mondediplo.com/2016/02/08china\#nb2}{2}) --- the beginning of a slow shift towards a tripolar world monetary system no longer exclusively centred in the West.

\textqq{Great powers have great currencies,} says Robert Mundell~(\href{https://mondediplo.com/2016/02/08china\#nb3}{3}). Like the gradual rise of the dollar to reserve status in the early 20th~century, the internationalisation of the RMB is part of the larger story of China's re-emergence as a semi-autonomous core of the world economy, and as the primary driver of the movement of East-West and North-South rebalancing that is a key feature of current world politics. China's share of world GDP, in purchasing power parity (PPP), has risen in 40~years from less than~2\% to over~16\%, ahead of the US. Though Chinese economic growth declined with the \textqq{great recession} after the~2008 crisis, average per capita GDP (PPP) is now expected to increase within a few years (from~\$9,800 to~\$16,000, against \$250 in~1980). Longer-range projections, which showed a growth trajectory similar to Japan in the late 20th~century, suggest that per capita GDP will reach current Japanese or European levels by the mid-21st century. China is gradually regaining a position in the world economy in keeping with its demographic weight.

\section{Key South-South actor}

It has become the core of South-South trade and investment linkage, supplanted Japan and the US as the leading trade partner of nearly all East Asian countries, and become a crucial actor in South American, African and South Asian trade. China is now Brazil's leading trade partner, ahead of the US. There are similar trends in Chile, Argentina and smaller South American economies. South Africa's export share to China has risen from 1.8\% in 1998 to over 12\%, while imports have risen from 3\% to 15\%. Indian exports to China have risen from 2.9\% to over 10\%, and imports from 2\% to 12\%.

This points to a re-patterning of the geography of world trade, and an East Asia-centric South-South trading system, with China at the core. Thanks to new regional and transcontinental linkages centred in East Asia, South-South trade (SST) has expanded more rapidly over 20 years than global trade. In 2013 SST was 25\% of world trade, 21\% of manufacturing exports, and 25\% of exports of manufactures with medium and high technological intensity. Trade between developed countries fell from 46\% to less than 30\%. The result is a reduction of traditional single-market dependency in Latin American, sub-Saharan and Southeast Asian countries on Europe and the US. This has been accompanied, particularly in East Asia, by industrial upgrading: the \textqq{developing} world's share of manufacturing value-added has risen from 8.8\% to nearly 30\%, and of merchandise trade from 25\% to 47\%.

China's re-emergence has begun to alter the vertical North-South relations of the age of western empire and industry. This generates new dependencies: when China sneezes, emerging country partners catch colds --- as with the current downturns for exporters of primary goods because of China's slowing growth. The impact of the global crisis on China evinces the need for a sustainable development strategy focused on domestic improvements rather than export growth. Even so, we are witnessing a structural change in the global economic and political hierarchy.

\section{Bye-bye Bretton Woods?}

Politics has lagged behind economics, but a gradual reordering of world politics is visible in the growing voice of the South in organisations and clubs (G20), its activism in world politics, and new international institution-building efforts bypassing traditional centres of authority. Frustrated by the slow pace of change, notably in the IMF and World Bank, China has been leading the effort of global South countries to create a new system to govern the world economy. In~2013 Brazil, Russia, India, China and South Africa (the BRICS) set up the New Development Bank (NDB),headquartered in Shanghai, with a capital base (money paid in and pledged) of~\$100bn; this will combine investment and monetary functions, serving as a lending institution for infrastructure development projects as well as a reserve facility for balance of payments issues. In~2014 China founded the Asian Infrastructure Investment Bank (AIIB), which, like the NDB, has a \$100bn capital base (the Japanese-led Asian Development Bank has \$160bn, the World Bank \$220bn).

The importance of these initiatives should not be underestimated. They significantly challenge the authority of traditional rule-makers, whose ability to set the international agenda and shape the global environment is waning. The erosion of Euro-Atlantic authority is clear from the irrelevance of old clubs (G7) and the difficulty of previously dominant countries and blocs (NATO) in setting policy agendas reflecting their preferences. It is also clear from the steady decline of the legitimacy and authority of the Bretton Woods institutions, notably the IMF, whose grip on the middle-income countries of the South has loosened since the late~1990s.

Recognising the challenge, the US intensively lobbied allied states in Asia and Europe to stay out of the AIIB, arguing that it would not meet IMF and World Bank standards of transparency, environmental and social responsibility, and democratic governance. The argument would have been more convincing had the IMF not been the arm of coercive Euro-Atlantic discipline for the South. With the exception of Japan, the US proved unable to sway its closest partners. The~\emph{New York Times}noted: \textqq{The surge to join the Bank was considered a major victory for China in a rare public showdown with the United States}~(\href{https://mondediplo.com/2016/02/08china\#nb4}{4}).

The decision to found the NDB and AIIB is the outcome of a movement building since the~1990s in East Asia and Latin America in reaction to IMF mismanagement of regional financial crises. The 1997-8 Asian crisis convinced many East Asian policy makers that it was time to take the future into their own hands and seek greater autonomy. The creation of the new system has huge implications: the ability to set policy frameworks and maintain international regimes through multilateral institutions is an essential dimension of power in world politics. Former Treasury Secretary Lawrence Summers wrote that October~2014 (when the AIIB was formed) \textqq{may be remembered as the moment the United States lost its role as the underwriter of the global economic system. {[}...{]} I can think of no event since Bretton Woods comparable to the combination of China's efforts to establish a major new institution and the failure of the US to persuade dozens of its traditional allies {[}...{]} to stay out of it}~(\href{https://mondediplo.com/2016/02/08china\#nb5}{5}).

\section{Rewriting the rules}

While it is implausible that the East (China) will supplant the West (the US) at the centre and apex of a new hierarchy in the foreseeable future, there is no question that power will become more diffuse. For those sensitive to the need to redress historic injustices, greater equality is desirable. Yet it also implies increased interstate competition for resources, capital, technology, status and political voice. Global rebalancing, still in an early phase, carries significant risks.

China and the US are deeply interdependent. China is the US's main creditor and America remains a crucial market for products made or assembled in China. The long-term interest of China is to gradually reduce its massive dollar holdings that expose it to US discretionary monetary decisions. The US has a similar interest in reducing its exposure to potential sovereign decisions that would expose the dollar --- and the US position in the world --- to significant risk.

The Obama administration is keenly aware that China's re-emergence is part of a \textqq{historic transition {[}...{]} that will unfold over decades}, with major consequences for the US in the 21st century. Without overtly seeking to contain China, and aiming to avoid confrontation, it wants to \textqq{manage competition from a position of strength}~(\href{https://mondediplo.com/2016/02/08china\#nb6}{6}). As Obama put it in his~2015 State of the Union address: \textqq{China wants to write the rules for the world's fastest-growing region. {[}...{]} Why would we let that happen? We should write those rules.}

Yet China, with other postcolonial states, is now in a position to co-author the rules. Power will have to be shared. The internationalisation of the RMB is just part of this story. While the predominance of the dollar will not be seriously challenged soon, a slow shift in global monetary relations seems inevitable. As Benjamin Cohen notes, the \textqq{most probable outcome is {[}like{]} the interregnum between the two world wars, when Britain's pound sterling was in decline and the dollar on the rise but neither was dominant. Coming years {[}will see{]} several monies in contention and none as clearly in the lead as in the recent past. The economic and political impacts of a more fragmented currency system could be considerable}~(\href{https://mondediplo.com/2016/02/08china\#nb7}{7}).



\textbf{More from~\href{https://mondediplo.com/_Philip-S-Golub_}{Philip S Golub}}

Philip S Golub is professor of international relations at the American University of Paris. His latest book is~\emph{East Asia's Reemergence}~(Polity Press, forthcoming March 2016).

(\href{https://mondediplo.com/2016/02/08china\#nh1}{1}) \textqq{The Future of the Renminbi} report,~\emph{Financial Times,}~London, 30~November 2015.

(\href{https://mondediplo.com/2016/02/08china\#nh2}{2}) Eswar Prasad quoted by James Kynge in \textqq{Pivotal moment for the renminbi and China} in \textqq{The Future of the Renminbi} report,~\emph{Financial Times,}~London, 30~November 2015.

(\href{https://mondediplo.com/2016/02/08china\#nh3}{3}) Robert Mundell quoted by Benjamin J Cohen in \textqq{Currency and State Power} in Martha Finnemore and Judith Goldstein (eds),~\emph{Back to Basics: State Power in a Contemporary World,}~Princeton University Press,~\\
Princeton, 2013.

(\href{https://mondediplo.com/2016/02/08china\#nh4}{4}) Jane Perlez, \textqq{\href{http://www.nytimes.com/2015/04/03/world/asia/china-asian-infrastructure-investment-bank.html?_r=0}{Stampede to Join China's Development Bank Stuns Even its Founder}},~\emph{The New York Times,}~2~April 2015; and \textqq{\href{http://www.nytimes.com/2015/12/05/business/international/china-creates-an-asian-bank-as-the-us-stands-aloof.html}{China Creates a World Bank of its Own, and the US Balks}},~\emph{The New York Times,}~4~December 2015.

(\href{https://mondediplo.com/2016/02/08china\#nh5}{5}) Lawrence Summers, \textqq{Time US leadership woke up to new economic era},~\emph{Financial Times,}~London, 6~April 2015.

(\href{https://mondediplo.com/2016/02/08china\#nh6}{6}) US National Security Strategy 2015.

(\href{https://mondediplo.com/2016/02/08china\#nh7}{7}) Benjamin J Cohen, \textqq{The Future of Reserve Currencies},~\emph{Finance \& Development,}~vol 46, no 3, International Monetary Fund, Washington DC, September 2009.

\end{document}

