% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Alain Gresh}
\title{ The new Arab awakening  }
\date{ 2011, március 18 \thanks{LMD } }
\maketitle

\begin{abstract}
  Neither with the West, nor against it

  The new Arab awakening - by Alain Gresh
\end{abstract}


The upheavals taking place across the Arab world have implications not just for the region but the world. As the United Nations attempted to calm the situation in Libya, the US told Gadafy it was time to go. While the EU fears mass immigration from Libya, the US faces the impact on the regional order of the fall of Mubarak, pillar of US policy across the region, from the Israeli-Palestinian conflict to Iran

A large Muslim country is overwhelmed by strikes and demonstrations. This pillar of US regional policy is damaged by authoritarian rule and its resources are looted by the president's family; there is social and economic crisis; Washington abandons an old ally and the US Secretary of State calls on a dictator to stand down and allow for democratic transition.

This may sound like Egypt in 2011. In fact, it was Indonesia in May 1998, and the call for President Suharto to stand down came from Madeleine Albright, not Hillary Clinton. He had seized power in 1965 with the help of the CIA in a coup in which half a million communists, or suspected communists, were killed. After the fall of the Berlin Wall in 1989 and the collapse of the Soviet Union in 1991, Indonesia was no longer needed as a bulwark against communism; the US decided it would rather support democratic movements, and direct them to suit its interests. President Bill Clinton wanted to project a more open image of the US. It turned out to be a wise choice, and Indonesia has maintained close relations with the US, even though, as an active member of the Organisation of the Islamic Conference, it has taken an independent stance on the Iranian nuclear issue.

What do we learn from this? No dictatorship lasts forever, even when it rules the world's most populous Muslim nation. Internal changes influence foreign policy, but the extent of evolution depends on the context: Egypt is not Indonesia, and the Middle East is not Southeast Asia.

It has been commonplace for western politicians and diplomats to sneer at the \textqq{Arab street}; they asked if we really needed to listen to hundreds of millions of people with their Islamist and anti-western slogans when we got on so well with their leaders, who were so good at maintaining order, and extended such warm hospitality. (Between 1995 and 2001, 400 French government ministers spent their holidays in Morocco.) These leaders maintained the fiction of the Israel-Palestine peace process, even as Israeli settlements spread.

The fantasy that the Arabs are passive and unsuited to democracy has evaporated in weeks. Arabs have overthrown hated authoritarian regimes in Tunisia and Egypt. In Libya, they have fought a sclerotic regime in power for 42 years that has refused to listen to their demands, facing extraordinary violence, hundreds of deaths, untold injuries, mass exodus and generalised chaos. In Algeria, Morocco, Bahrain, Yemen, Jordan, Iraqi Kurdistan, the West Bank and Oman, Arabs have taken to the streets in vast numbers. This defiance has spread even to non-Arab Iran.

And where promises of reform have been made but were then found wanting, people have simply returned to the streets. In Egypt, protesters have demanded faster and further-reaching reform. In Tunisia, renewed demonstrations on 25-27~February led to five deaths but won a change of prime minister (Mohamed Ghannouchi stepped down in favour of Beji Caid-Essebsi). In Iraq, renewed protests led to a promise to sack unsatisfactory ministers. In Algeria, the 19-year emergency law was repealed amid continuing protests. The demands are growing throughout the region, and will not be silenced.

The revolutions in Tunisia and Egypt, the uprising in Libya, and all the other popular movements that have shaken the region are not just about how people want to live and develop, but about regional politics. For the first time since the 1970s, geopolitics cannot be analysed without taking into account, at least in part, the aspirations of people who have retaken control of their destinies.

This is certainly the case with Egypt. Even if it is too early to predict foreign policy, Washington has lost an unconditional ally: US regional strategy has relied on Egypt, along with Israel (with which Sadat signed a peace treaty in 1979), for the last 30 years. Egypt took part in the 1990-91 Gulf war against Iraq, and Mubarak was at the forefront of the fight against the \textqq{Iranian threat}. He maintained the illusion of the Middle East \textqq{peace process}, putting pressure on the Palestinian Authority to continue negotiations, and regularly welcomed Israeli leaders to Sharm el-Sheikh, even though it was clear they had no intention of agreeing a peace accord. Egypt under Mubarak participated in the economic blockade of Gaza and helped scupper all attempts at reconciliation by Hamas and Fatah, even one negotiated by another \textqq{moderate} country, Saudi Arabia (the Mecca accord of May 2007). During the uprising, some demonstrators waved placards in Hebrew, claiming the only language Mubarak understood was that of Israel's leaders.

\section{Peace and stability}\label{peace-and-stability}

The Supreme Council of the Armed Forces, currently in charge in Egypt, has reassured Washington and Tel Aviv that it will respect Egypt's international commitments, a reference to the 1978 Camp David accords and 1979 Egypt-Israel peace treaty. It is unlikely Egyptians would want to return to a state of war, but they do not see these agreements as the basis of regional peace and stability: quite the opposite. As Steven Cook of the Council of Foreign Relations in New York put it: \textqq{From the perspective of many Egyptians, this arrangement hopelessly constrained Cairo's power while freeing Israel and the US to pursue their regional interests unencumbered. Without the threat of war with Egypt, Israel poured hundreds of thousands of Israelis into settlements in the West Bank and the Gaza Strip, invaded Lebanon (twice), declared Jerusalem its capital, and bombed Iraq and Syria}~(\href{http://mondediplo.com/2011/03/02arabworld\#nb1}{1}).

Egyptians have expressed their sympathy with the Palestinians and Lebanese whenever they have had the chance: during the war with Lebanon in 2006, portraits of the Hizbullah leader Hassan Nasrallah were displayed in Cairo shops even as the Egyptian regime condemned Hizbullah's recklessness. The protesters who fought for multi-party democracy do not much like Iran -- a non-Arab, Shia Muslim country and historic rival, whose repressive rule worsens by the day -- but they do value its refusal to bow to the diktats of the US and Israel. A more representative future government in Egypt will need to take account of popular feeling over Gaza and relations with Israel, and will probably be more wary of US attempts to form a common (if undeclared) front between Arab countries and Israel against Iran.

Egypt's room for manoeuvre will also depend on its economic base, which has been weakened by years of \textqq{liberalisation}, begun by Sadat's \emph{infitah} (opening up of the economy). Egypt remains dependent on US military and food aid, and funding from the EU, which now has a fragile economy. Some commentators suggest that Egypt could adopt an independent foreign policy like Turkey; but Turkey's diplomatic freedom is based on a dynamic economy, and a GNP three times that of Egypt's, with roughly the same population.

The upheaval in Egypt worries other Arab countries which are presented as \textqq{moderate}. Saudi Arabia's King Abdullah intervened with President Barack Obama on Mubarak's behalf. The king, and other leaders, are haunted by the fear of a decline of US power in the region. The fact that the US has managed to put together a broad front against Iran's nuclear programme and impose sanctions does not hide its failure in Iraq (US troops are due to withdraw by the end of this year, and Iraq has been affected by the protests spreading across the region), the stalemate in Afghanistan, and its inability to get the Israelis to halt expansion of settlements.

\section{Saudi media warning}\label{saudi-media-warning}

The resignation of Saad Hariri's government in Lebanon in January and the abandonment of Mubarak worsened the fears of these \textqq{moderate} leaders, already alarmed by the way the movement for democracy had spread. The youth of the Gulf are not immune to events in Tunisia and Egypt. On 16~February, the Saudi newspaper \emph{Al-Watan} called on the authorities to take account of the aspirations of young people, who \textqq{take an interest in development projects, follow their implementation and how quickly they are carried out, measure their effectiveness and cost, and share information on who gains and who loses from them} -- a reference to the corruption that blights many projects in the kingdom. Saudi Arabia was already trying a more independent path by getting closer to Syria. It responded favourably to overtures by the new Iranian foreign minister, Ali Akbar Salehi, in January.

The Palestinian Authority (PA) has lost a faithful ally in Mubarak, who was opposed to PA reconciliation with Hamas and supported its policy of negotiation with Israel. The PA has to recognise the change. In February Obama asked PA president Mahmoud Abbas to withdraw a UN Security Council resolution the PA had tabled, condemning Israeli settlement building. Abbas refused, marking a hardening of position towards the US (see \emph{West Bank wind of change}, page 6). Will the lack of political progress inspire the youth of the West Bank -- and Gaza -- to express their desire for freedom and dignity? Will they present their struggle in terms of human rights and equality, and protest peacefully in the streets, against both their leaders and the occupation? According to \emph{The Jerusalem Post}~(\href{http://mondediplo.com/2011/03/02arabworld\#nb2}{2}), the Israeli army is creating a rapid reaction force to counter this.

In Israel, Binyamin Netanyahu's government was more concerned than the US's Arab allies by events in Egypt, and made clear its strong support for Mubarak. Daniel Levy, of the New America Foundation thinktank, says this attitude illuminates Israel's frequent claim to be the \textqq{only democracy in the Middle East}: it indicates not a fear of being isolated among surrounding dictatorships but a wish to remain the only democracy~(\href{http://mondediplo.com/2011/03/02arabworld\#nb3}{3}). Successive Israeli governments have felt comfortable with pro-western authoritarian regimes because they were aware of the Arab street's solidarity with the Palestinians. For the moment Israel is paralysed, deliberately exaggerating the role of the Islamists, drawing parallels with Iran's Islamic revolution of 1979 and rattling sabres more loudly over the \textqq{Iranian threat}, which it believes the world does not understand. It has told its soldiers they might be ordered to invade Lebanon again, as minister of defence Ehud Barak warned on a recent visit to the northern front~(\href{http://mondediplo.com/2011/03/02arabworld\#nb4}{4}).

If the West has lost (with allies already overthrown), does that mean the Syria-Iran axis, and its allies Hamas and Hizbullah, have won? It does, but their weaknesses are clear. Hamas is confined to Gaza, and the likelihood that the UN special tribunal for Lebanon, into the assassination of Rafik Hariri, will indict Hizbullah's leaders is weakening the movement. The Iranian leadership may have welcomed the revolution in Egypt, but it has put down its own protesters and intensified repression.

In Syria, President Bashar al-Assad has two trump cards: fear among Syrians that unrest will lead to Iraqi-style instability and sectarian conflict, and his firm stance against Israel, which has popular support. However, economic liberalisation and a fast-growing population mean Syria faces severe economic and social problems. Young Syrians want freedom too.

\section{Palestine not forgotten}\label{palestine-not-forgotten}

The US adapted well to the fall of Suharto in Indonesia, but the situation now in the Middle East is very different -- mainly because of Palestine, which many commentators mistakenly believe was a minor issue for the protesters. The organisers of Cairo's protests banned anti-American and anti-Israeli slogans, deciding to concentrate on opposing the Mubarak regime. But at the huge victory celebration in Cairo on 18~February, after Mubarak stood down, many protesters chanted for the liberation of Jerusalem.

For decades the US has been able to give Israel almost unconditional support with impunity: Arab leaders have remained faithful, and the US has cared little about being unpopular on the Arab street. But this is coming to an end. In March 2010, General David Petraeus, then head of US Central Command, said: \textqq{Arab anger over the Palestinian question limits the strength and depth of US partnerships with governments and peoples in the {[}region{]} and weakens the legitimacy of moderate regimes in the Arab world}~(\href{http://mondediplo.com/2011/03/02arabworld\#nb5}{5}). The new geopolitical context will force the US administration to make crucial choices, but does it have the will, and ability, to do so?

These questions also apply to the EU, which has been compromised by its staunch support for Ben Ali and Mubarak. The EU was incapable of maintaining distance from dictators, has made many agreements with an Israeli government that is hostile to peace, and has promoted neoliberal economic policies that have worsened poverty and facilitated massive corruption south of the Mediterranean. Will it now have the courage to listen to the Arab street, which is not in fact a crowd of bearded fundamentalists and women in niqabs? Perhaps, as the Lebanese writer Georges Corm suggests, civil society in the North should follow the Arab example and \textqq{raise the level of protest against the dreadful neoliberal oligarchy that impoverishes European economies, creates too few jobs and every year forces more Europeans of all nationalities into insecurity. This backwards evolution benefits a narrow layer of managers whose annual pay eats up more and more of the nations' wealth}~(\href{http://mondediplo.com/2011/03/02arabworld\#nb6}{6}).

In only a few years, the world has become polycentric. Every large country, including Brazil, China, India and South Africa, is trying to find its place -- neither in opposition nor subservient to the US, but beside it, defending its own interests. Turkey is a member of Nato and a US ally, but plays an important role in the region by taking an independent stance towards Iran's nuclear programme and Palestine~(\href{http://mondediplo.com/2011/03/02arabworld\#nb7}{7}). North Africa and the Middle East want to join this global movement. \textqq{What the people of the region demand,} wrote Graham Fuller, former CIA officer and author of \emph{The Future of Political Islam}, \textqq{is to be able to take control of their own lives and destinies. ... In the near term, the prescription is stark -- Washington must back off and leave these societies alone, ending the long political infantilisation of Middle Eastern populations ... based on a myopic vision of American interests}~(\href{http://mondediplo.com/2011/03/02arabworld\#nb8}{8}).

\textqq{Neither East nor West} chanted Iranian protesters in 1979, opposing both the US and the Soviet Union. \textqq{Neither with the West nor against it} could be the slogan now across the Arab world, expressing a desire for independence and sovereignty in a multi-polar world. They will judge the West by its ability to defend the principles of justice and international law everywhere, particularly in Palestine. But they will no longer allow their governments to use the struggle against the West to justify tyranny.

\end{document}

