% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Sam Pizzigati}
\title{ The war on salaries - Enough is enough  }
\date{2012, február 6 \thanks{LMD } }
\maketitle

US radicals came up a century ago with sound proposals for a maximum income, enforced through progressive taxation, to ensure that the rich couldn't so easily buy political influence, as well as to adjust inequality

The Occupy Wall Street movement hasn't yet demanded a cap on individual income but it probably will. Ever since the \textqq{golden age} after the Civil War, great American popular surges for economic justice always come to demand an income cap, now called a \textqq{maximum wage}. This is more than a cap on salary, it means total annual income; and the term derives from the more familiar concept of a \textqq{minimum wage}.

America's first serious maximum wage proposal was made by the philosopher Felix Adler, remembered as the founding chairman of the National Child Labor Committee, the early 1900s advocacy group that led the campaign to end the exploitation of the young. The exploitation of workers young and old, he believed, generated private fortunes that exerted a \textqq{corrupting influence} on US politics. To curb this, he proposed a steeply graduated income tax --- with a 100\% top rate at the point \textqq{when a certain high and abundant sum has been reached, amply sufficient for all the comforts and true refinements of life} (1).

The New York Times gave Adler's call ample publicity, but the idea didn't take a specific legislative form until the first world war, when progressives demanded a 100\% tax on all income over \$100,000 to help finance the war effort. The group backing this, the American Committee on War Finance, would assemble a network of 2,000 volunteers across the country and put ads in newspapers that readers could sign to pledge their commitment \textqq{to further the prompt enactment into law} of the boldest tax proposal any American political grouping had ever advanced. They were demanding a fixed limit on income, \textqq{a conscription of wealth}.

The committee chairman Amos Pinchot, a New York attorney, declared: \textqq{If the government has a right to confiscate one man's life for public purposes, it certainly ought to have the right to confiscate another man's wealth for the same purposes.} He later testified to Congress that the richest 2\% of Americans owned 65\% of the nation's wealth. \textqq{Neither the United States nor any other country can carry on a war which will make the world safe for democracy and the plutocracy at the same time,} Pinchot told lawmakers. \textqq{If the war is to serve God, it cannot serve Mammon} (2).

A changed discourse

Pinchot and his fellow progressives did not manage to enforce the rate. But by war's end their campaign had totally changed the tenor of the US's political discourse on taxes and the top rate, on incomes over \$1m, just 7\% in 1914, would rise to 77\% in 1918.

The \textqq{red scare} that followed the first world war in the US quickly dashed progressive hopes for a more egalitarian nation --- and ushered in a right-wing political reaction that again made the country safe for plutocracy. Incomes and wealth concentrated at a ferocious pace throughout the 1920s and, in Congress, both Democrats and Republicans pushed hard for lower taxes on the richest. By 1925, no income over \$100,000 faced more than a 25\% tax rate.

The crisis of 1929, which almost collapsed the economy, changed things. By 1933, 25\% of US workers were unemployed and there was a renewed call for income caps. From the state of Louisiana, a flamboyant young senator, Huey P Long, mobilised a Share Our Wealth movement that swept the nation, urging a \$1m cap on individual annual income (the equivalent of \$15m in 2010) and an \$8m cap on individual net worth.

President Franklin Roosevelt tried to steal Long's thunder in June 1935, outraging corporate America and the nation's deepest pockets with a \textqq{soak the rich} tax plan that, later that year, raised the top tax rate on income over \$5m (\$78m in 2010) to 79\%. This manoeuvre --- and the assassination of Long in August 1935 --- removed income caps from the agenda. By April 1942 they were back: FDR, inspired by trade unions, called for a maximum wartime income of \$25,000 a year (\$350,000 in 2010). In 1944 Congress hiked the top tax rate on income over \$200,000 to a record 94\%.

For the next two decades America's top tax rate hovered around 90\%, before dropping to 70\% under Lyndon Johnson (November 1963-January 1969). Under Ronald Reagan, the top tax rate dropped to 50\% in 1981, then to 28\% in 1988. The current top rate is 35\%. But this overstates today's tax burden of America's rich: much of their income comes from capital gains --- the profits they make buying and selling stocks, bonds and other assets --- which are taxed at only 15\%. In 2008, the US's 400 highest-earning taxpayers had some \$270.5m each in income. They paid just 18.1\% on that, exploiting all the loopholes in federal income taxes. In 1955, they had averaged only \$13.3m (in today's dollars) and paid 51.2\% of that in tax.

A true 'maximum wage'

Today the heirs to Adler, Pinchot and Long are focused on enterprise more than income tax. They advocate that US authorities --- local, state, and national --- leverage the power of the public purse to deny tax dollars to corporations that pay their top executives high multiples of workers' earnings. Almost every major US corporation currently depends on tax dollars. Companies get tax dollars to perform government contracts or for economic development subsidies or, indirectly, via tax breaks and preferences. No tax dollars should go to corporations that pay their executives over 10 or more times what their workers are making (3). \textqq{The federal government currently denies contracts to companies that increase, through discriminatory employment practices, racial or gender inequality in the United States,} notes an Institute for Policy Studies report. \textqq{The same principle could be invoked to deny contracts to companies that, through excessive executive compensation, increase the nation's economic inequality} (4).

The ultimate goal is a true \textqq{maximum wage}, tied to the minimum wage, to be enforced through a progressive income tax, just as Adler proposed over a century ago. The maximum would be set as a specific multiple of the minimum wage and all income over a given multiple of that minimum would then be subject to a 100\% tax. This would encourage and nurture a solidarity economy: society's most wealthy would have a vested interest in the well-being of society's least wealthy.

Before Occupy Wall Street, this was political fantasy. No longer. In a sign of our changing times, two respectable US academics --- a law professor at Yale and a Berkeley economist --- have recently published in The New York Times a cogent case for tax reform that would limit the average incomes of America's richest 1\% to 36 times the nation's median income (5). Today we take the idea of a minimum wage for granted. Why not a maximum?

Sam Pizzigati is an associate fellow at the Institute for Policy Studies in Washington DC, editor of the journal Too Much and author of The Rich Don't Always Win, Seven Stories Press, New York, forthcoming 2012

(1) Felix Adler, \textqq{Social Reform: Proposing a System of Grand Taxation}, The New York Times, 9 February 1880.

(2) The Public: an International Journal of Fundamental Democracy, New York, 28 September 1917.

(3) Major corporate CEOs are currently averaging 325 times the average US weekly wage.

(4) \textqq{Executive Excess 2007: the Staggering Social Cost of US Business Leadership}, 14th Annual CEO Compensation Survey, Institute for Policy Studies, Washington DC, 29 August 2007.

(5) Ian Ayres and Aaron Edlin, \textqq{Don't Tax the Rich. Tax Inequality Itself}, The New York Times, 18 December 2011.

\end{document}

