% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Serge Halimi}
\title{ Tunisia's revolution annexed  }
\date{2013, március 1 \thanks{LMD } }
\maketitle


Almost everyone in Tunisia believes that the benefits of the revolution are in danger. Perhaps from a \textqq{secular} opposition that refuses to admit that the conservative An-Nahda Islamists were the clear winners in the National Constituent Assembly elections in October~2011. Or from the An-Nahda Islamists, who want to use their victory to infiltrate the state from within, while exploiting the fear inspired by the Salafist militias. Or simply from a political circus reminiscent of the Fourth French Republic, with parliamentary alliances breaking up as soon as a member was not appointed, big surprises forgotten next day, innumerable tiny groups endlessly redefining their position. Meanwhile, mining production is down, tourism is shaky, insecurity is worsening and hundreds of young Tunisians have left to join the jihadis fighting in Syria, Algeria and Mali.

Jihadi flags and An-Nahda Islamist flags were seen side by side in the demonstrations in Tunis on 16~February. There was a big crowd, although not nearly as big as that of opposition supporters for the funeral the previous week of Chokri Belaid, the leftwing militant assassinated by an unidentified group. His murder weakened public support for An-Nahda, united its opponents and provoked discord in its ranks. The prime minister and secretary general of the Islamist party, immediately deserted by most of his friends, proposed to form a \textqq{government of national expertise without political affiliations}. Encouraged by opposition bodies, by the General Tunisian Workers' Union (UGTT), the army, employers, Algeria, western embassies, the idea was to suspend the An-Nahda government temporarily, pending the introduction of a new constitution and fresh elections. The demonstrators on 16~February opposed this, claiming that An-Nahda was \textqq{legitimate} and that it was the victim of conspiracies by the media, foreigners, especially the French, \textqq{counter-revolutionaries} and \textqq{remnants of the old regime}.

It is surprising to hear such Jacobin tirades from such a conservative political body. Since they came to power after elections in October~2011, the An-Nahda Islamists have shown little inclination to upset the economic and social order. Like their Egyptian counterparts and their ---~dwindling~--- paymasters in the Gulf monarchies, they have tried to combine extreme capitalism with ancient family and moral values. Their speeches appeal to similar parties by describing opponents in the strongest terms: \textqq{First they cut the roads and block the factories, and now they are claiming the government is not legitimate,} Rachid Ghannouchi told his partisans. \textqq{An-Nahda is Tunisia's backbone. To break it or exclude it would damage the country's national unity.}

That's what the argument is all about.

\section{Shared plan to break up the state}\label{shared-plan-to-break-up-the-state}

Where does national unity begin and end? What sacrifices must the Tunisians make and what risks must they take to preserve it? That an Islamist party held a key position in the government was not an issue a year ago when it was a matter of drafting a new constitution, not very different from the old one, and readjusting economic development to the benefit of provinces that had been neglected for decades. But the question is not quite the same when An-Nahda's failure ---~the constitution not yet adopted, public order in danger, no investors in sight, poor regions still as poor as ever~--- lends strength to more radical Islamist groups that must be included in the political game lest they take up armed violence. But, while co-opting these groups might gradually normalise a few extreme militants, the corollary is further Islamisation of Tunisian society.

Hence the opposition suspicions. Far from conceding that dialogue and persuasion have so far enabled An-Nahda to forestall a violent expression of Salafist and jihadi views, the opposition considers that there are no firm distinctions between these groups and that they all harbour the same political and religious plan to break up the state. There is a video from last April that has become famous, in which Ghannouchi explains to the Salafists that they will have to be patient, suggesting that they and An-Nahda might have simply split tasks to achieve their common aim: one group to make placatory speeches, the other to scare off opponents. The obscurity of the internal workings of An-Nahda tends to confirm this interpretation.

But there is a danger of underestimating the tensions within the ruling party, seen to spectacular effect in the latest government crisis. In a thoughtful and informative report on the Salafist challenge, the International Crisis Group concluded that \textqq{An-Nahda itself is divided: between religious preachers and pragmatic politicians as well as between its leadership's more flexible positions and the core beliefs of its militant base. Politically, such tensions give rise to an acute dilemma: the more the party highlights its religious identity, the more it worries non-Islamists; the more it follows a pragmatic line, the more it alienates its constituency and creates an opening for the Salafis}~(\href{http://mondediplo.com/2013/03/01tunisia\#nb1}{1}).

However, the opposition is unwilling to admit that the worst has been avoided thanks to An-Nahda, still less to accept that the Islamisation of public institutions ---~education, culture, justice~--- in a country with 11~million inhabitants may be the price for quelling the violent tendencies of some 50,000~jihadis. Stirred by Chokri Belaid's assassination and emboldened by the immense crowd at his funeral, it doesn't give much credence to the An-Nahda leader's troubles. \textqq{Ghannouchi was never willing to denounce the Salafists or the jihadis,} said Riadh~Ben~Fadhel, leader of a centre-left opposition group, the Modernist Democratic Pole. \textqq{He said they were the heart and soul of the revolution, they reminded him of his youth, they were part of the Islamist family, they were lost sheep. With them, Ghannouchi has an enormous pool of potential voters and a militant intervention force at his disposal, enabling him to mount a direct attack on the democratic camp, employing highly organised militias and never showing his own face. He uses them to do his dirty work. But now the masks are off.}

\section{UGTT under attack}\label{ugtt-under-attack}

The tone is little softer on the UGTT side~(\href{http://mondediplo.com/2013/03/01tunisia\#nb2}{2}). The two main forces are openly at war with each other. The Leagues of Protection of the Revolution (LPR) attacked the trade union headquarters in December. Six months earlier, the UGTT regional office in Jendouba had been targeted by Salafist forces. \textqq{We are fighters, accustomed to the hostility of the government and of violent groups,} said Nasredine Sassi, director of the central trade union research department. \textqq{But this is the first time the UGTT has been attacked in this way. It reflects official political statements, including statements by a number of ministers, denouncing union activities.}

The Tunisian left, in politics and associations, is now united in its opposition to An-Nahda. It no longer hesitates to describe it as \textqq{extreme right} or even as a \textqq{fascist party}. The memory of the common sufferings of democrats and Islamists during the dictatorship faded in just a few months. And at the same time the leaders of what was, under Ben Ali, the single party ceased to be ostracised.

Ahlem Belhadj, chair of the Tunisian Association of Democratic Women (AFTD), responded ironically to the suggestion of An-Nahda peacefully coopting religious extremists: \textqq{They are so well integrated that there are training camps in Tunisia and hundreds of Tunisians are dying in Syria and Mali.} She considers that An-Nahda's economic policy, which is \textqq{even more neoliberal than Ben Ali's}, leads to increased unemployment among young people in working-class districts, with the danger that some of them may become radical and violent.

Fabio Merone, a researcher specialising in Tunisian Salafism at the Gerda Henkel Foundation, also thinks that Salafism, like all jihadism, is the product of social change. He said that in Ben Ali's time the Tunisian myth worked for the middle class but excluded those who lived in a different Tunisia, who fled to Italy or joined religious groups. Salafism \textqq{does not come from the moon or from Saudi Arabia: it is the political formation of young people reacting to banishment and denied education.} But Ben Ali's cultural desert may also have prompted a quest for identity, a gap that Wahhabi preachers filled.

One of them, Saudi-trained Bechir Ben Hassen delivered an address during the demonstration on 16~February to an audience that included militant members of the ruling party, jihadi groups and ministers, including the (unveiled) minister for women's affairs. The event was duly noted at UGTT headquarters. Sassi said: \textqq{This government should be in its ministerial offices settling the Tunisians' problems, not out on the street organising demonstrations and haranguing the crowd.}

\textqq{The problems to be settled} can be identified by a look at the vacancies column in\emph{La Presse de Tunisie}. A Canadian entry on 17~February invited qualified \textqq{bricklayers, butchers, nurses and dental hygienists} to emigrate. A Tunisian semi-trailer company was seeking a warehouseman \textqq{with a university degree}.

\section{\texorpdfstring{`\textbf{It's Ben Ali all over again'}}{`It's Ben Ali all over again'}}\label{its-ben-ali-all-over-again}

`In Sassi's opinion, \textqq{The government is no nearer to solving the economic and social problems, particularly unemployment. It's Ben Ali all over again.} The UGTT, noting with dismay that the absence of regional development encouraged the informal economy, called on the government \textqq{to develop an appropriate infrastructure at Gafsa, Sidi Bouzid and Kasserine, in the border areas where smuggling is rife.} Milk, tomatoes, pasta and mineral water, often bought at state-subsidised prices~(\href{http://mondediplo.com/2013/03/01tunisia\#nb3}{3}), are exported illegally to Libya to be sold for much more --- to such an extent that there are now shortages in Tunisia and the price of basic commodities is rising sharply. As journalist Thameur Mekki said: \textqq{We haven't had to import milk since the second world war. The state stands by and does nothing. They are powerless, they strut about on TV instead of being at their desks. And when they are at their desks, they are busy Islamising the state.}

According to Workers' Party spokesman Jilani Hammami, a pillar of Chokri Balaid's Popular Front, \textqq{the new government had to start again from scratch. So there was no recovery programme. It pursued the same course as Ben Ali, counting on Qatar and Saudi Arabia and getting nothing.} An-Nahda's dream of \textqq{Muslim-Arab} solidarity never materialised. Instead of gifts from the Gulf states (the Tunisian authorities had expected \$5bn from Qatar, according to the economic information site African Manager), Tunisia got only small loans (\$500m) at relatively high rates of 2.5\%. At about the same time Japan lent it \$350m at 0.95\%.

The International Monetary Fund had a \textqq{very good opinion} of Ben Ali's Tunisia. In~2008, then-managing director Dominique Strauss-Kahn considered that \textqq{its economic policy is sound and is, I think, an example that the emerging countries would do well to follow.} Could the IMF make up for the Gulf states' shortcomings? The trade union headquarters expressed few reservations. As Sassi said: \textqq{The UGTT has nothing against the IMF. The secretary general received Christine Lagarde and delegations from the World Bank in these very offices. We know the country can no longer survive outside the global system but we are trying to guide policies in the right direction. We have told the World Bank `you supported Ben Ali. Now you must show that you are willing to support democracy with pilot projects for the development of less-favoured regions'.}

The Popular Front is more proactive. It is against Tunisia's status as special associate of the European Union: \textqq{A relatively unproductive economy based on exports and depending on highly vulnerable SMEs will always be at the mercy of key decision-makers in Europe}. It also insists that payment of the foreign debt should be suspended for three or four years, so that the 18\% of the Tunisian budget usually allocated to these payments can go to creating jobs. As Jilani Hammami says: \textqq{If France, Germany, Belgium, Italy, the United States and the Gulf states really care about Tunisia, they should suspend payment of the debt.} He does not have much hope of this.

According to Thameur Mekki: \textqq{Should purchasing power continue to decline and insecurity to increase, it will spell the end of consent for democracy. At present, the Tunisian people have no use for it.} The Salafists are well established in the least-favoured districts and they plan to take advantage of the state's failings to become key players in the economy, including the informal and underground economy, to preach and take root. \textqq{They say `look, nothing works, it's because people are not listening to the prophet'. They want to persuade them to reject elections and political parties, and accept of their own free will what the Salafists present as the final solution: strict application of Islamic law}~(\href{http://mondediplo.com/2013/03/01tunisia\#nb4}{4}).

Others are more optimistic. Ahlem Belhadj considers that women's rights are already \textqq{agreed even in parties that were not inclined to demonstrate earlier. Thanks to resistance in civil society, right and left, the laws have remained intact.} The vigilance of the popular movement, the crowd at Chokri Belaid's funeral, the incipient regrouping of progressive forces and the divisions within An-Nahda also encourage Ben Fadhel to think that \textqq{the battle to Islamise Tunisia can only be lost.}

\end{document}

