% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Laurent Geslin}
\title{ The kingdom of Greater Hungary --- again?  }
\date{ \thanks{} }
\maketitle

\begin{abstract}
The end of internal border controls after the EU enlargement of 2004 was meant to check a surge in nationalism. But in Hungary and Slovakia, as in Ukraine on the other side of the EU border, people still claim national identities.

by Laurent Geslin and Sébastien Gobert
  
\end{abstract}



\textqq{My grandfather lived in five different countries without ever leaving his home village,} said Sándor Igyártó, a retired customs officer who is a Ukrainian citizen of Hungarian origin. \textqq{The Hungarians came back in 1938, and during the second world war he was drafted into the Hungarian army and sent to fight on the eastern front. When the Soviets annexed this region in 1944 he was deported to Siberia, and he was only able to come home after Stalin died.}

The small town of Tyachiv, on the River Tisza, lies in the Transcarpathian plain, west of the Central Carpathians. Transcarpathia is in Ukraine today, but was part of the Austro-Hungarian Empire until the Treaty of Trianon of 1920, when it was ceded to the new Czechoslovak Republic. Tyachiv has Calvinist, Roman Catholic, Greek Catholic and Orthodox churches, plus memorials to partisans who fought in the \textqq{great patriotic war}~(\href{http://mondediplo.com/2013/05/11borders\#nb1}{1}), Soviet soldiers killed in Afghanistan, and Austro-Hungarian soldiers from the first world war. In the late 19th century, Transcarpathia was the centre of Europe, according to Austro-Hungarian geographers who erected a marker in 1887 a few kilometres west of Tyachiv. Today it is a forgotten region at the eastern border of the European Union, trapped on the other side of the Schengen Area's external border, which separates Hungary, Slovakia and Poland from Romania and Ukraine.

Like the rest of central Europe, Transcarpathia was for many centuries multicultural --- its population included Hungarians, Ruthenes, Ukrainians, Germans, Roma and Jews. But the simplification of identities that began in the early 20th century accelerated after Ukraine gained independence in 1991. According to census data only 150,000 Hungarians remained in Transcarpathia in 2001~(\href{http://mondediplo.com/2013/05/11borders\#nb2}{2}), 12\% of the population, compared with 17\% in 1921~(\href{http://mondediplo.com/2013/05/11borders\#nb3}{3}). \textqq{Every year five to six thousand people emigrate to Hungary,} said Igyártó. \textqq{There's no future here; the economy is in a terrible state.}

Yet many large houses have been built over the past few years in Chop, near the borders with Hungary and Slovakia. \textqq{It's no mystery: the people who own the big houses are involved in smuggling --- mostly cigarettes and people,} said a local journalist who asked to remain anonymous. Chop is close to a major railway junction that was once a gateway to the Soviet Union, but the town is cut off from Kiev by the Carpathians and more than 800km of bad roads. The Hungarian population there watch Hungarian television and set their clocks to Hungarian time, an hour behind Ukrainian time.

The border with Hungary is guarded by police with dogs and heat detectors. Every year hundreds of migrants from Pakistan, Afghanistan and Somalia try to cross it. \textqq{It costs around \$6,300, and it's virtually impossible to get across without bribing the border guards,} said Haruni, a Somali who has been waiting to cross for two years, in Uzhhorod, the regional capital. In 2012 the authorities discovered two tunnels to Slovakia and seized 13,000 cartons of contraband cigarettes, worth around \$160,000. According to sociologist Antal Örkény, most locals only survive by small-scale cross-border trafficking. They are even more cut off from the EU than they were before the creation of the Schengen Area.

\section{The young won't stay}\label{the-young-wont-stay}

Fifty kilometres south is another drowsy small town, Berehove~(\href{http://mondediplo.com/2013/05/11borders\#nb4}{4}). Dogs roam the streets and time and weather have faded the pastel colours of the old Hungarian buildings. Here too, work is getting hard to find, except in a few Italian-owned textile factories where wages are only \$315 a month. \textqq{We are trying to persuade our young people to stay, but many emigrate to Hungary as soon as they've finished their studies,} said Ildikó Orosz, chairperson of the Ferenc Rákóczi II Transcarpathian Hungarian Institute, in Berehove. \textqq{In areas where we are a minority, the children go to Ukrainian schools and are gradually assimilated, though in the past nobody spoke Ukrainian locally.} In the past, the Slav communities spoke Rusyn (Ruthenian), a language whose separate identity is disputed by Ukrainian nationalists (see~\href{http://mondediplo.com/2013/05/12borders}{\emph{Who are the Ruthenians?}}). \textqq{We suffered a great deal during the 20th century; our intellectuals were decimated. Our institute, which opened in 1996, gives young Hungarians in Transcarpathia access to a complete education in their mother tongue. The Ukrainian university doesn't give them that option.}

On the road from Berehove to Uzhhorod is Palanok Castle, former seat of Prince Ferenc Rákóczi II, who led a war of independence against the Habsburgs between 1703 and 1711. Betty Henkel, an unemployed graduate, said: \textqq{Hungarians have always lived here; Ukraine is only 20~years old. Today, they try to portray us as an anomaly but we have our own schools, our own political parties, our own cultural associations. Of course, we all live together and we do our best to get on with the Ukrainian authorities. But we have no expectations of the Ukrainian government.}

Roman Ofitsynskyi, a vice-rector of Uzhhorod National University, disapproves of this system: \textqq{The curriculum at the institute in Berehove doesn't lead to jobs. Hungarian is useless when it comes to finding work in Ukraine, and the programmes emphasise philology and history rather than vocational training. How many history teachers do they really need? Uzhhorod offers Hungarian studies and we have scholarships that are reserved specifically for Hungarian speakers, sometimes to the detriment of other students. They talk of discrimination, but in fact they are very privileged.}

Ukraine's national identity is still under construction. Language divisions between the east of the country, inhabited largely by Russian speakers, and the west, inhabited by Ukrainian speakers and many ethnic minorities, are open to political manipulation. \textqq{During the Soviet era, Russian civil servants and soldiers moved into Transcarpathia, and we experienced Russification on a massive scale,} said István Csernicskó, a vice-rector at the Ferenc Rákóczi II Institute. \textqq{So in 2004 Hungarians in Ukraine supported Viktor Yushchenko and Yulia Tymoshenko's Orange Revolution. But the new government had no intention of favouring regionalism in Transcarpathia, for fear that the east would also claim too much autonomy. On the contrary, it imposed the use of Ukrainian wherever possible. Our minorities are victims of these disputes over national identity.}

In August 2012 Viktor Yanukovych, who had defeated Tymoshenko to become president of Ukraine in 2010, signed a law that gives official status to minority languages in regions where they are spoken by more than 10\% of the population. With a general election in October, this was a good way to mobilise the Russophone electorate and win the votes of the minorities. The law was ratified by Transcarpathia's regional parliament in December.

\section{Need to use our mother tongue}\label{need-to-use-our-mother-tongue}

István Gajdos, then mayor of Berehove and chairman of the Hungarian Democratic Federation in Ukraine (UMDSZ), one of two political parties representing the Hungarian community in Transcarpathia~(\href{http://mondediplo.com/2013/05/11borders\#nb5}{5}), was elected to the Verkhovna Rada, Ukraine's parliament, nominated by Yanukovych's Party of Regions. \textqq{Hungarian is now the official language of the district and the municipality. It was essential for us to be able to use our mother tongue in our own town. The law allows us to make all road signs and official communiqués bilingual. I am aware that not all council employees speak good Hungarian, but in future job applicants will be required to have a command of the language.}

This decision is unacceptable to the Ukrainian far-right party Svoboda, which won a historic 10\% of the vote in the election and now has 38 seats in the Ukrainian parliament. Oleh Kutsin, head of Svoboda's regional branch, sees the \textqq{re-Hungarianisation} of Berehove (Beregszász in Hungarian) as a sign of the separatist tendencies of Hungarians in Transcarpathia, and of the imperialism of Budapest. \textqq{Hungary spends a million dollars a year on development projects for the benefit of Hungarians in Ukraine, and the Hungarian consulate issues passports to all comers, even though Ukrainian citizens are only allowed to have one nationality. Budapest is doing everything it can to divide the Hungarian areas from the rest of Ukraine, with a view to absorbing them into its territory in the long term. It's a serious threat to our national security.}

Since 2001 a \textqq{Hungarian status card} has made it easier for ethnic Hungarians to study and work in Hungary, and an agreement between Kiev and Budapest has allowed people living less than 50km from the border to enter Hungary without a Schengen visa. Since January 2011, 2.5 million Hungarians living outside Hungary --- in Romania, Slovakia, Serbia and Ukraine --- can also apply for a Hungarian passport, even in Ukraine, where having dual nationality is theoretically prohibited. István Tóth, consul general at Berehove, explained: \textqq{Ukrainian law doesn't punish citizens for holding two passports, as long as the matter isn't made public.} He would not say how many had asked for Hungarian citizenship~(\href{http://mondediplo.com/2013/05/11borders\#nb6}{6}), but recognised that Budapest does provided massive financial aid that keeps the Transcarpathian Hungarian Institute, cultural associations and political parties going.

\textqq{It's normal that the Hungarian government should be concerned for Hungarians in other countries. But in fact, it doesn't do enough,} said Miklós Kovács, president of the Cultural Association of Hungarians in Transcarpathia (KMKSZ), a political party affiliated to Hungarian prime minister Viktor Orbán's Fidesz. \textqq{Every year there are fewer of us and it's more difficult to mobilise Hungarians {[}in Ukraine{]} to defend their interests. We are able to keep up a certain level of cultural activity, but it's little more than folklore. In a few years, we will be gone, not necessarily as a community, but as a political force, and the Ukraine Hungarian question will be resolved once and for all.}

\section{Worrying the neighbours}\label{worrying-the-neighbours}

Many believe Orbán encourages Hungarian nationalism in other countries because he is unable to deal with his own economic crisis. Since returning to power in 2010, after a crushing victory over the Socialist Party, he has made controversial statements aimed at those who feel nostalgia for \textqq{Greater Hungary}; these worry neighbouring countries with Hungarian communities. The Fundamental Law of Hungary, the new constitution adopted in April 2011, recalls the country's Christian roots and \textqq{thousand-year} history, and states: \textqq{Hungary shall bear responsibility for the fate of Hungarians living beyond its borders.} This is strong language in a country hard hit by the economic crisis and still traumatised by the 1920 Treaty of Trianon, which lopped off two thirds of the former Kingdom~(\href{http://mondediplo.com/2013/05/11borders\#nb7}{7}).

In Miskolc, eastern Hungary's biggest industrial centre, 150km from the Ukrainian border, disused steelworks are falling into ruin. In the 1980s the Lenin Foundry employed more than 18,000 workers, and two thirds of the town's population of 200,000 was directly dependent on heavy industry. All that vanished with Hungary's transition to a market economy. \textqq{In the 1990s unemployment reached 30\% of the workforce,} said György Mike, in charge of public enterprises on Miskolc town council, which Fidesz has controlled since the last election. \textqq{The textile and banking sectors have gradually developed, replacing heavy industry, but people have got into debt and many are now ruined.} Since the economic crisis in 2008, shops in the town centre have been closing; EU subsidies paid for the renovation of the town hall, but are not enough to restart the economy. \textqq{Our last Socialist mayor tripled the town's debt; we can't ask for any more loans. The fall in the standard of living has led many former manual workers to vote for Fidesz.}

Miskolc, strongly Communist in the 1990s, is now fertile soil for the far-right party Jobbik (Movement for a Better Hungary), which won 16.67\% of the vote and 47~seats in the Hungarian parliament in 2010. In his small office, local party chief Miklós Arpád looked at a map of Greater Hungary on the wall. \textqq{Ukraine got Transcarpathia by unfair means, and the Hungarians in Slovakia and Transylvania didn't want to be cut off from their motherland. Hungary was the biggest victim of the peace treaties. Many Hungarians still live in those regions. It's our country's role to protect them.}

By issuing passports to Hungarians in other countries, Orbán may be trying to build an electorate that favours him, and revitalise Hungary's demographics. The country has lost 350,000 people since the early 1990s and its fertility rate is around 1.3 children per woman (2.1 are needed for a stable population level); perhaps attracting Hungarians back will alleviate this. Sociologist Zoltán Kántor rejects this view and sees the \textqq{Trianon passport} as a \textqq{natural affirmation of national identity}: \textqq{In the 1990s, before Hungary joined the EU, handing out passports could have encouraged Hungarians living in other countries to move to Hungary, but that this is no longer the case because the borders are open. But by tabling the question of Hungarian communities in neighbouring countries, Orbán has appropriated one of the main themes of Jobbik's campaign, and that may allow him to halt the advances the movement has made.}

\section{Many mixed marriages}\label{many-mixed-marriages}

Slovakians are watching Hungarian politics anxiously, especially as incidents between the countries are frequent. In 2010 Slovakia's prime minister Robert Fico, then at the head of a coalition government formed with the Slovak National Party (SNS), said that Fidesz wanted to \textqq{go back to the pre-Trianon Greater Hungary} and called this \textqq{a threat to Slovakia's national security. Imagine what would happen if thousands of Slovakian citizens took Hungarian nationality: Hungarian politicians would start behaving as if the southern part of Slovakia was part of their territory}~(\href{http://mondediplo.com/2013/05/11borders\#nb8}{8}). In July 2009 Slovakia banned the use of Hungarian in government and administration, and from all public spaces, and since 2010 it has prohibited nationals from holding two passports, on pain of losing their Slovak citizenship. This pleased the SNS, whose leader Ján Slota, had called Hungarians a \textqq{cancer in the body of the Slovak nation}. Cyril Lesko, SNS chief in the town of Pre¹ov, said: \textqq{It's a completely legitimate measure. Otherwise, the Hungarian minority could be tempted to secede.}

Some 500,000 Hungarians live in Slovakia, mostly near the southern border with Hungary. The village of Vel'ké Ra¹kovce, not far from Kosice, is surrounded by a marshy plain. Jakab Elemér looked south, towards the border: \textqq{We are Hungarians, but we are Slovak citizens: we have more than one identity. We don't think Budapest should be telling us what to do.} Elemér is one of 14 Most-Híd~(\href{http://mondediplo.com/2013/05/11borders\#nb9}{9}) party members of the Národná Rada, the Slovak parliament. \textqq{We left the Hungarian Coalition Party (SMK) in 2009 to found a new party, because we didn't see ourselves as an ethnic party: we encourage integration into Slovakian society, without abandoning our roots and our culture.} It's an unusual political choice --- most Hungarian parties in other countries have close links with Budapest --- but it has paid off. In the June 2010 parliamentary election, Most-Híd won 8.12\% of the vote, SMK, which takes a harder nationalist line, only 5\%. \textqq{Unlike Ukraine and Romania, we have many mixed marriages in Slovakia,} said Örkény. \textqq{The Hungarian minority have less influence socially, so it's in their interest to integrate into Slovakian society.} Örkény believes that the identity strategies of minorities are determined largely by socioeconomic circumstances.

Istvánné Szöllösi, Slovakian National Council member for the Hungarian village of Bükkszentkereszt, very near Miskolc, showed off the village hall, where concerts of Slovakian folk music are held: \textqq{Our ancestors came here in the mid-18th century, to work in the glassworks that were being built around Miskolc. The glass industry has gone now, but we are very well integrated into Hungarian society. No one dreams of leaving.} The number of Slovak speakers in the village is dwindling, even though there is still a Slovakian primary school. \textqq{If the economic situation in Hungary deteriorates, and it becomes advantageous to emigrate to Slovakia, I have no doubt that the people of Bükkszentkereszt will seriously consider returning to their roots,} said Örkény.

When the former Eastern Bloc countries joined the EU in 2004, it was hoped that the upsurge of nationalism in the 1990s would diminish with the gradual elimination of border controls. Integration was supposed to resurrect an idealised Mitteleuropa, but owing to the economic crisis and the lack of success of EU policies, it has not been possible to look beyond national priorities. Kosovo's declaration of independence in 2008 set a precedent, showing that national borders can still change.

\end{document}

