% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Martine Bulard}
\title{ Sayonara, America?  }
\date{2010, június 8 \thanks{LMD } }
\maketitle

\begin{abstract}
Tokyo reassesses its options and its allies


Prime Minister Hatoyama's resignation at the beginning of June highlights Japan's dilemma. It wants an eastern equivalent of the EU to guarantee stability and trade. But who will join, and will the US accept Japan's exit from a co-dependency guaranteed by treaty since the end of the second world war?

by Martine Bulard

  
\end{abstract}




\textqq{You can choose your allies, but not your neighbours,} said Japanese diplomat Yasushi Masaki, summing up the geopolitical options. Among his country's awkward neighbours are Russia (getting strong again), China (emerging as a world power) and North Korea (making no secret of its nuclear ambitions). The image of Japan as a kind of Alamo, defended only by the US, is popular with the Tokyo elite, many of whom are Americanophiles.

Yet 50 years after the revision of the US-Japan Security Treaty~(\href{http://mondediplo.com/2010/06/04japan\#nb1}{1}), which codifies relations and the presence of US troops on Japanese soil, the mood is not festive. The election of Prime Minister Yukio Hatoyama of the Democratic Party of Japan (DPJ) in September 2009, after more than half a century of almost uninterrupted rule by the Liberal Democratic Party (LDP), complicated the situation. Hatoyama had no wish to break the alliance but affirmed his intention of normalising relations with the US and getting it to treat Japan like any other sovereign state. He also wanted to create an East Asian Community on EU lines. Did this indicate a slight shift away from the West? The idea was enough to cause panic in the business world and incur the hostility of the rightwing opposition, a significant proportion of Japan's ultra-powerful civil service, and many in the DPJ.

The debate has centred on the US air base at Futenma, in the centre of Ginowan, a city of 90,000 on the island of Okinawa. Under an agreement signed in 1996, which expired at the end of 2009, the camp was to have been relocated to Nago, some distance to the north. This January Nago elected a mayor who opposed the move. When Hatoyama asked for time to come up with another site, opponents of the relocation went on the offensive -- and won. The backlash when Hatoyama finally gave up trying to move the base forced him to resign on 1~June.

\section{No question of breaking free}\label{no-question-of-breaking-free}

As early as October 2009 the US secretary of defence Robert Gates had declared that no other site was suitable, and refused to attend a dinner given in his honour. Even the Japanese ministry of foreign affairs (MOFA), which generally supports the US position, is still in shock; \textqq{We did have an election,} protested a high-ranking MOFA official, who listed many good reasons for honouring the agreement. Japan's foreign minister, Katsuya Okada, contradicted Hatoyama, citing political realism as justification for maintaining the air base. The mixed messages from the top mean diplomats and senior civil servants will speak only on condition of anonymity. The press agrees with Okada. Even the \emph{Asahi Shimbun}, Japan's leading centre-left daily, has cited an ancient proverb that compares an imperial edict to sweat (once out, it can never be recalled). So there is no question of Japan breaking free from the US.

Professor Hidetaka Ishida, president of the University of Tokyo's Interfaculty Initiative in Information Studies (IIIS), is not surprised: \textqq{There is a powerful American lobby which affects everyone, newspapers, diplomats and governments, regardless of their political leanings. They were educated at American universities, they meet regularly and some are friends. In their eyes, changes of government should not break the continuity. There should be no change -- as if there were no other choice. We must break free from this way of thinking, which ties us to the US.}

To understand how difficult it is for Japan to imagine a future unsecured by the US, you have to go back to the postwar period. Osamu Nishitani, a professor of philosophy at Tokyo University of Foreign Studies explained: \textqq{The US represented a political ideal then; for most Japanese, it was \emph{the} democracy.} (He is among the signatories of a manifesto against the maintenance of the Futenma air base.) You also need to go into the history of the LDP, which has many ties to the US, including the rehabilitation of senior military figures stained by war crimes whose sins have been washed away by the grace of America.

Times have changed. Masami Honto, a young company executive and peace activist, told me: \textqq{America has been weakened by the economic crisis and competition from emerging powers and is bogged down in wars that have nothing to do with us. The cold war is over, but they still talk as if it was on our doorstep.}

Japan is now the world's second-largest economy. But as a defeated nation it adopted a pacifist constitution under which it is forbidden to wage war -- although it does now have Self Defence Forces. The US-Japan Security Treaty ratified the US military presence (up to 260,000~troops) and established Tokyo's dependence on Washington. The countries entered into an unspoken pact: the US would enjoy the use of an \textqq{unsinkable aircraft carrier} while Japan would benefit from a nuclear umbrella; the US would open its market to Japanese goods while tolerating Japan's protectionist policies; Japan would pay for the maintenance of a US military presence on its territory; the US would take the lead in diplomatic matters and Japan would follow, the common enemy being communism. Nothing remains of this geopolitical landscape, but the contract survives.

In a speech at the National Defence Academy this March, Hatoyama reassured new graduates that Japan's alliance with the US remained a cornerstone of its foreign policy~(\href{http://mondediplo.com/2010/06/04japan\#nb2}{2}). But Japan wants a more balanced relationship. The cost of maintaining the US garrisons is high -- around \$3bn a year -- at a time when Japan's budget deficit is rocketing.

Another symbol of the past is the legal immunity of US military personnel. \textqq{When there's a road accident,} said a citizen of Naha, Okinawa, \textqq{first you get a white car, with the Japanese police, then a black-and-white one belonging to the US military police -- only they are allowed to deal with an incident involving a soldier.} Locals are eager to talk about serious offences that have never been brought to court, including rapes, road accidents and robberies.

\section{A new way}\label{a-new-way}

\textqq{Japan is the only country in this kind of alliance with the US,} said Councillor Yukihisa Fujita. A member of Japan's upper house and director-general of the DPJ's international department, he is one of the few in government in favour of change, although he stressed that he was speaking in his own name. \textqq{The original aim of the treaty was to ensure that Japan was protected. Since the end of the cold war, it has mainly served to bolster US strategy,} and its wars in Iraq and Afghanistan. \textqq{We must work with America, but in some other way.} He added the relationship needed to be transparent, alluding to the agreement signed by President Richard Nixon and Prime Minister Eisaku Sato in 1969, which remained a secret until March this year.

Back then, when Japan was still traumatised by the bombing of Hiroshima and Nagasaki, Sato had persuaded the Diet to adopt the Three Non-nuclear Principles, which it espouses to this day: Japan would not manufacture, possess or allow nuclear weapons on its soil. This earned him the 1974 Nobel peace prize. But he allowed the US complete freedom of manoeuvre, including the use of bases in Japan to mount overseas combat operations without prior approval from the Japanese government (although the treaty officially required it), and the creation of nuclear weapons depots. To most experts and political leaders, the arrangement had not been a secret for a long time. But, with the exception of the communists and the pacifists, the conspiracy of silence lasted for 40 years.

Even now, it is difficult to get a clear picture: the division is not simply between advocates of independence and pro-Americans. Among those who would like Japan to break free from US influence are rightwing nationalists, staunch militarists and even a few who remember the pre-war empire with nostalgia. They see the renegotiation of the treaty as an opportunity to end Japan's exceptionality and its pacifist constitution, especially the famous Article 9, which prohibits offensive military action~(\href{http://mondediplo.com/2010/06/04japan\#nb3}{3}). Since the US began pressuring Japan to speed up its rearmament so that it could reduce its own military spending, they have found themselves in the same camp as those who favour close ties with Washington. As Osamu Nishitani pointed out, \textqq{Japan is almost the only country with a pacifist constitution, but it has the fifth highest military spending.}

Since the first Gulf war (1990-1), to which Japan felt unable to send troops, in spite of US pressure, Japanese governments have continued to modify the law (in 1992, 1999 and 2001) to allow overseas operations. However, these are restricted to peacekeeping, humanitarian work and election observation, under a UN mandate. This has resulted in attempts to justify the unjustifiable. Sending troops to Cambodia, Mozambique and Rwanda met with little opposition, but the mission to Iraq was seen as a major break with the principles, as was the dispatch of supply ships to the Indian Ocean, to support Nato operations in Afghanistan. Hatoyama kept his promise and recalled the ships this January However, the joint development of a missile defence shield in the Pacific with the US is still on the agenda.

Japan's defence minister, supported by the foreign minister, makes no secret of his ambition of deploying troops overseas~(\href{http://mondediplo.com/2010/06/04japan\#nb4}{4}). But the two smaller parties in the coalition government, the Social Democratic Party (SDP) and the New People's Party, are against it. And Japan's pacifist movement, although not as strong as it used to be, is still active.

Hatoyama finally bowed to pressure from the Americanophiles and decided to revert to the 1996 plan for Okinawa~(\href{http://mondediplo.com/2010/06/04japan\#nb5}{5}). During a visit to the island, he apologised its people and asked them to continue to bear the \textqq{burden} of the bases for a little while longer. This drew further censure from the media and the opposition -- after all, the US bases are supposed to ensure national security rather than be a burden. The leader of the SDP, Mizuho Fukushima, resigned as consumer and social affairs minister and withdrew her party from the ruling coalition.

Before Hatoyama's resignation, Yukihisa Fujita said he felt the treaty should be reviewed: \textqq{US troops should remain in Japan, but their location should be reviewed, and the number of bases reduced.} He believes there are reasons for a US military presence. \textqq{First, we don't have adequate forces without rearming -- which is not on the agenda. Second, neighbouring countries would be very worried to see Japan increasing its military strength. They are reassured by the fact that we are working with the US.}

It is not certain that all Japan's neighbours see the US as guarantor of security and peace in Asia. But they would like it less if Japan were to return to its militarist ambitions. In the early 2000s Prime Minister Junichiro Koizumi's visits to the Yasukuni Shrine, where Japanese war criminals are venerated alongside other war dead, led to protests in China and South Korea, where they were seen as provocation. The visits have ceased, but Hidetaka Ishida feels that Japan has still not settled its account with the past. Diplomats and experts agree that this is an impediment to Japan's ambitions in Asia, although the continent is now a main priority.

\section{China is the prize}\label{china-is-the-prize}

The big prize is China, whose huge market guarantees Japan's future. After more than a decade of crisis, the Japanese economy is still geared to exports, although these are in freefall. Asia, which seems to have coped better with the financial storm, has replaced the West and China has replaced the US as Japan's main trading partners. In 2009 nearly a quarter of Japan's exports went to China, while 16\% went to the US and 12\% to Europe.

But trade has not been enough to soften attitudes. China and Japan share cultural roots (including their writing systems, Confucianism and Buddhism), which should encourage cooperation, but the past is also charged with conflict. In ancient times, the emperors of China regarded Japan as a nation of \textqq{dwarves}, barely worthy of paying China tribute. In the early 20th century, the Japanese pursued their colonial conquests brutally, from the occupation of Manchuria to the Nanking massacre. The Japanese are the only nation to have been nuclear-bombed, but, unlike the Germans, not all have learned from the consequences of aggression. Even today, the museum attached to the Yasukuni Shrine justifies the Pacific war as a \textqq{defensive war}.

The tensions have subsided. In Taiwan -- an issue on which Beijing and Tokyo differ -- the election of Ma Yingjeou, who, unlike his more pro-independence predecessor, favours cooperation with China, has allowed the resumption of less adversarial relations. In Japan, the end of the Koizumi administration signalled a return to normal relations. Hatoyama suggested to President Hu Jintao that the East China Sea should be transformed into a \textqq{sea of fraternity}~(\href{http://mondediplo.com/2010/06/04japan\#nb6}{6}), an allusion to the territorial dispute regarding the Senkaku/Diaoyu islands~(\href{http://mondediplo.com/2010/06/04japan\#nb7}{7}). On a visit to Tokyo in December 2009, China's vice president, Xi Jinping, was able to bypass normal protocol and gain an audience with Emperor Akihito. The influential DPJ figure Ichiro Ozawa led a 600-strong delegation to China last December.

Professor Ryosei Kokubun of Keio University, Tokyo, feels considerable progress has been made, although he warns against taking a simplistic view. There is no Sino-Japanese alliance seeking to drive the US out of Asia, he says; it is a three-way relationship (US-China-Japan). Relations had their ups and downs during the 20th century: a US-Chinese alliance against Japan during the second world war was succeeded by a US-Japanese alliance against China, at least \textqq{until the 1970s, when all three saw the Soviet Union as the enemy}. America's U-turn over China took Japan by surprise. The Japanese still talk of the \textqq{Nixon Shock}, referring to President Richard Nixon's 1972 visit to and recognition of China, which the US had until then shunned. They had the same kind of shock in 1998: \textqq{President Clinton stayed in China for more than a week, without coming to Tokyo,} said Kokubun. \textqq{People started to worry about America bypassing Japan and feared that a new strategic partnership would be formed at Japan's expense.}

The return of a Democrat to the White House revived their fears. Officially, everyone applauded Barack Obama's speech on nuclear disarmament, \textqq{which we have wanted for a long time,} said an MOFA official. But they also noticed the length of the US president's visit to China, compared to his brief stopover in Tokyo last November.

\section{Arc of freedom}\label{arc-of-freedom}

Japan's new administration is keen to avoid the establishment of a one-to-one relationship between Washington and Beijing, from which it would be excluded. Hence the rapprochement with China and a determination to earn political credibility in the region. It is true that the concept of the East Asian Community does not belong exclusively to Japan (it first emerged after the Asian financial crisis of the 1990s, when it was opposed by the US and China), but Hatoyama revived the long term idea of a common (Asian) currency, to mark Asia's new role in world affairs. His resignation is unlikely to change Japan's desire to free itself from the International Monetary Fund and the burden of the dollar. An embryonic Asian monetary fund, in which South Korea is participating, is already in place. Even so, the Asian leadership contest has yet to be decided.

China seized the diplomatic initiative by signing a free trade agreement with the 10 Asean countries~(\href{http://mondediplo.com/2010/06/04japan\#nb8}{8}), which came into force on 1~January 2010. While trying to catch up, Japan has turned to India, Australia and New Zealand to form an \textqq{arc of freedom and prosperity}, in opposition to China's authoritarianism. The Japanese government is keen to woo India, which it sees as the perfect counterweight to China. Japan and India signed a strategic partnership agreement in October 2008 and are planning joint military manoeuvres. This privileged relationship is promising but as yet marginal: India accounts for less than 1\% of Japan's trade.

Some members of the Japanese administration have greater hopes of a Japano-Korean axis, similar to the Franco-German axis in Europe. An expert on Asian relations at the MOFA admitted, \textqq{We speak the same economic language as China, but we still have great differences over the rest. In southeast Asia, there are only two countries that have both a market economy and democracy: Korea and Japan. They could be an engine of regional cooperation.} However, China may be scary, but Japan does not inspire confidence.

Historical disputes remain an obstacle to a new alliance. In spite of three years working together on a 2,200-page report published this March, a commission of South Korean and Japanese historians was unable to agree on key issues such as Japan's imposition of forced labour and recruitment of \textqq{comfort women} (the euphemism for those forced to work in Japanese military brothels) during the second world war. \textqq{It will require patience,} admitted the MOFA expert. \textqq{Political and trade negotiations are already much easier. Now that it has joined the developed nations, Korea is more confident. And Hatoyama's people are ready to take a calmer look at history.} But a trilateral alliance between Japan, China and South Korea has quietly come into being. Brought together by economic issues, they have been meeting on the fringes of Asean summits since 1999. In December 2008, they met independently for the first time. \textqq{The East Asian Community concept is taking shape at last,} remarked Kokubun.

But judging from their different approaches to North Korea, it will be a long march. So far, China has favoured pressure and negotiation, South Korea has taken a firm stance and Japan has refused to talk to North Korea. North Korea's missile launches across Japanese airspace and its nuclear ambitions worry Japan, even if privately nobody really believes it is a serious threat.

Another source of friction that prevents Japan from repositioning itself strategically is Russia. Things have not moved on from the second world war. Russia and Japan have yet to sign a peace treaty, owing to their dispute over the Kurile Islands, which Japan refers to as the Northern Territories. Keen to secure energy resources over which Japan is in competition with China, Hatoyama was negotiating.

Sixty years after its defeat in the Pacific and 20~years after the collapse of the Soviet empire, Japan is still looking for a way into the post-cold war era. Hatoyama tried to secure greater autonomy from the US and was pushing for a new kind of regional cooperation, possibly with China. But he was hesitant to take the plunge and, lacking an innovative strategic vision and with very little support from his own party, he resigned. Does this mean that Japan has once again missed an opportunity to draw a line under the cold war? It is too soon to draw such a conclusion.

\end{document}

