% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Serge Halimi}
\title{  Paris attacks: terror at home and away }
\date{ \thanks{} }
\maketitle

\begin{abstract}
  \textsc{Paris attacks: terror at home and away}

\textbf{This is a dumb war}

\textsc{by Serge Halimi}

\end{abstract}


A~little-known US senator named Barack Obama said in 2002: \textqq{I don't oppose all wars. {[}...{]} What I am opposed to is a dumb war. What I am opposed to is a rash war. {[}...{]} A war based not on reason but on passion.} Americans were angry after 9/11 and President George Bush Jr chose to channel their anger not against Saudi Arabia (most of the Al-Qaida terrorists involved came from there), but against Iraq, which the US invaded six months later. The media wanted the war and most Democratic senators, including Hillary Clinton, favoured it too. But the invasion of Iraq created the chaos that produced so-called Islamic State (ISIS).

The Paris killings of 13~November are about to help realise ISIS's two main objectives. The first is to create a coalition of \textqq{apostates}, \textqq{infidels} and \textqq{Shia renegades} who will come to fight it in Iraq and Syria, then in Libya. The second is to make the majority of westerners believe that their Muslim compatriots could be a fifth column hiding in the shadows, a \textqq{domestic enemy} in the service of the jihadists.

War and fear --- even an apocalyptic objective contains a grain of rationality. The jihadists have calculated that the \textqq{crusaders} and \textqq{idolaters} may launch airstrikes on Syrian cities or patrol Iraqi provinces intensively but will never manage to occupy an Arab country for long. ISIS also hopes that its attacks in Europe will stir up mistrust of western Muslims, and lead to heavy policing of them. This will breed resentment, and some will want to join the \textqq{caliphate}; only a very few, but then the followers of Salafist jihad aren't trying to win an election. In fact, an anti-Muslim party win would advance their cause.

\textqq{France is at war,} François Hollande told the French congress on 16~November. He has been trying for a long time to intervene militarily in Syria, and has been pushing for greater US involvement. What is peculiar is that Hollande now wants to fight ISIS in Syria, yet two years ago, seized by the same war fever, he was trying to convince the US to punish Bashar al-Assad's regime.

Will Obama persist in opposing Hollande's \textqq{dumb} war? The pressure on Obama is all the stronger because ISIS wants the same thing as Hollande. As Pierre-Jean Luizard, a researcher at France's Centre National de la Recherche Scientifique (CNRS), says, it was at first \textqq{as if ISIS had consciously made a list of everything that would disgust popular opinion in the West: infringing the rights of minorities and of women, particularly through forced marriage, executing homosexuals, reinstating slavery, {[}...{]} beheadings and mass executions}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-1}{1}). When this macabre catalogue was not enough, ISIS cut the throat of an American hostage, posting a video of it, then carried out deadly shootings in Paris. At this point, ISIS expected the \textqq{crusaders} to respond.

A head of state is almost duty bound to react to such spectacular acts. He is under political pressure to announce some kind, almost any kind, of response --- the destruction of a warehouse or a munitions depot, airstrikes on a city. He is expected to show determination, to promise new, even tougher, legislation and condemn those in favour of appeasement. He must use martial language, talk of blood and assert that retaliation will be ruthless. Thus, he will seek standing ovations, and his approval rating will go up ten points. Eventually, all of this will prove \textqq{dumb} --- but not until a few months later. And the temptation to escalate grows ever stronger, especially with frenetic 24/7 news coverage making it seem that every act, every statement, requires an immediate answer.

During the Gulf war in 1991, US hawks criticised George Bush Sr for not ordering the troops that had just freed Kuwait to go on to Baghdad. Four years later, chief of staff General Colin Powell justified their relative restraint: \textqq{From the geopolitical standpoint, the coalition, particularly the Arab states, never wanted Iraq invaded and dismembered.~{[}...{]} It would not contribute to the stability we want in the Middle East to have Iraq fragmented into separate Sunni, Shia, and Kurd political entities. The only way to have avoided this outcome was to have undertaken a largely US conquest and occupation off a remote nation of twenty million people. {[}...{]}~It is nave however, to think that if Saddam had fallen, he would necessarily have been replaced by a Jeffersonian in some sort of desert democracy where people read the Federalist Papers along with the Koran. Quite possibly, we would have wound up with a Saddam by another name}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-2}{2}). In 2003 George Bush Jr completed his father's military project. The neocons hailed in him a new Churchill, courage, even democracy. But Powell had forgotten to read his own book, as the fears he had once expressed came true under the president he was serving as secretary of state.

Bush Jr was criticised for the childish, almost criminal naivety of his war on terror. He seems to have found his true heirs in Paris. \textqq{Let's put it simply,} France's foreign minister Laurent Fabius said, talking down to us like a teacher to a class of small children. \textqq{ISIS are monsters, but there are only 30,000 of them. If all the countries in the world are unable to eliminate 30,000~people who are monsters, then nothing makes sense any more}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-3}{3}).

Let's try to explain it to him: the 30,000 monsters have widespread support in the Sunni regions of Iraq and Syria, where the armies they face are often seen as instruments of Shia dictatorships, themselves responsible for many massacres. That is why ISIS was able to capture some cities without any fighting, when the soldiers holding them fled, abandoning their weapons and uniforms. The US has tried funding the training and equipment of more than 4,000 \textqq{moderate} Syrian fighters but, according to the Americans, only four or five are operational --- and the unit cost has been several million dollars. At Mosul, 30,000~Iraqi troops were defeated by 1,000 ISIS fighters, who captured more than 2,000~armoured vehicles and hundreds of millions of dollars from the vaults of local banks. At Ramadi, the jihadists defeated 25~times their number of Iraqi troops. Syria's armed forces are exhausted by four years of war. And the Kurds are not prepared to die for territory they do not claim. \textqq{In reality,} Luizard observed, \textqq{ISIS is only strong because its opponents are weak, and is flourishing on the ruins of institutions that are in the process of collapsing}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-4}{4}).

It's the same in Libya. Under the influence of strong emotions and led by the shock team of Nicolas Sarkozy and Bernard-Henri Lévy, France made an important contribution to the fall of Muammar Gaddafi. It imagined that getting a dictator lynched would be enough to bring about a western-style liberal democracy. But Libya has fallen apart and ISIS controls several cities from which it attacks neighbouring Tunisia. France's defence minister has admitted: \textqq{I am very concerned about Libya. Daesh {[}ISIS{]} has moved in, taking advantage of internal clashes between Libyans,} but \textqq{if Tobruk and Tripoli were to work together, Daesh would no longer exist}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-5}{5}). That problem had presumably been solved, three years ago, when Lévy explained: \textqq{Contrary to what the Cassandras predicted, Libya has not split into three confederate entities. {[}...{]} Tribal law has not prevailed over the sense of national unity. {[}...{]} Compared to Tunisia and Egypt, Libya appears to have achieved a successful {[}Arab{]} Spring --- and those who helped it can be proud of themselves}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-6}{6}). Proud indeed: apart from Bernard Guetta, who broadcasts the French foreign ministry's viewpoint~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-7}{7}), nobody is better at tall stories.

Hollande now wants \textqq{a grand and unique coalition} against ISIS. This would include Assad. But Assad has already replied: \textqq{You cannot fight Daesh and still be allied with Qatar and Saudi Arabia, which are arming the terrorists}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-8}{8}). President Vladimir Putin feels that Turkey, another presumed member of the coalition, has stabbed Russia in the back by shooting down one of its planes on 24~November. As soon as the motley coalition that France is trying to cobble together had won the war, it would face the question of what next, under even more difficult conditions than in Afghanistan, Iraq or Libya. US neocons have already forgotten all these failures (as has Hollande), and are demanding that 50,000~troops be sent into the ISIS-occupied zone~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-9}{9}).

In~\emph{Foreign Affairs,}~Steven Simon and Jonathan Stevenson, experts on the Middle East, list the conditions for a sustainable western military success on territory currently controlled by ISIS: \textqq{the support of the American public; a large cadre of deployable civilian experts in reconstruction and stabilisation; deep knowledge of the society for whose fate a victorious United States would take responsibility; {[}...{]} a sustained military force to provide security for populations and infrastructure {[}...{]} local constituents or clients, or indeed allies, to assist.} They point out that \textqq{if this sounds familiar, it is because it is the same list of things that Washington wasn't able to put together the last two times it launched major military interventions in the Middle East {[}Iraq and Libya{]}. {[}...{]} The United States would likely lose another war in the Middle East for all the same reasons it lost the last two}~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-10}{10}).

France, already heavily engaged in Africa, cannot win a war in the Middle East. The fact that ISIS is trying to draw it into this trap should not lead Hollande to rush into it, taking with him a coalition of countries that are often more cautious. Terrorism kills civilians, but so does war. The intensification of western airstrikes in Iraq and Syria, which will create as many jihadist fighters as they kill, will not restore the territorial integrity of those countries, nor the legitimacy of their governments in the eyes of their peoples. A lasting solution will depend on the peoples of the region, on a diplomatic solution, not on former colonial powers or the US, which are disqualified both by their support for the worst policies of Israel and by the disastrous results of their military adventurism --- disastrous from their own viewpoint too, since by invading Iraq in~2003, after supporting Saddam Hussein in his war against Iran (which killed more than a million) for eight years, they turned Iraq into an ally of Iran. And states that sell arms to the oil dictatorships of the Gulf, propagators of Salafist jihad, are not qualified to talk of peace, or to teach Arabs the virtues of pluralist democracy.

Historian Eric Hobsbawm wrote that when they operate in stable states with stable regimes and do not have significant support from a section of the population, small groups of terrorists are a police problem, rather than a military one. He added that it is understandable that such groups make the population very nervous, especially in major western cities, and especially when government and media are working together to create a climate of fear~(\href{http://mondediplo.com/2015/12/01dumbwar\#nb15-11}{11}).

This creation of a climate of fear, and repeated denigration of those who refuse to face up to reality, make it possible to stifle the voices of those who reject the accumulation of repressive measures that are not only ineffective but threaten civil rights. Xenophobic measures (as demanded by the National Front) have been added to the mix, such as revoking the French nationality of some citizens with dual nationality. The declaration of a state of emergency was approved almost unanimously by French parliamentarians, and, as if this was not enough, the prime minister asked them not to refer to the constitutional council the legally shaky measures he wanted them to approve.

Obama told Bush in 2002: \textqq{You want a fight, President Bush? Let's fight to make sure that {[}...{]} the arms merchants in our own country stop feeding the countless wars that rage across the globe. {[}...{]} Let's fight to make sure our so-called allies in the Middle East {[}...{]} stop oppressing their own people, and suppressing dissent, and tolerating corruption and inequality {[}...{]} so that their youth grow up without education, without prospects, without hope, the ready recruits of terrorist cells.} Obama has not taken his own advice, and neither have other heads of state. Hence, the situation we are in today. ISIS attacks and France's disastrous foreign policy have led to a new \textqq{war}, solely military and therefore already lost.

\end{document}

