% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Cédric Gouverneur}
\title{ Poland's populist revenge  }
\date{2016, március 15 \thanks{LMD } }
\maketitle

\begin{abstract}
  \textsc{Protest against darwinian capitalism}

\textbf{Poland's populist revenge}

\textbf{The poor, old and workless helped elect Poland's rightwing government, hoping for better welfare --- and respect.}

\textsc{by Cédric Gouverneur}

\end{abstract}


Adam Kalabis, 46, and I drank tea at the Solidarity office at Pokoj (Peace) colliery in Ruda Slaska (Silesia). He had close-cropped hair and the build of a wrestler, and his huge hands were still dark with coal dust, half an hour after the end of his shift 800m underground. \textqq{There are 4,000~employees here, half of them miners. I've worked here since I was~18. I began carrying sacks of coal. Now I'm in maintenance.} Kalabis said he's been \textqq{worn out by the mine. I hope I don't end up like my father: he retired at~45 and died a year later.} But Kalabis is still a long way from retirement: \textqq{The previous government decided sick days and days when we gave blood wouldn't count {[}towards their period of continuous employment{]}.} Some miners used to donate blood to get a day off. \textqq{The liberals even stopped counting years of military service performed under the Communists.}

Kalabis works a seven-and-a-half-hour day, five days a week for the publicly owned KW company and receives 2,900~zloty (less than~\$740) a month. \textqq{My wages have increased by 150~zloty {[}\$38{]} in 15~years. Even so, I'm better off than some. The widow of a friend who was killed in the Halemba methane explosion {[}which caused 23~deaths in November~2006{]} got compensation for six months and then nothing. Everyone in my family has worked in the mines, for generations. But I'm the last. My wife cleans public toilets. It's a junk contract --- 800~zloty {[}\$200{]} a month, full time.} (In Poland \textqq{flexible contracts} are more commonly known as \textqq{junk contracts}.)

\textqq{It's hard to find permanent work,} Kalabis admitted. \textqq{That's why young people go abroad.} Since their country joined the EU in 2004, at least two million Poles have emigrated, many of them to the UK. \textqq{My son and daughter dream of living in England. Capitalism is fine for those who have a head for business, but not for the rest of us.} Pinned to the wall were a Solidarity banner, Poland's coat of arms (a crowned white eagle on a red background), the ubiquitous portrait of Pope John Paul II, a signed photo of a local boxing champion --- and a 2016 Law and Justice Party (Prawo i Sprawiedliwosc, PiS) calendar.

Kalabis, though a miner and a Solidarity rep, also campaigns for this rightwing party. Solidarity even called on people to vote for the PiS candidate, Andrzej Duda, in the May 2015 presidential election. In the parliamentary elections that followed, the union did not give any instructions on voting, but everyone has got the message. \textqq{I'm Catholic, but that's not why I got involved. PiS are the only party who're supporting us, who're close to the people. After the methane explosion at Halemba, President Kaczynski~(\href{https://mondediplo.com/2016/03/02poland\#nb1}{\emph{1}}) came to see us. That was touching.} Kalabis detests the neoliberals of Civic Platform (PO), the centre-right party in power from 2007 to 2015~(\href{https://mondediplo.com/2016/03/02poland\#nb2}{\emph{2}}). He was shocked that President Bronislaw Komorowski attended the funeral of General Wojciech Jaruzelski, the Communist president (1981-9) who suppressed Solidarity. And he still hasn't come to terms with the PO government's January~2015 plan to close the mines without consultation: \textqq{I heard about the closure plans for my pit on TV.} He believes that the former PO prime minister Donald Tusk, now president of the European Council, \textqq{wants to close all the mines, whereas the PiS has vowed to save them. Most of my workmates vote PiS.}

In October 2015 PiS won the parliamentary elections in both the lower house (Sejm) and the Senate, with 37.6\% of the vote, against 24.1\% for the neoliberals and 8.8\% for the populist Kukiz 15. The progressive camp failed to clear the threshold (5\% for parties, 8\% for coalitions) and have no parliamentary representation~(\href{https://mondediplo.com/2016/03/02poland\#nb3}{\emph{3}}). The left --- which is divided between United Poland and Poland Together --- has had its welfare ideas co-opted by the reactionary right and won no seats. The presidential election was a foretaste of this groundswell of support for the right: the incumbent, Bronislaw Komorowski, was beaten in the second round by the virtually unknown Duda

\section{Polish values matter}

Despite many attempts, no PiS representative agreed to be interviewed~(\href{https://mondediplo.com/2016/03/02poland\#nb4}{\emph{4}}). But there is an insight into the party's ideology in what foreign minister Witold Waszczykowski told the German tabloid~\emph{Bild}~in January: \textqq{Who says the world had to evolve according to a Marxist model in a single direction --- towards a mixing of cultures and races; a world of cyclists and vegetarians who only use renewable energy and fight all forms of religion? None of this has anything to do with traditional Polish values. It goes against what the majority of Poles hold dear: tradition, a sense of their history, a love of their country, faith in God and normal family life with a man and a woman}~(\href{https://mondediplo.com/2016/03/02poland\#nb5}{\emph{5}}).

Conservative values are not the only motivation for PiS voters. The party has found recruits in the Poland of job insecurity and falling living standards concealed behind strong macro-economic indicators; the Poland specialised in manufacturing low-end goods for big European companies, especially German ones; the Poland of pensions of less than~\$330 a month. Ordinary Poles, like Kalabis and his family, have suffered under neoliberal reforms and often have to choose between a \$250-a-month junk contract and emigrating. The nationalist, pro-religion, protectionist, xenophobic PiS has attracted these disappointed people with an ambitious welfare programme: a family allowance of 500 zloty (\$130) a month per child, funded through a tax on banks and big business; a minimum wage; and a return to a retirement age of~60 for women and 65 for men (PO had planned to raise it to~67 for both).

Professor Radoslaw Markowski, a political scientist at the University of Warsaw, has studied PiS's evolution: \textqq{When they were in power between 2005 and 2007, they were conservative, but economically neoliberal. They have become increasingly populist, xenophobic and Eurosceptic: it's a form of Catholic nationalism, sweetened with a welfare package.} He identifies three groups of PiS voters: \textqq{First, there's what I call the Smolensk sect, the people who're convinced that the April~2010 crash~(\href{https://mondediplo.com/2016/03/02poland\#nb6}{\emph{6}}) was the result of a plot by Donald Tusk and Vladimir Putin. Then there are the practising Catholics, whose knowledge of the world is often limited to what their priest tells them. A third of Poland's practising Catholics have had experience of the Church's political propaganda.} Lastly, there are the poor, who are attracted by the party's welfare programme: \textqq{PiS has successfully worked out what workers and peasants want.} The low turnout at the polls --- nearly 50\% did not vote --- did the rest.

Jakub Majmurek, a sociologist at the leftwing thinktank Krytyka Polityczna (political critique), has analysed the factors that led to PO being thrown out: \textqq{PO had been in control for eight years. That's a long time for a young democracy. Ewa Kopacz, the prime minister who succeeded Donald Tusk when he went to Brussels in 2014, lacked charisma.} PO never got over the \textqq{Waitergate} scandal. In June~2014 the conservative weekly~\emph{Wprost}~published private conversations of senior government officials secretly recorded by waiters in a Warsaw restaurant. Their coarse conversation, self-interest and smugness destroyed any illusion of a civic platform: \textqq{After that, they were seen as an elite who were cut off from reality.}

He also emphasised the liberals' self-satisfaction: \textqq{The PO's leaders are relatively old. They have lived through Communism and its shortages. Their boast was: `Look how Poland has leapt forward.' But that meant nothing to the young, who had no experience of that period. When they go and work in western Europe, they see that salaries there are much better. In Berlin, rents are a bit higher than in Warsaw, but people earn three times as much. The expectations of Poland's youth are very high.} As are their frustrations.

\section{Unemployment at 20\%}

Pawel Michalski, 34, owns a business in Bytom, a city in Silesia that has suffered since the closure of its mines. Shops have permanently pulled down iron shutters, and pensioners beg on the streets. He said: \textqq{We have 20\% unemployment here.} He campaigns for the Kukiz 15 movement, a populist, iconoclastic, anti-system party founded by Pawel Kukiz, a former rock musician. The party has been infiltrated by the ultranationalist far right. Kukiz got a 20\% share in the first round of the 2015 presidential elections and his movement, now ahead of the left, is Poland's third-largest political force. Michalski stood on the Kukiz 15 ticket in the parliamentary elections and got 15\% in Bytom. \textqq{Young people are emigrating. In England or Germany it's easy to find work. I have a friend who worked as a nurse here and earned 1,700~zloty a month {[}\$430{]}. You can't live on that. It's shameful. So she's gone to Germany.} Michalski is pro free market, but backs the PiS family allowance plan: \textqq{People are so poor. They need help.} He plays down the presence of ultranationalists in his party: \textqq{You know, we have all sorts in Kukiz.}

Robert Piaty, 33, studied political science, but has had a succession of junk contracts. He's currently working in a call centre for 1,400~zloty a month (\$355). \textqq{Half of my friends have gone to England. I spent six months there and made pounds1,000 {[}\$1,430{]} a month.} He's a member of the Sierpien 80~union (August 1980, a reference to the Solidarity strike) and votes for the Poland Together party (Razem, 3.6\% share of the vote), which would like to be Poland's Podemos. But he understands why young Poles worried about employment prospects vote for PiS in the hope of benefiting from its welfare programme: \textqq{They've promised a minimum hourly wage from July 2016.}

PiS has turned the screws on Poland's institutions: in the final week of 2015, the new government appointed five judges to the highest judicial body, the Constitutional Tribunal, passed a law that modified how it operates, and dismissed the heads of Poland's public broadcasting companies. In March a law will make the justice minister the new public prosecutor. Since mid-December, tens of thousands of Poles have demonstrated in response to calls from the Committee for the Defence of Democracy (KOD). And, in a move without precedent in EU history, the European Commission has launched a procedure against Poland to safeguard the rule of law --- a preliminary inquiry to determine whether Poland is contravening democratic principles.

\textqq{Polish democracy is very healthy,} says Aleksandra Rybinska, a journalist on the rightwing weekly~\emph{W Sieci,}~which ran a cover story under the headline \textqq{Conspiracy against Poland} above a photomontage linking Chancellor Angela Merkel and the president of the EU parliament, Martin Schulz, with the partition of Poland in~1772. Rybinska defends PiS policy: \textqq{PO appointed its own judges shortly before it lost the elections. So PiS would have been unable to get any laws through. As for media appointments, that's normal here: in 2008, colleagues on the right were sacked on PO orders. That didn't offend western sensibilities. The truth is that PiS represents everything despised by the '68 generation who are in power in the West. The West thought that Viktor Orban's Hungary would be an exception; but now Poland is also turning to traditional values. Brussels is afraid of conservative forces.}

Majmurek says: \textqq{PiS supporters think they're looked down on, persecuted by the liberal elites. Because their leaders were a bit younger than the liberals, they were mocked and referred to as `Pampers' {[}disposable nappies{]}, which fuelled their resentment of the political class. After their spell in the wilderness, they reckon their time has come. They want revenge.}

\section{`We want to defend our freedoms'}

Mateusz Kijowski, 47, wearing a lilac jacket, earrings and a ponytail, personifies everything the PiS abhors. He's an IT specialist who started the KOD on Facebook in November. \textqq{Within a few days, we had 55,000 followers.} We met in January; he was just back from Strasbourg, where the KOD delegation had received \textqq{a very warm welcome from liberal, socialist and green MEPs}. He was now preparing for a second wave of demonstrations in 46 cities and among the Polish diaspora across Europe. I showed him an online video made by the far right, which claims the KOD is funded by the American billionaire George Soros. He laughed and said \textqq{Unfortunately not. Seriously, no one expected these attacks on freedom. PiS didn't mention them in its campaign. It's behaving as though its mandate gives it unrestricted power, as though democracy means the absolute power of the electoral majority. It's undermining the fundamental principle of the EU, which is the separation of powers. We want to defend our freedoms.}

On the snowy Saturday after we met, 2,000~KOD supporters gathered in Solidarity Square in front of Gdansk's shipyards. Placards called for the defence of democracy. A cartoon compared the new head of state television, Jacek Kurski --- a native of Gdansk, \textqq{PiS's pitbull} --- to Jerzy Urban, the discredited spokesman of the old Communist regime. Demonstrators waved Polish, European and rainbow-coloured LGBT flags. Some wore white Guy Fawkes, V for Vendetta, masks. A drone flew overhead. The crowd waved in mockery at its spy camera.

\textqq{It's our duty to be here,} two pensioners said as the demonstration snaked its way to the city centre. \textqq{We demonstrated in 1980. We don't want any more dictatorships. We came here for the young people who don't realise what they could lose.} The average age of demonstrators was high: most were over 40. \textqq{I'm here of my own free will,} said a young woman. \textqq{I'm not in the pay of George Soros.} Why were young people not motivated to get involved? \textqq{They're apathetic. They have no political consciousness and they don't feel concerned. My younger brother, who's~18, wanted to vote for Kukiz. I managed to convince him to vote PO.} When they reached Dluga Street in the old town, the demonstrators chanted \textqq{We want to be ourselves}, a Solidarity slogan. Through a megaphone, the former dissident Alexander Hall condemned the fact that the PiS leader, Jaroslaw Kaczynski, calls all the shots in Poland, though he has no official function. A banner showed him as a puppeteer pulling the strings of President Duda and Prime Minister Beata Szydlo. After singing the national anthem and hearing the EU one played, the demonstrators dispersed, ignoring a few youths who shouted \textqq{the pigs are out of their sty}, a PiS taunt.

The KOD has a problem beyond young people's apathy or their support for the right and the populists: it finds it hard to extend its reach beyond liberal circles. No one I met on the left was willing to take part in its demonstrations. \textqq{The liberals are well off, the part of society that did well from the economic reforms,} said Piaty, a call-centre worker from Katowice. Ania Zawadzka, a feminist activist from Warsaw, turns out for Gay Pride marches and for the 11~November antifascist demonstration against the ultranationalists' march, but refuses to support the KOD: \textqq{The liberal intelligentsia are responsible for this situation. They refused to make the right to abortion more flexible to avoid offending the Church~(\href{https://mondediplo.com/2016/03/02poland\#nb7}{\emph{7}}). They made Poland an ultraliberal country, passed many laws against workers and despised and marginalised the poor. Because of them the people have moved to the right.}

Karol Guzikiewicz was a 16-year-old apprentice mechanic at the time of the strike in Gdansk led by Lech Walesa. He became Solidarity vice-president in the shipyards, but now campaigns for PiS: \textqq{The yards are a wilderness. They covered 100~hectares in 1990 but today it's just~20. There were 17,000~workers in 1990, but there are only~1,000 today. Now we mainly make wind turbines.} He's in no doubt that \textqq{all this is Donald Tusk and Europe's fault. Because of the liberals, Poland's labour laws are the worst in Europe. So yes, I've been campaigning for PiS since~2008. I joined this party because its welfare programme was similar to Solidarity's.} His former mentor Walesa told a Radio interviewer that the new government was \textqq{acting against democracy and liberty} and was making Poland an international laughing stock. But Guzikiewicz dismisses Brussels' criticisms: \textqq{Europe should look after its million migrants and leave Poland alone.}

In Gdansk I met former dissident Stefan Adamski, who in 1980 was editor of Solidarity's newsletter: \textqq{Solidarity members have been betrayed by the neoliberals. It's a brutal transition to Darwinian capitalism. It's not surprising that they turn to a party with a welfare programme, even an irresponsible one.} Adamski, a founder of Attac Poland~(\href{https://mondediplo.com/2016/03/02poland\#nb8}{\emph{8}}), campaigns for the leftwing Poland Together party. \textqq{Solidarity wasn't a supporter of capitalism. The union demanded respect for workers' rights from the Communist regime. PiS isn't questioning capitalism: it's simply promising greater solidarity. The saddest part is that Kaczynski won't be stopped by defenders of democracy. He'll be disciplined by the financial markets, which will oppose the implementation of welfare and protectionist measures.}

\section{Fact file}

\textbf{Population:}~\\
38.5 million (sixth largest in the EU)

\textbf{Economic growth:}~\\
3.6\% in 2015 (3.3\% in 2014; sixth largest economy in the EU)

\textbf{GDP:}~\\
EUR 413bn

\textbf{Per capita GDP:}~\\
EUR 18,600

\textbf{Unemployment:}~\\
9.8\% in December 2015 (11.4\% in December 2014)

\textbf{Gross median wage:}~\\
EUR 686 (EUR 1,717 in France, EUR 397 in Romania)

\textbf{Hourly labour costs:}~\\
EUR 8.4 (EUR 31.4 in Germany)

\textbf{Precarious employment:}~\\
1.4 million workers on flexible contracts (aka \textqq{junk contracts})

\emph{Sources}: Central Statistical Office of Poland (GUS); Krytyka Polityczna; Eurostat

\end{document}

