% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\usepackage{hyperref}
\begin{document}
\author{Jean-Arnault Dérens}
\title{ Ukraine: not all rebellions are revolutions  }
\date{2014, március 28 \thanks{LMD } }
\maketitle

\begin{abstract}
  \textsc{Ukraine: not all rebellions are revolutions}

\textbf{New deal, same players}

\textbf{Ukraine's president has fled and there is an interim government, but the power brokers who will make or break the country's future include many of the same oligarchs who backed the last regime. And the corruption won't stop.}

\textsc{by Jean-Arnault Dérens and Laurent Geslin}

\end{abstract}


The Donbass Palace Hotel, off Lenin Square in Donetsk, is the most luxurious in eastern Ukraine. Rooms in this flagship of Rinat Akhmetov's empire cost around \$500 a night, much more than the average monthly wage. Akhmetov, the richest man in Ukraine, was a close associate of former president Viktor Yanukovych but after the Kiev uprising pragmatically switched his support to the new regime. Besides the hotel and real estate, Akhmetov owns the city's football team, Shakhtar Donetsk, mines, steelworks and factories. Among the clans of Ukraine's oligarchy, the largest fortunes have been made in the industrial and mining valley of the Don. This region, made up of the Donetsk and Luhansk oblasts, was a major centre of mining and industry in the Soviet era.

Donbass still generates a quarter of Ukraine's foreign currency earnings, even though only 95 mines are still officially active, compared to~230 two~decades ago. Over the same period, seven million people have left the country. Soon after independence in~1991, local people faced economic meltdown and the first state mine closures, and began mining illegally. \textqq{Round here you only have to dig down a metre to find coal,} said a former miner from Torez, an industrial city near Donetsk named after a French Communist leader~(\href{http://mondediplo.com/2014/04/03ukraine\#nb1}{\emph{1}}). In the makeshift tunnels, shored up with wooden props, accidents are common. Yet miners are willing to risk going underground for the prospect of just a few hundred dollars a month. After Yanukovych came to power in~2010, the network of these~\emph{kopanki}~(illegal mines) was regulated and controlled.

\textqq{Coal from the~\emph{kopanki}~was sold cheaply to the state mines, which sold it on at market price,} said Anatoly Akimochkin, vice-president of the Ukraine's Independent Miners' Union. As well as these profits, state mines also received government subsidies. \textqq{A large part of this money disappeared into the pockets of cronies of the regime,} said Akimochkin. According to expert estimates, 10\% of Ukraine's coal output in recent years has come from illegal mines. Behind this network is the former president's elder son, Oleksandr Yanukovych, who went into competition with the owners of the privatised mines, led by Akhmetov.

\section{Different deal of the cards}

\textqq{A revolution? No, it's just a different deal of the cards,} said sociologist Volodymyr Ishchenko, deputy director of the Centre for Society Research in Kiev. A few weeks after Yanukovych's removal, his frustration was clear: \textqq{This government defends the same values as the previous one: economic liberalism and getting rich. Not all rebellions are revolutions. It's unlikely that the Maidan movement will lead to profound changes that will justify calling it a revolution. The most serious candidate in the presidential election on 25~May is Petro Poroshenko, the `chocolate king' {[}because of the fortune he made in that industry{]}, one of the richest men in the country.} Even as demonstrators were being shot in the Maidan (Independence Square), the centre of popular anger since 22~November, a bizarre handover of power was being brokered behind closed doors with the powerful businessmen who have now taken control of Ukraine.

Over the past 20 years, Ukraine has experienced a form of development referred to as oligarchic pluralism. Many businessmen who amassed huge fortunes buying up mines and factories privatised cheaply after the fall of the Soviet Union have gone into politics. Oil and gas traders have become ministers or heads of major institutions. Former prime minister Yulia Tymoshenko, a leading figure in the 2004~Orange Revolution who was held up in the West as a martyr when she was imprisoned in~2011, made a fortune in the gas industry. A revolving door has developed between business and politics. Some powerful businessmen have played a more discreet role by financing the campaigns of politicians whom they expect to represent their interests. This system, which became the accepted way of doing things under President Leonid Kuchma (1994-2005), assumes constant reconfiguration shaped by the competing interests of the powerful, and their alliances and feuds.

Near the Donbass Palace, on top of an expensive building that houses the headquarters of Metinvest and DTek, two of Akhmetov's companies, there used to be an illuminated sign advertising Mako, the Swiss-registered holding company through which Yanukovych junior exports Ukrainian coal. A few days after Yanukovych senior fell from power, the sign came down, an indication that the alliance between Akhmetov and the president's men was over.

After 2010, Yanukovych, long viewed as the political representative of the Donetsk clan's interests, decided to demonstrate his independence. He appointed men he trusted ---members of the \textqq{family} --- to key state posts. Among them was Serhiy Arbuzov, his personal banker, who was put in charge of the national bank in~2010. He was briefly made prime minister at the height of the crisis on 28~January, after the departure of Mykola Azarov. The president also relied on Vitaliy Zakharchenko, a close friend of his son Oleksandr, whom he put in charge of the tax authorities in December~2010, before promoting him to interior minister. He also favoured the influential Dmytro Firtash, who for a time enjoyed a monopoly on Russian gas imports, before diversifying into chemicals and banking. Zakharchenko has now fled to Russia and Firtash was arrested in Vienna on 13~March.

\section{The young oligarchs}

The \textqq{family} also fostered the emergence of the \textqq{young oligarchs} group, in which Serhiy Kurchenko was the rising star. Kurchenko was born in Kharkiv in~1985 and owns Gas Ukraine, which controls 18\% of the liquid gas market and has a global turnover of \$10bn. He bought the Odessa refinery in~2012 as well as the football club in his home town, Metalist Kharkiv. His rise has been made possible by his close relationship with the son of Ukraine's former prosecutor general, Viktor Pshonka, another member of the \textqq{family}. In acquiring the Odessa refinery, Kurchenko went into competition with Ihor Kolomoyskyi, Ukraine's third wealthiest man and a major player in the oil market. \textqq{The competition was unfair,} journalist Anna Babinets said, \textqq{because Kurchenko enjoyed the support of the regime.}

After the family's fall, Kurchenko and the Pshonkas fled to Russia. On 2~March the new authorities appointed Kolomoyskyi as governor of the Dnipropetrovsk oblast. The same day Serhiy Taruta, a major figure in the steel industry and owner of the Industrial Union of Donbass, was put in charge of the Donetsk oblast. Taruta had backed the Orange Revolution financially, but was careful to avoid making his political allegiances too obvious. \textqq{Taruta and Akhmetov have never been friends. But after many clashes, they have managed to reach a kind of agreement over the control of our region,} said Valentin Kokorski, a political scientist at Donetsk University. \textqq{It is unthinkable that the appointment of Akhmetov's rival could have happened without his approval.} The struggle was fierce for a long time, though: Akhmetov pushed up his prices to try to force Taruta to relinquish control of his company.

One of the few advantages claimed for the oligarchic system is that it protects Ukraine from the influence of Russian capital~(\href{http://mondediplo.com/2014/04/03ukraine\#nb2}{\emph{2}}). \textqq{And yet,} Kokorski said, \textqq{it would be unrealistic to think the Ukrainian economy, especially in Donbass, could do without Russia. All our processing industries are directed to that market and very often do not meet EU standards. Our oligarchs are well aware that Ukraine can only find salvation by fulfilling its role as bridge between the EU and Russia.} Akhemtov's fortune may have its roots in Donbass, but his empire extends into Russia and the EU --- Bulgaria, Italy, the UK. He owns factories there as well shell companies and cross-holdings.

Taruta comes from a Greek community on the Azov Sea coast. His home town, the major port Mariupol, is one of the Akhmetov group's strongholds and site of the Azovstal and Ilyich iron and steel works, and the train manufacturer Azovmach, which exports almost all of its output to Russia. Soon after his appointment, Taruta met representatives of these industries in Mariupol. \textqq{The meeting was fruitful. No one has an interest in Ukraine splitting up,} said Nikolai Tokarsky, editor of the influential local paper~\emph{Priazovsky Rabochy,}~who attended the meeting. The paper belongs to Akhmetov's SCM holding company. Tokarsky is also an \textqq{independent} member of parliament for the Donetsk oblast, where he represents the oligarchs' interests. At the risk of alienating his readership, which is very sensitive to the siren call of Russia,~\emph{Priazovsky Rabochy~}is an advocate for the \textqq{territorial integrity} of Ukraine, a clear sign of Akhmetov's support for the new regime in Kiev.

\section{Defence against the `Russian threat'}

The government is counting on the oligarchs to try to compensate for the failure of the state and almost total disappearance of its apparatus. It is trying to involve them in the defence against the \textqq{Russian threat}, given that a lasting conflict would be disastrous for their interests. Akhmetov and Taruta seem fully aware of this danger and have increased their appeals for calm. After the violent clashes of 13~March, in which a demonstrator died in central Donetsk, Akhmetov put out a statement saying that \textqq{Donbass is a responsible region} populated by \textqq{brave, hardworking people} and that it would never give in to violence.

Throughout March, a bizarre battle pitted pro-Russian demonstrators against the security forces for control of public buildings in eastern Ukraine. They were occupied by protestors, then retaken by the police a few days later. When the siege of the regional administrative headquarters in Luhansk began on 9~March, 300~police officers with riot shields abandoned the building, cheered on by a crowd of~2,000, most of them women and pensioners. Many of the policemen smiled at the people who had just ejected them. Similar scenes were played out in Donetsk. \textqq{The police don't know who to obey any more. Their leaders used to serve the old authorities,} said Denis Kazantsky, a Donetsk blogger.

Chains of command are unclear at all levels of the security forces. Central administrations, where new leaders have been appointed, are barely functioning: \textqq{As far as corruption is concerned, the prosecutor's department is relying on the information that we journalists can provide, because the archives have disappeared,} said Anna Babinets. Since the Ukrainian army, according to interim president Oleksandr Turchynov, has just 6,000~combat-ready men, parliament has voted to establish a national guard. This force, which may incorporate nationalist extremists, including members of the far-right Pravyi Sektor (Right Sector)~(\href{http://mondediplo.com/2014/04/03ukraine\#nb3}{\emph{3}}), stands little chance of fulfilling security needs, but does risk fuelling the distrust of the people in the east. In Kharkiv on 14~March, Pravyi Sektor activists clashed with pro-Russians, with fatal consequences.

With the state apparently falling apart, the history of the \textqq{revolution} may already be a tale of missed opportunities. Alexander Tkachenko, an official in the Party of Regions in Luhansk, 30km from the Russian border, admits he was shocked, \textqq{as everyone was,} by Yanukovych's palatial villa with its gold toilets: \textqq{When we were young, we learned the old slogan `Peace to the Cottages! War on the Palaces!' But corruption has eaten away at the whole country.}

The people of eastern Ukraine might have joined those of western Ukraine in a single opposition movement against the oligarchs and corruption. But over-enthusiastic Ukrainian nationalism has turned off Russian-speakers in the east, and supporters of the former president talk of the \textqq{the fascist threat}. In just a few weeks, the manipulation of these fears and identity issues has brought the country to the verge of civil war.


\end{document}

