% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Nick Thorpe}
\title{Hungary: Parliament votes for new constitution   }
\date{2011, április 19 \thanks{BBC } }
\maketitle



Hungary's parliament has voted in favour of a new constitution which ends the transition from a totalitarian to a democratic system, its authors claim.

Two of the three opposition parties walked out of parliament before the vote.

They accuse the governing Fidesz party of imposing divisive right-wing ideology on the country.

The votes - 262 in favour to 44 against with one abstention - reflect the two-thirds majority enjoyed by Fidesz.

A storm of applause erupted from the benches of the governing party, followed by the Hungarian national anthem, once the result was declared.

The opposition far-right party Jobbik voted against the constitution.

The Socialist and Green parties walked out of the chamber before voting began.

\section{Democracy undermined}

The authors of the constitution call it a basic law for the 21st century.

One article limiting the size of the national debt has won international praise.

More controversial is the preamble which stresses Hungary's Christian roots.

Other criticisms include limits to the authority of the constitutional court and a reduction in the number of parliamentary ombudsmen.

A paragraph on the protection of the unborn child has the potential to open the way to limits on abortion.

The new constitution is the latest and most spectacular in a series of laws which the government has rushed through since election a year ago.

The government insists these strengthen Hungarian democracy.

Opponents say democracy is being steadily undermined. 

\end{document}

