% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{anonim}
\title{ Hungarians protest against new Fidesz constitution  }
\date{2012, január 2 \thanks{BBC } }
\maketitle


Tens of thousands of people have been protesting in Budapest over Hungary's controversial new constitution, a day after it came into force.

The country's governing Fidesz party pushed the law through parliament in April after winning a two-thirds majority in parliamentary elections.

Opponents say it threatens democracy by removing checks and balances set up in 1989 when Communism fell.

The EU and US had also asked for the law to be withdrawn.

The dispute has cast doubt over talks on a new financing agreement with the EU and IMF, seen as vital for market confidence in the central European country.

'Darkest spot'

Several centre-left opposition parties joined in the protests, held near a gala event organised by the government to celebrate the new constitution.

Protesters chanted slogans denouncing the centre-right Prime Minister, Viktor Orban, and carried placards denouncing his "dictatorship" as officials arrived for the event.

"Viktor Orban and his servants turned Hungary from a promising place to the darkest spot in Europe," said Socialist MP Tibor Szanyi, quoted by AFP news agency.

Aspects of the new constitution and accompanying laws which have come in for criticism include:

A preamble committed to defending the intellectual and spiritual unity of the nation, which experts warn could be a future source of tension

The inclusion of social issues - like the rights of the unborn child, marriage between a man and a woman, and the definition of life sentences - which experts say should be left to ethical debates within society

The rewriting of the electoral system, in a way which opponents say favours Fidesz

But Fidesz says the new constitution, or basic law, improves the legal framework of life in Hungary.

"Despite political debates we think it is an important value that for the first time, a freely elected parliament created the Basic Law," said Fidesz MP Gergely Gulyas, quoted by the Reuters news agency.

Mr Gulyas co-wrote the new law and shepherded it through parliament.

\section{Hungary's forint falls to record euro low}\label{hungarys-forint-falls-to-record-euro-low}

The Hungarian forint has dropped to a new record low against the euro as the country scaled back a debt sale and continued talks over a bailout.

The forint tumbled to 323.40 against the euro, the lowest since the 17-nation currency was founded in 1999.

Thousands of Hungarians have taken to the streets in protest at the country's new constitution.

The government said it now wants to strike a deal with international lenders "as soon as possible".

The constitution has been a source of tension in talks with the International Monetary Fund and European Union.

Hungary has been seeking a standby credit line of 15-20bn euros (\$19.5bn, \pounds12.6bn) from the IMF and EU as it runs into trouble issuing new debt.

But both have both cast doubts over aid because of the new laws passed by the ruling Fidesz party, which critics say undermine the independence of its central bank.

The new law allows the government to appoint more deputy governors, and Hungary's central bank governor, Andras Simor, has said the bill amounts to a takeover of the central bank.

The government wants to keep interest rates low to boost growth - but Hungary's central bank increased rates for the the second month in a row in December, to 7\% from 6.5\%.

Demonstrators say that other changes to the constitution harm democracy and remove the checks and balances set up in 1989 when communism fell.

The EU and US have asked for the law to be withdrawn.

Bond sale

Tamas Fellegi, who is in charge of negotiations with the IMF and EU, said on Thursday that the government wanted to strike a deal "as soon as possible".

"We are ready to negotiate without preconditions, and we are ready to discuss everything at the negotiating table," he said.

The state debt agency cut its sale of 12-month treasury bills by 10bn forints from the planned 45bn forints after there were fewer bids, and the yield surged - to 9.96\% from 7.91\% at the last auction.

Ratings agency Standard \& Poor's downgraded Hungary's debt to junk status in December, partly due to the changes to the constitution.

Hungary was last given a 20bn-euro standby loan by the IMF in 2008 to prevent it having to default on its debts.

\end{document}

