% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Hugh Schofield}
\title{ France's autism treatment 'shame'  }
\date{2012, április 13 \thanks{BBC} }
\maketitle

In many countries, the standard way of treating autistic children is with behavioural therapy - stimulating and rewarding them to develop the skills they need to function in society - but France still puts its faith in psychoanalysis. And an increasing number of parents are now demanding change.

For autism campaigners, it is one of the most serious health scandals of our times.

How for decades France turned its back on the latest scientific thinking, and treated autism as a form of psychosis.

How, as a result, tens of thousands of children were misdiagnosed - or not diagnosed at all - and consigned to lives of misery.

And how, to this day, in its approach to autism, the French medical establishment continues to believe in the powers of psychiatry and psychoanalysis - long after the rest of the world has switched to alternative methods of treatment.

"It is an out-and-out disgrace," according to Daniel Fasquelle, a member of parliament who campaigns on the issue.

"Every day I am contacted by parents with the same story - how their child's autism was not detected in time, so they never had the treatment that they needed.

"Thousands of children could have been saved. They do it everywhere else. Why not here? It is France's shame."

The row over autism in France has festered unreported for years. But recently it has become public. Independent associations have been created, lobbying for a move away from psychoanalysis and psychiatry, and over to the "behaviourist" treatments that prevail elsewhere.

In early March, these groups scored an important victory - with a ruling from the health ministry that calls into question the use of psychoanalysis as a treatment for autism.

But the psychoanalysts are not taking it lying down. From their point of view, behaviourism is a form of superficial social conditioning that does not address root causes - and they resent the way they have been typecast as the villains of the piece, when their aims are as sincere as those of their opponents.

"One thing that never pays in the field of autism is triumphalism," said Lauriane Brunessaux, a child psychiatrist. "Autism is far too complex, and we understand it so badly.


"Today it is the behaviourists who are being triumphalist."

The behaviourist approach to autism was developed in the 1970s and 80s in the US and Canada, and it is now the norm in most of the world.

Under the so-called ABA method (Applied Behavioural Analysis), autism is treated as primarily an educational - rather than a medical - problem.

With a set of rewards (which can be granted or withheld) - and with plenty of individual attention - children can learn to function in society, and be much less of a burden on their families.

"If you diagnose early, and then give the right treatment between the ages of two and seven, 70\% of autistic children can acquire functional language skills. Here in France, we are way off that figure," says Fasquelle.

"And the same pattern continues later in life. In the UK, there are 17 times more university students with autism than in France. It is unacceptable."

The blame - Fasquelle and autism associations argue - lies with a medical establishment that remains fixated with Freud.

"Today everyone knows that autism is a neuro-developmental problem. It is not a psychosis or mental disorder," says Muhamed Sajidi, president of the association Conquer Autism.

"But in France it is the psychiatrists - heavily influenced by Freudian psychoanalysis - who remain in charge. And they have shut themselves off from all the changes in our knowledge of autism."

Sajidi set up the association after his life was "destroyed", as he puts it, by the medical establishment's failure to diagnose his son Sami's autism.

For him, as for many others, one of the worst aspects is how blame for autism has been laid at the door of the child's parents, and more especially of the mother.

"The first time I went to see a doctor when my (autistic) son Gael was three and we thought there was a problem, the psychiatrist asked me if I had wanted him - if it had been a wanted pregnancy!" says Candy Lepenuizic, a British woman married to a Frenchman.

"Then she asked what sort of dreams I had had while I was pregnant with him. And suggested the whole family have a course of psychotherapy.

"At that point I got up and walked out. It was only because I had been warned this might happen that I did not burst into tears."

Such horror stories are typical in French families of autistic children.

"The whole idea was that it was la faute á maman (the mother's fault). It was the 'refrigerator mum', or there was some problem with the family dynamic," says Lepenuizic.



\end{document}



