% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{Nick Thorpe}
\title{ Hungary denies Roma families evacuated over attack fear  }
\date{2011, április 27 \thanks{BBC } }
\maketitle

Hungary's government has accused its political opponents of staging an unnecessary "evacuation" of Roma families from a northern village.

The village, Gyongyospata, was recently the scene of ethnic tension between local Roma and right-wing vigilantes.

A vigilante group was due to hold a paramilitary training exercise in the village at the weekend.

But officials have told the BBC there is no connection between the two events.

A total of 277 Roma were taken from the village by bus by the Hungarian Red Cross on Friday morning.

The action was reported by both local and international media as an evacuation of terrified Roma women and children, linked to the fact that a new far-right group called Vedero, or "Defensive Strength", were planning a paramilitary training weekend in the village.

This followed incidents last month in the same village, when another far-right group mounted anti-Roma patrols, ostensibly to protect the local population.

But Red Cross spokesman Erik Selymes said there was no connection between what he called a "pre-planned holiday" organised for the Roma and the paramilitary training exercise.

Zoltan Kovacs, a state secretary in the Hungarian Justice Ministry, told the BBC that the so-called evacuation was "a clear-cut political provocation" and completely unwarranted.

The government this week passed new legislation which aims to prevent paramilitary or other uniformed groups from encroaching on the role of the police to keep law and order.

27 April 2011 Last updated at 11:54 GMT

\section{Hungary Roma battle far-right vigilantes}\label{hungary-roma-battle-far-right-vigilantes}

Far-right vigilantes have clashed with Roma (Gypsies) in a village in north-east Hungary where ethnic tensions escalated last week.

At least four people were hurt in the brawl in Gyongyospata, in which dozens were involved, police say.

Hungarian media report that fighting erupted after Vedero ("Defensive Strength") uniformed paramilitaries hurled stones at a Roma house.

Far right groups have staged "public order" patrols in Roma areas recently.

The groups allege that such patrols are necessary to crack down on crime.

Police reinforcements have gone to Gyongyospata since the clash on Tuesday night, the Magyar Hirlap news website says.

Last week Hungary's government accused political opponents of staging an unnecessary "evacuation" of Roma families from the village, 81km (50 miles) from Budapest.

A total of 277 Roma were taken from the village by bus by the Hungarian Red Cross last Friday. They have now returned to their homes.

Media reports described it as an evacuation of Roma women and children frightened by Vedero's activities. But a Red Cross spokesman denied any connection, saying it was a pre-planned holiday.

Vedero announced it would set up a paramilitary training camp in the Roma part of Gyongyospata, which has about 2,800 inhabitants in total. However, the police prevented Vedero from doing so.

The government has passed new legislation aimed at preventing paramilitary or other uniformed groups from acting like self-appointed police.

\end{document}

