% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{anonim}
\title{ Hungary internet tax cancelled after mass protests  }
\date{2014, október 31 \thanks{BBC http://www.bbc.co.uk/news/world-europe-29846285} }
\maketitle


\textbf{Hungary has decided to shelve a proposed tax on internet data traffic after mass protests against the plan.}

"This tax in its current form cannot be introduced," Prime Minister Viktor Orban said on Friday.

Large-scale protests began on Sunday, when demonstrators hurled old computer parts at the headquarters of Mr Orban's ruling Fidesz party.

The draft law - condemned by the EU - would levy a fee on each gigabyte of internet data transferred.

The protesters objected to the financial burden but also feared the move would restrict free expression and access to information.

The levy was set at 150 forints (pounds0.40; 0.50 euros; \$0.60) per gigabyte of data traffic.

After thousands protested the government decided to cap the tax at 700 forints per month for individuals and 5,000 forints for companies. But that did not placate the crowds.

\textbf{'It should not be done'}

Fidesz had said the special tax was needed to balance Hungary's budget in 2015.

Speaking on Kossuth public radio, Mr Orban said that "if the people not only dislike something but also consider it unreasonable then it should not be done...

"The tax code should be modified. This must be withdrawn, and we do not have to deal with this now."

He said a measure seen by the government as a technical issue had become "a fear-inducing vision".

There will be a national consultation on it in January, he said.

A European Commission spokesman, Ryan Heath,~\textbf{\emph{said the tax was "bad in principle"}}~because it was a unilateral measure applied to a global phenomenon.

He said it was "part of a pattern... of actions that have limited freedoms or sought to take rents without achieving wider economic or social interest" in Hungary.

The Commission has previously criticised Mr Orban's government for constitutional proposals seen to be cementing the Fidesz party's political dominance.

\end{document}

