% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{anonim}
\title{ Hungary students in Kaposvar topless protest  }
\date{2013, október 13 \thanks{BBC http://www.bbc.co.uk/news/world-europe-24393250} }
\maketitle



A new dress code imposed by a university head in southern Hungary has prompted students to strip down to their underwear in protest.

Under the rules imposed on 1 October, women are barred from wearing mini-skirts and men have to wear dark suits.

The code, announced by Kaposvar University Rector Ferenc Szavai, is aimed at classes and exams.

A group of 10 art school students went topless in class on Thursday to protest at the instructions.

They were joined by their speech instructor who was taking the class.

According to the new code, women will have to wear long skirts or trousers and a blouse and jacket, while avoiding low neck-lines, heavy make-up and inappropriate fashion accessories.

Flip-flops are banned, as are unkempt hair and fingernails.

But the students objected to the conservative regulations and complained that the university had failed to legislate for warmer days.

Before the class took part in their underwear protest, they posed for a picture in their full clothing.

On its website, Kaposvar University says it aims to foster a "vibrant, vivacious, active life on campus".

\url{http://www.bbc.co.uk/news/world-europe-24393250} 

\end{document}

