% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{anonim}
\title{ Hungary's debt downgraded by Moody's to junk status  }
\date{2011, november 25 \thanks{BBC } }
\maketitle


Moody's has cut its rating of Hungarian government debt to junk status.

The ratings agency blamed Hungary's high levels of debt and weak prospects for growth, as well as uncertainty about whether the government can achieve its goals for the economy.

The government said the move by Moody's was part of a series of financial attacks against the country.

Earlier, Standard \& Poor's decided not to downgrade Hungary until talks with the EU and IMF had been completed.

Last week, the government said it would seek a financial safety net from the European Union and the International Monetary Fund, but no new loans.

The Hungarian economy ministry said the Moody's downgrade had "no basis because, despite all the external difficulties, in the past year-and-a-half there has been an expressly favourable change in most areas of the Hungarian economy".

Moody's cut Hungary's debt one notch, from Baa3 to Ba1, which is just below investment grade.

Hungary was given a 20bn euro standby loan by the IMF in 2008 to prevent it having to default on its debts.

But the newly-elected Prime Minister Viktor Orban decided not to renew the standby facility last year.

Mr Orban has followed an unconventional policy of allowing Hungarian households to convert their mortgages - which were predominately taken out in foreign currencies, particularly the Swiss franc, prior to the 2008 financial crisis - back into Hungarian forint at the expense of the banks.

At the same time, he has retaken control of private pension assets, to help pay off public debt.

The government has dismissed the views of the ratings agencies, saying that they are biased against Hungary because its economic policies do not favour banks.

\end{document}

