% -*- mode: LaTeX; coding: iso-8859-2 -*-
\documentclass[a4paper,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[latin2]{inputenc}
\usepackage[magyar]{babel}
\begin{document}
\author{anonim}
\title{  Hungary's Viktor Orban condemns anti-Semitism as WJC meets }
\date{2013, május 6 \thanks{BBC } }
\maketitle

Hungarian Prime Minister Viktor Orban has condemned anti-Semitism at the opening of a meeting of the World Jewish Congress (WJC) in Budapest.

He acknowledged anti-Semitism was on the rise in the country because of the economic crisis but said his government had a "zero-tolerance" towards it.

On Saturday the far-right Jobbik party held a protest against the Hungarian capital hosting the WJC meeting.

The party's leaders regularly issue anti-Semitic statements.

The WJC usually hosts its assembly in Jerusalem, but chose Hungary this year to highlight what it says is growing anti-Semitism in Europe.

Addressing the opening session, Mr Orban said: "Anti-Semitism is unacceptable and cannot be tolerated."

He added that his government had a "moral duty to declare zero tolerance on anti-Semitism".


\section{Not for sale}

However the WJC expressed its disappointment that Mr Orban had not specifically talked about Jobbik, the third-largest party in parliament.

"The prime minister did not confront the true nature of the problem - the threat posed by the anti-Semites in general and by the extreme-right Jobbik party in particular," it said in a statement.

The president of the World Jewish Congress, Ronald Lauder, said Jews were again wondering if they were safe in Hungary.

Saturday's rally by Jobbik was billed as a tribute to what organisers called the victims of Bolshevism and Zionism.

"The Israeli conquerors, these investors, should look for another country in the world for themselves because Hungary is not for sale," party chairman Gabor Vona told the crowd.

Jobbik MP Marton Gyongyosi said Hungary had "become subjugated to Zionism, it has become a target of colonisation while we, the indigenous people, can play only the role of extras".

Mr Orban had ordered police to ban the march, but a Budapest court overruled the ban.

\end{document}

